title: "MERTK Adult-Only Dataset Processing and Analysis"
author: "Meite I, Pesce L, Feinstein M, Thorp EB, Lantz C"
date: "2024-11-12"
output: html_document


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#------------------
#PACKAGES
#-----------------
```{r packages, include=FALSE}
library(Seurat)
library(BiocManager)
library(scater)
library(scran)
library(dplyr)
library(stringr)
library(utils)
library(ggplot2)
library(cowplot)
library(AnnotationHub)
library(patchwork)
library(Matrix)
library(remotes)
library(scuttle)
library(bigstatsr)
library(sctransform)
library(SingleR)
library(monocle3)
```

#------------------
#LOADING DATA
#-----------------

1. Loading Output Data from the CellRanger Pipeline
Reading the raw data for a single sample using a sparse matrix  with the Seurat function Read10X(). This function uses the CellRanger output directory to create the sparse matrix and save on RAM (memory) and CPU (processing capacity)

2. After Calling Read10X(), Cellranger creates metadata for each of the cells: 

orig.ident: this often contains the sample identity if known, but will default to "SeuratProject"
nCount_RNA: number of UMIs per cell
nFeature_RNA: number of genes detected per cell

```{r Loading Data, include = FALSE }

#-----------------------
#ONBOARDING DATA
#----------------------

#MER_Ad_MI
MER_Ad_MI <- Read10X( data.dir = "/projects/p31275/Isaac/MERTK/MER_Ad_MI")

#MER_MI
MER_MI <- Read10X(data.dir =  "/projects/p31275/Isaac/MERTK/MER_MI")

#WT_Ad_MI
WT_Ad_MI <- Read10X(data.dir =  "/projects/p31275/Isaac/MERTK/WT_Ad_MI")

#WT_P1_MI_1
WT_P1_MI_1 <- Read10X(data.dir =  "/projects/p31275/Isaac/MERTK/WT_P1_MI_1")

#WT_P1_MI_2
WT_P1_MI_2 <- Read10X(data.dir =  "/projects/p31275/Isaac/MERTK/WT_P1_MI_2")


#------------------------
#Convert to Seurat Object
#------------------------


MER_Ad_MI <- CreateSeuratObject( counts = MER_Ad_MI, project = "MER_Ad_MI")
MER_MI <- CreateSeuratObject( counts = MER_MI, project = "MER_MI")
WT_Ad_MI <- CreateSeuratObject( counts = WT_Ad_MI, project = "WT_Ad_MI")
WT_P1_MI_1 <- CreateSeuratObject( counts = WT_P1_MI_1, project = "WT_P1_MI_1")
WT_P1_MI_2 <- CreateSeuratObject( counts = WT_P1_MI_2, project = "WT_P1_MI_2")

#----------------------------
#Merging Biological Replicates
-----------------------------
merged_seurat_WT_P1 <- merge(x = WT_P1_MI_1, y = WT_P1_MI_2,
                             add.cell.ids = c("WT_P1_MI_1", "WT_P1_MI_2"),
                             project = "WT_P1_MI")

#----------------------------
#Merging All Datasets 
-----------------------------
#Labeling Cells 
WT_P1_MI_1$orig.ident <- "WT_P1_MI_1"
WT_P1_MI_2$orig.ident <- "WT_P1_MI_2"
MER_Ad_MI$orig.ident <- "MER_Ad_MI"
MER_MI$orig.ident <- "MER_MI"
WT_Ad_MI$orig.ident <- "WT_Ad_MI"

merged_seurat <- merge(
  x = merged_seurat_WT_P1,
  y = list(MER_Ad_MI, MER_MI, WT_Ad_MI))

head(merged_seurat@meta.data)
```


#------------------
#DATA QUALITY ASSESSMENT
#-----------------


1. Quality Control by accounting for the biology. Specifically, calculating the mitochondrial and ribosomal ratio. 
This number should be low due to the cell types (cardiac tissue) and sequencing (nucleus)
- number of genes detected per UMI: more genes per UMI detected represent larger complexity of the data 
- Mitochondrial ratio: indicates the number of genes originating from the mitochondrial genes
- Novelty Score: log10 of the number of UMIs per cell, then divide the log10 number of genes by the log10 number of UMIs.

#Quality Control

```{r Quality Control, include = FALSE}

# Novelty score (log10 genes per UMI)
merged_seurat$log10GenesPerUMI <- log10(merged_seurat$nFeature_RNA) / log10(merged_seurat$nCount_RNA)

# Mitochondrial Feature Expression Ratio - searches through all gene identifiers in the dataset beginning with "mt"
merged_seurat$mitoRatio <- PercentageFeatureSet(merged_seurat, pattern = "^mt-")
merged_seurat$mitoRatio <- merged_seurat$mitoRatio / 100  # Ensure we get ratio and not percentage value 

# Ribosomal Feature Expression Ratio
merged_seurat$riboRatio <- PercentageFeatureSet(merged_seurat, pattern = "^Rp[ls]")
merged_seurat$riboRatio <- merged_seurat$riboRatio / 100  # Ensure we get ratio and not percentage value 

```


#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata <- merged_seurat@meta.data
metadata$cells <- rownames(metadata)

# Assign metadata based on orig.ident patterns
metadata <- metadata %>%
  mutate(
    genotype = case_when(
      str_detect(orig.ident, "^MER_") ~ "KO",
      str_detect(orig.ident, "^WT_") ~ "WT",
      TRUE ~ "Other"
    ),
    sample = case_when(                     
      genotype == "WT" ~ "WT",           
      genotype == "KO" ~ "KO",
      TRUE ~ NA_character_
    ),
    age_group = case_when(
      orig.ident %in% c("MER_Ad_MI", "WT_Ad_MI") ~ "Adult",
      orig.ident %in% c("MER_MI", "WT_P1_MI_1", "WT_P1_MI_2") ~ "Neonatal",
      TRUE ~ NA_character_
    ),
    timepoint = "7dpi",
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )

# Assign back to Seurat object
merged_seurat@meta.data <- metadata
head(merged_seurat@meta.data)
```



#################################################
     ASSESSING DATA QUALITY: PLOTS
#################################################

Metrics to assess:
Cell counts: number of cell counts per sample (nCells)
UMI counts per cell
Genes detected per cell
Complexity (novelty score)
Mitochondrial counts ratio

#Notes
Doublet detection was not performed

#Displaying QC Metrics 
```{r Assessing Data Quality, include = FALSE}
# Define levels for the conditions
levels <- c("WT", "KO")

# Number of Cells
p1 <- metadata %>% 
  ggplot(aes(x = factor(x = merged_seurat@meta.data$sample, levels = levels), fill = sample)) + 
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  xlab("Condition") +
  ggtitle("nCells") +
  NoLegend()

# Number of UMIs
p2 <- metadata %>% 
  ggplot(aes(color = sample, x = nUMI, fill = sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = 500) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("nUMIs")

# Number of Genes
p3 <- metadata %>% 
  ggplot(aes(x = factor(x = merged_seurat@meta.data$sample, levels = levels), y = log10(nGene), fill = sample)) +
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  xlab("Condition") +
  ggtitle("nGenes") +
  NoLegend()

# Mitochondrial Counts Ratio
p4 <- metadata %>% 
  ggplot(aes(color = sample, x = mitoRatio, fill = sample)) + 
  geom_density(alpha = 0.1) + 
  scale_x_log10() + 
  theme_classic() +
  geom_vline(xintercept = 0.3) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Mito Ratio") +
  NoLegend()

# Ribosomal Counts Ratio
p5 <- metadata %>% 
  ggplot(aes(color = sample, x = riboRatio, fill = sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  geom_vline(xintercept = 0.5) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Ribo Ratio") +
  NoLegend()

# Novelty Gene (Unique genes per UMI)
p6 <- metadata %>%
  ggplot(aes(x = log10GenesPerUMI, color = sample, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Unique genes per UMI") +
  NoLegend()

# Combine all plots into a single grid
library(gridExtra)
plot1 <- grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)

# Display the combined plot
plot1



```


#Saving Quality Control Figures
```{r}
ggsave(
  filename = "QCmetrics.tiff", plot = plot1,path = "/projects/p31275/Isaac/MERTK/Figures", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)
```


```{r Genes versus UMI Correlations, echo=TRUE, message=FALSE, warning=FALSE}

#Correlation of genes v. UMI
plot2 <- metadata %>% 
  ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  geom_point() + 
  scale_colour_gradient(low = "deepskyblue3", high = "firebrick4") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500) +
  geom_hline(yintercept = 250) +
  facet_wrap(~sample)


plot2
```

#Saving Genes vs UMI correlations
```{r  eval=FALSE, include=FALSE}

# Save the nGene vs nUMI plot to a file
ggsave(
  filename = "nGene_v_nUMI.tiff", plot = plot2, path = "/projects/p31275/Isaac/MERTK/Figures/QC_Round1", device = "tiff", dpi = 500, width = 7, height = 5, units = "in", compression = "lzw"
)
```


#################################################
     FILTERING CELLS BASED ON QC METRICS 
#################################################

1. Cell-Level filtering 
Removal of low quality cells:
nUMI > 500
nGene > 250
log10GenesPerUMI > 0.8
mitoRatio < 0.3

2. Gene-Level Filtering 
Removal of genes with zero counts 

3. keeping Genes expressed in >= 10 cells 

```{r Filtered_seurat, include = FALSE}

filtered_seurat <- subset(x = merged_seurat, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
filtered_seurat <- JoinLayers(filtered_seurat)
counts <- GetAssayData(filtered_seurat, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
filtered_seurat <- subset(filtered_seurat, features = rownames(counts)[keep_genes])

```



### Revisiting QC Metrics with Filtered Data
```{r QC_metrics_revisted, include = FALSE}

metadata2 <- filtered_seurat@meta.data
#nCells
p7 <- metadata2 %>% 
  ggplot(aes(x=sample, fill=sample)) + 
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("nCells") +
  NoLegend()

#nUMIs
p8 <- metadata2 %>% 
  ggplot(aes(color=sample, x=nUMI, fill= sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = 500) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("nUMIS")

#nGenes 
p9 <- metadata2 %>% 
  ggplot(aes(x=sample, y=log10(nGene), fill=sample)) + 
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("nGenes") +
  NoLegend()

#mito counts ratio
p10 <- metadata2 %>% 
  ggplot(aes(color=sample, x=mitoRatio, fill=sample)) + 
  geom_density(alpha = 0.1) + 
  scale_x_log10() + 
  theme_classic() +
  geom_vline(xintercept = 0.6) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("Mito Ratio") +
  NoLegend()

#ribo counts ratio
p11 <- metadata2 %>% 
  ggplot(aes(color=sample, x=riboRatio, fill=sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("Ribo Ratio") +
  NoLegend()

#novelty gene
p12 <- metadata2 %>%
  ggplot(aes(x=log10GenesPerUMI, color = sample, fill=sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("Unique genes per UMI") +
  NoLegend()

p13 <- metadata2 %>% 
  ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  geom_point() + 
  scale_colour_gradient(low = "deepskyblue3", high = "firebrick4") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500) +
  geom_hline(yintercept = 250) +
  facet_wrap(~sample)
table(metadata$sample)


# Combine all plots into a single grid
library(gridExtra)
plot3 <- grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)

# Display the combined plot
plot3

```



#Saving Quality Control Figures
```{r}

ggsave(
  filename = "QCmetrics_revisited.tiff", plot = plot3,path = "/projects/p31275/Isaac/MERTK/Figures/QC_Round2", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

```



```{r Genes versus UMI Correlations_revisited, echo=TRUE, message=FALSE, warning=FALSE}

#Correlation of genes v. UMI
plot4 <- metadata2 %>% 
  ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  geom_point() + 
  scale_colour_gradient(low = "deepskyblue3", high = "firebrick4") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500) +
  geom_hline(yintercept = 250) +
  facet_wrap(~sample)


plot4
```

#Saving Genes vs UMI correlations
```{r  eval=FALSE, include=FALSE}

# Save the nGene vs nUMI plot to a file
ggsave(
  filename = "nGene_v_nUMI_revisited.tiff", plot = plot4, path = "/projects/p31275/Isaac/MERTK/Figures/QC_Round2", device = "tiff", dpi = 500, width = 7, height = 5, units = "in", compression = "lzw"
)


```






#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

1. Normalizing the gene expression values and identifying variant genes
Expression counts must be comparable between genes/samples (account for sequencing depth and gene length)

2. Scaling 
  - first step to normalization is to multiply each UMI count by a cell specific factor to get all cells to have the same UMI count 
  -Different cells have different amounts of mRNA (*differences in cell types/ variations)
  
3. Log Transformation
- log transformation: issue is that genes with different abundances are affected differently 
- effective normalization is observed with low/medium abundance genes

4. SCT Transform - Pearson Residual 
-Pearson Residual --> measurements are multiplied by a gene-specific weight
- each gene is weighted based on how much evidence there is that it is non-uniformly expressed across cells
- genes expressed in small fraction of cells will be favored (useful for finding rare cell populations)
- accounts for both the expression level but also the distribution of that expression across all cells 

```{r}
#Normalize the counts
Normalized_seurat <-NormalizeData(filtered_seurat)

seurat_phase <- CellCycleScoring(Normalized_seurat, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

```


The metadata now of the seurat object contains the assigned phase of the cycle for each cell



#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_phase <- FindVariableFeatures(seurat_phase, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_phase <- ScaleData(seurat_phase)

```


```{r}

# Perform PCA
seurat_phase <- RunPCA(seurat_phase)
# Plot the PCA colored by cell cycle phase
plot5 <- DimPlot(seurat_phase,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot5
```


#Saving Quality Figures
```{r eval= FALSE, include=FALSE}

ggsave(
  filename = "Cell_phase.tiff", plot = plot5, path = "/projects/p31275/Isaac/MERTK/Figures", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

```


####SAVING OBJECT######
```{r}
saveRDS(seurat_phase,"/projects/p31275/Isaac/MERTK/DATA/seurat_phase.rds")
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################


```{r Analysis without Integration, include = FALSE}

library(SeuratData)


# Creating a combined grouping factor (WT_Neo, WT_Adult, KO_Adult, KO_Neonatal)
seurat_phase$group <- paste(seurat_phase$sample, seurat_phase$age_group, sep = "_")

# Splitting the RNA assay using this combined factor
seurat_phase[["RNA"]] <- split(seurat_phase[["RNA"]], f = seurat_phase$group)

seurat_phase <- Seurat::SCTransform(seurat_phase, method = "glmGamPoi")

seurat_phase <- RunPCA(seurat_phase)
seurat_phase <- RunUMAP(seurat_phase, dims = 1:30)

#Plotting by Sample
sample_no_integration <- DimPlot(seurat_phase, reduction = "umap", group.by = "sample")


#Plotting by age_group
age_group_no_integration <- DimPlot(seurat_phase, reduction = "umap", group.by = "age_group")

#Plotting by sample and age_group =  (group)
group_no_integration <- DimPlot(seurat_phase, reduction = "umap", group.by = "group")







```







#Saving No Integration  Figures
```{r eval= FALSE, include=FALSE}

ggsave(
  filename = "Sample.tiff", plot = sample_no_integration, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_NO_Integration", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

ggsave(
  filename = "Age_Group.tiff", plot = age_group_no_integration, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_NO_Integration", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)


ggsave(
  filename = "Group.tiff", plot = group_no_integration, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_NO_Integration", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

```



# ── CCA Integration───────────────
# Harmony was selected over CCA and MNN based on UMAP mixing quality.
# These chunks are retained for reproducibility but not used downstream.
```{r Analysis with Integration - Integrating Datasets, eval= FALSE}

seurat_phase <- IntegrateLayers(object = seurat_phase, method = CCAIntegration, normalization.method = "SCT", verbose = F)
seurat_phase <- FindNeighbors(seurat_phase, reduction = "integrated.dr", dims = 1:30)
seurat_phase <- FindClusters(seurat_phase, resolution = 0.6)

seurat_phase <- RunUMAP(seurat_phase, dims = 1:30, reduction = "integrated.dr")

#Plotting by Sample (Integrated)
sample_integration <- DimPlot(seurat_phase, reduction = "umap", group.by = "sample")

#Plotting by age_group (Integrated)
age_group_integration <- DimPlot(seurat_phase, reduction = "umap", group.by = "age_group")

#Plotting by sample and age_group =  (group) (Integrated)
group_integration <- DimPlot(seurat_phase, reduction = "umap", group.by = "group")

```



#Saving  Integration  Figures
```{r eval= FALSE}

ggsave(
  filename = "Sample.tiff", plot = sample_integration, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_Integration_(Seurat CCA)", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

ggsave(
  filename = "Age_Group.tiff", plot = age_group_integration, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_Integration_(Seurat CCA)", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)


ggsave(
  filename = "Group.tiff", plot = group_integration, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_Integration_(Seurat CCA)", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

```



####SAVING OBJECT######
```{r eval=FALSE}
saveRDS(seurat_phase, "/projects/p31275/Isaac/MERTK/DATA/seurat_cca.rds")

```


########################################
Differential Expression Analysis 
########################################




####HARMONY INTEGRATION######
```{r Harmony Batch Correction, include=FALSE}

library(Seurat)
library(harmony)

# Set SCT as default assay
DefaultAssay(seurat_phase) <- "SCT"

#Confirming variable features are set
head(VariableFeatures(seurat_phase))

# Dimensionality reduction
seurat_phase <- RunPCA(seurat_phase, npcs = 50)

# Harmony on PCA embeddings, correcting for seq_folder
seurat_phase$seq_folder <- as.factor(seurat_phase$seq_folder)
seurat_phase <- RunHarmony(seurat_phase, group.by.vars = "seq_folder")

# UMAP on Harmony embeddings
seurat_phase <- RunUMAP(seurat_phase, reduction = "harmony", dims = 1:30)


#Visualization by biological Conditions
sample_harmony <- DimPlot(seurat_phase, reduction = "umap", group.by = "sample")       
age_group_harmony <- DimPlot(seurat_phase, reduction = "umap", group.by = "age_group")     
group_harmony  <- DimPlot(seurat_phase, reduction = "umap", group.by = "group")         


sample_harmony
age_group_harmony
group_harmony





```







#Saving  Harmony  Figures
```{r eval= FALSE, include=FALSE}

ggsave(
  filename = "Sample.tiff", plot = sample_harmony, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_Harmony", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

ggsave(
  filename = "Age_Group.tiff", plot = age_group_harmony, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_Harmony", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)


ggsave(
  filename = "Group.tiff", plot = group_harmony, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_Harmony", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

```


####SAVING OBJECT######
```{r}
saveRDS(seurat_phase, "/projects/p31275/Isaac/MERTK/DATA/seurat_harmony.rds")

```




# ── CCA Integration ───────────────
# Harmony was selected over CCA and MNN based on UMAP mixing quality.
# These chunks are retained for reproducibility but not used downstream.
```{r}

# Load required libraries
library(Seurat)
library(SingleCellExperiment)
library(batchelor)

# 1. Split the SCTransformed Seurat object by technical batch (seq_folder)
split_seurat <- SplitObject(seurat_phase, split.by = "seq_folder")

# 2. Make sure SCT is the default assay in each split object
split_seurat <- lapply(split_seurat, function(obj) {
  DefaultAssay(obj) <- "SCT"
  return(obj)
})

# 3. Extract variable features for integration
integration.features <- Reduce(intersect, lapply(split_seurat, VariableFeatures))
cat("Number of common variable features:", length(integration.features), "\n")

# 4. Convert Seurat objects to SingleCellExperiment (SCE) format
sce_list <- lapply(split_seurat, as.SingleCellExperiment)

# 5. Subset each SCE to common variable features
sce_list <- lapply(sce_list, function(sce) {
  sce[rownames(sce) %in% integration.features, ]
})

# 6. Run fastMNN to batch correct across seq_folder
mnn_out <- fastMNN(sce_list, d = 30)

# 7. Convert the "reconstructed" assay to a standard matrix
assay(mnn_out, "reconstructed") <- as.matrix(assay(mnn_out, "reconstructed"))

# 8. Convert integrated SCE back to Seurat object
seurat_mnn <- as.Seurat(mnn_out, counts = "reconstructed", data = "reconstructed")

# 9. Add corrected embeddings from fastMNN
seurat_mnn[["mnn"]] <- CreateDimReducObject(
  embeddings = reducedDim(mnn_out, "corrected"),
  key = "MNN_",
  assay = DefaultAssay(seurat_mnn)
)

# 10. Run UMAP on MNN reduction
seurat_mnn <- RunUMAP(seurat_mnn, reduction = "mnn", dims = 1:30)

#Recover metadata 
seurat_mnn@meta.data <- seurat_phase@meta.data[colnames(seurat_mnn), ]

# 11. Visualize
plot6 <- DimPlot(seurat_mnn, reduction = "umap", group.by = "sample")       # WT vs KO
plot7 <- DimPlot(seurat_mnn, reduction = "umap", group.by = "age_group")    # Neonatal vs Adult
plot8 <- DimPlot(seurat_mnn, reduction = "umap", group.by = "group")        # Composite groups

plot6
plot7
plot8


```



#Saving  MNN  Figures
```{r eval= FALSE, include=FALSE}

ggsave(
  filename = "Sample.tiff", plot = plot6, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_MNN", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

ggsave(
  filename = "Age_Group.tiff", plot = plot7, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_MNN", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)


ggsave(
  filename = "Group.tiff", plot = plot8, path = "/projects/p31275/Isaac/MERTK/Figures/UMAP_MNN", device = "tiff", dpi = 500, width = 7, height = 7, units = "in",compression = "lzw"
)

```



####SAVING OBJECT######
```{r}
saveRDS(seurat_mnn, "/projects/p31275/Isaac/MERTK/DATA/seurat_mnn.rds")

```



#################################################
ATLAS CREATION 
ATLAS CREATION
ATLAS CREATION
################################################




#NORMALIZATION OF DATASET W/ ACCESSION E-MTAB-13264


```{r Normalizing Dataset ACCESSION: E-MTAB-13264}


seurat_13264 <- readRDS("/projects/p31275/Isaac/MERTK/Atlas_Datasets/E-MTAB-13264.rds")

#Checking metadata
table(seurat_13264$condition)
#selecting HFrEF condition
seurat_13264_HF <- subset(seurat_13264, subset = condition == "Mice-HFrEF")

#saving original clusters:
seurat_13264_HF$orig_cluster <- Idents(seurat_13264_HF)


```



#Quality Control

```{r Quality Control, include = False}

# Quality Control Setup
DefaultAssay(seurat_13264_HF) <- "RNA"

# Human patterns due to multi-species atlas: use MT-, RPS, RPL 
seurat_13264_HF[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_13264_HF,
  pattern = "^MT-"
)

seurat_13264_HF[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_13264_HF,
  pattern = "^RP[SL]"
)

# Convert to ratios
seurat_13264_HF$mitoRatio <- seurat_13264_HF$percent.mito / 100
seurat_13264_HF$riboRatio <- seurat_13264_HF$percent.ribo / 100

```




#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_13264_HF <- seurat_13264_HF@meta.data

# Assign metadata based on orig.ident patterns
metadata_13264_HF <- metadata_13264_HF %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )

# Assign back to Seurat object
seurat_13264_HF@meta.data <- metadata_13264_HF
seurat_13264_HF$log10GenesPerUMI <- log10(seurat_13264_HF$nGene) / log10(seurat_13264_HF$nUMI)

```



```{r Filtered_seurat, include = False}

seurat_13264_HF <- subset(x = seurat_13264_HF, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Get raw counts slot
counts <- GetAssayData(seurat_13264_HF, assay = "RNA", slot = "counts")

# Filter genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_13264_HF <- subset(seurat_13264_HF, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_13264_HF <-NormalizeData(seurat_13264_HF)

seurat_13264_HF <- CellCycleScoring(seurat_13264_HF, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)


```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_13264_HF <- FindVariableFeatures(seurat_13264_HF, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_13264_HF <- ScaleData(seurat_13264_HF)

```



```{r}

# Perform PCA
seurat_13264_HF <- RunPCA(seurat_13264_HF)
# Plot the PCA colored by cell cycle phase
plot19 <- DimPlot(seurat_13264_HF,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot19
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

NormalizeData, FindVariableFeatures, ScaleData and RunPCA were run above for cell-cycle scoring

```{r SCTransform, include = FALSE}

library(SeuratData)
library(glmGamPoi)

seurat_13264_HF <- Seurat::SCTransform(seurat_13264_HF, method = "glmGamPoi")

seurat_13264_HF <- RunPCA(seurat_13264_HF)
seurat_13264_HF <- RunUMAP( seurat_13264_HF, dims = 1:30)


```

####SAVING OBJECT######
```{r}
saveRDS(seurat_13264_HF, "/projects/p31275/Isaac/MERTK/DATA/seurat_13264_HF.rds")

```










#NORMALIZATION OF DATASET W/ ACCESSION EMTAB9816

```{r}

#---------------------
#.     MI@1D
#--------------------

counts_9816_1d <- Read10X( data.dir= "/projects/p31275/Isaac/MERTK/Atlas_Datasets/EMTAB9816/MI1D")

seurat_9816_1d <- CreateSeuratObject(counts = counts_9816_1d, project = "EMTAB9816_MI1D")


```


#Quality Control

```{r Quality Control, include = False}

seurat_9816_1d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_9816_1d,
  pattern = "^mt-"
)

seurat_9816_1d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_9816_1d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_9816_1d$mitoRatio <- seurat_9816_1d$percent.mito / 100
seurat_9816_1d$riboRatio <- seurat_9816_1d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_9816_1d <- seurat_9816_1d@meta.data

# Assign metadata based on orig.ident patterns
metadata_9816_1d <- metadata_9816_1d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_9816_1d@meta.data <- metadata_9816_1d
seurat_9816_1d$log10GenesPerUMI <- log10(seurat_9816_1d$nFeature_RNA) / log10(seurat_9816_1d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_9816_1d <- subset(x = seurat_9816_1d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_9816_1d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_9816_1d <- subset(seurat_9816_1d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_9816_1d <-NormalizeData(seurat_9816_1d)

seurat_9816_1d <- CellCycleScoring(seurat_9816_1d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_9816_1d <- FindVariableFeatures(seurat_9816_1d, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_9816_1d <- ScaleData(seurat_9816_1d)
```



```{r}

# Perform PCA
seurat_9816_1d <- RunPCA(seurat_9816_1d)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_9816_1d,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_9816_1d <- Seurat::SCTransform(seurat_9816_1d, method = "glmGamPoi")

seurat_9816_1d <- RunPCA(seurat_9816_1d)
seurat_9816_1d <- RunUMAP( seurat_9816_1d, dims = 1:30)



```

####SAVING OBJECT######
```{r}
saveRDS(seurat_9816_1d, "/projects/p31275/Isaac/MERTK/DATA/seurat_9816_1d.rds")

```








#-----------------
#EMTAB9816 MI3D
#-----------------



```{r}

#---------------------
#.     MI@3D
#--------------------
count_9816_3d <- Read10X( data.dir= "/projects/p31275/Isaac/MERTK/Atlas_Datasets/EMTAB9816/MI3D")

seurat_9816_3d <- CreateSeuratObject(counts = count_9816_3d, project = "EMTAB9816_MI3D")
```


#Quality Control

```{r Quality Control, include = False}

seurat_9816_3d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_9816_3d,
  pattern = "^mt-"
)

seurat_9816_3d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_9816_3d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_9816_3d$mitoRatio <- seurat_9816_3d$percent.mito / 100
seurat_9816_3d$riboRatio <- seurat_9816_3d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_9816_3d <- seurat_9816_3d@meta.data

# Assign metadata based on orig.ident patterns
metadata_9816_3d <- metadata_9816_3d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_9816_3d@meta.data <- metadata_9816_3d
seurat_9816_3d$log10GenesPerUMI <- log10(seurat_9816_3d$nFeature_RNA) / log10(seurat_9816_3d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_9816_3d <- subset(x = seurat_9816_3d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_9816_3d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_9816_3d <- subset(seurat_9816_3d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_9816_3d <-NormalizeData(seurat_9816_3d)

seurat_9816_3d <- CellCycleScoring(seurat_9816_3d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_9816_3d <- FindVariableFeatures(seurat_9816_3d, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_9816_3d <- ScaleData(seurat_9816_3d)
```



```{r}

# Perform PCA
seurat_9816_3d <- RunPCA(seurat_9816_3d)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_9816_3d,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_9816_3d <- Seurat::SCTransform(seurat_9816_3d, method = "glmGamPoi")

seurat_9816_3d <- RunPCA(seurat_9816_3d)
seurat_9816_3d <- RunUMAP( seurat_9816_3d, dims = 1:30)


```




####SAVING OBJECT######
```{r}
saveRDS(seurat_9816_3d, "/projects/p31275/Isaac/MERTK/DATA/seurat_9816_3d.rds")

```







#-----------------
#EMTAB9816 MI7D
#-----------------


```{r}

#---------------------
#.     MI@7D
#--------------------
count_9816_7d <- Read10X( data.dir= "/projects/p31275/Isaac/MERTK/Atlas_Datasets/EMTAB9816/MI7D")

seurat_9816_7d <- CreateSeuratObject(counts = count_9816_7d, project = "EMTAB9816_MI7D")

```


#Quality Control

```{r Quality Control, include = False}

seurat_9816_7d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_9816_7d,
  pattern = "^mt-"
)

seurat_9816_7d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_9816_7d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_9816_7d$mitoRatio <- seurat_9816_7d$percent.mito / 100
seurat_9816_7d$riboRatio <- seurat_9816_7d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_9816_7d <- seurat_9816_7d@meta.data

# Assign metadata based on orig.ident patterns
metadata_9816_7d <- metadata_9816_7d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_9816_7d@meta.data <- metadata_9816_7d
seurat_9816_7d$log10GenesPerUMI <- log10(seurat_9816_7d$nFeature_RNA) / log10(seurat_9816_7d$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_9816_7d <- subset(x = seurat_9816_7d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_9816_7d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_9816_7d <- subset(seurat_9816_7d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_9816_7d <-NormalizeData(seurat_9816_7d)

seurat_9816_7d <- CellCycleScoring(seurat_9816_7d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_9816_7d <- FindVariableFeatures(seurat_9816_7d, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_9816_7d <- ScaleData(seurat_9816_7d)

```



```{r}

# Perform PCA
seurat_9816_7d <- RunPCA(seurat_9816_7d)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_9816_7d,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_9816_7d <- Seurat::SCTransform(seurat_9816_7d, method = "glmGamPoi")

seurat_9816_7d <- RunPCA(seurat_9816_7d)
seurat_9816_7d <- RunUMAP( seurat_9816_7d, dims = 1:30)


```




####SAVING OBJECT######
```{r}
saveRDS(seurat_9816_7d, "/projects/p31275/Isaac/MERTK/DATA/seurat_9816_7d.rds")

```




#NORMALIZATION OF DATASET W/ ACCESSION GSE119355


```{r}

counts_Epelman <- Read10X(data.dir = "/projects/p31275/Isaac/MERTK/Atlas_Datasets/GSE119355")

seurat_Epelman <- CreateSeuratObject(counts = counts_Epelman, project = "GSE119355")



```


#Quality Control

```{r Quality Control, include = False}

seurat_Epelman[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_Epelman,
  pattern = "^mt-"
)

seurat_Epelman[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_Epelman,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_Epelman$mitoRatio <- seurat_Epelman$percent.mito / 100
seurat_Epelman$riboRatio <- seurat_Epelman$percent.ribo / 100
```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_Epelman <- seurat_Epelman@meta.data

# Assign metadata based on orig.ident patterns
metadata_Epelman <- metadata_Epelman %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_Epelman@meta.data <- metadata_Epelman
seurat_Epelman$log10GenesPerUMI <- log10(seurat_Epelman$nFeature_RNA) / log10(seurat_Epelman$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_Epelman <- subset(x =seurat_Epelman, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_Epelman, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_Epelman <- subset(seurat_Epelman, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_Epelman <-NormalizeData(seurat_Epelman)

seurat_Epelman <- CellCycleScoring(seurat_Epelman, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_Epelman <- FindVariableFeatures(seurat_Epelman, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_Epelman <- ScaleData(seurat_Epelman)

```



```{r}

# Perform PCA
seurat_Epelman <- RunPCA(seurat_Epelman)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_Epelman,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_Epelman <- Seurat::SCTransform(seurat_Epelman, method = "glmGamPoi")

seurat_Epelman <- RunPCA(seurat_Epelman)
seurat_EPelman <- RunUMAP(seurat_Epelman, dims = 1:30)


```




####SAVING OBJECT######
```{r}
saveRDS(seurat_Epelman, "/projects/p31275/Isaac/MERTK/DATA/seurat_Epelman.rds")

```










#NORMALIZATION OF DATASET W/ ACCESSION EMTAB7895


```{r}

#---------------------
#.     MI@1D
#--------------------

counts_7895_1d <- Read10X(data.dir = "/projects/p31275/Isaac/MERTK/Atlas_Datasets/EMTAB7895/MI1D/filtered_feature_bc_matrix")

seurat_7895_1d <- CreateSeuratObject(counts = counts_7895_1d, project = "EMTAB7895_1d")
```


#Quality Control

```{r Quality Control, include = False}

seurat_7895_1d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_7895_1d,
  pattern = "^mt-"
)

seurat_7895_1d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_7895_1d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_7895_1d$mitoRatio <- seurat_7895_1d$percent.mito / 100
seurat_7895_1d$riboRatio <- seurat_7895_1d$percent.ribo / 100
```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_7895_1d <- seurat_7895_1d@meta.data

# Assign metadata based on orig.ident patterns
metadata_7895_1d <- metadata_7895_1d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_7895_1d@meta.data <- metadata_7895_1d
seurat_7895_1d$log10GenesPerUMI <- log10(seurat_7895_1d$nFeature_RNA) / log10(seurat_7895_1d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_7895_1d <- subset(x =seurat_7895_1d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_7895_1d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_7895_1d <- subset(seurat_7895_1d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_7895_1d <-NormalizeData(seurat_7895_1d)

seurat_7895_1d <- CellCycleScoring(seurat_7895_1d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_7895_1d <- FindVariableFeatures(seurat_7895_1d, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_7895_1d <- ScaleData(seurat_7895_1d)
```



```{r}

# Perform PCA
seurat_7895_1d <- RunPCA(seurat_7895_1d)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_7895_1d,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_7895_1d <- Seurat::SCTransform(seurat_7895_1d, method = "glmGamPoi")

seurat_7895_1d <- RunPCA(seurat_7895_1d)
seurat_7895_1d <- RunUMAP(seurat_7895_1d, dims = 1:30)


```




####SAVING OBJECT######
```{r}
saveRDS(seurat_7895_1d, "/projects/p31275/Isaac/MERTK/DATA/seurat_7895_1d.rds")

```











#--------------------
#EMTAB7895 3D
#--------------------


```{r}

#---------------------
#.     MI@3D
#--------------------

counts_7895_3d <- Read10X(data.dir = "/projects/p31275/Isaac/MERTK/Atlas_Datasets/EMTAB7895/MI3D/filtered_feature_bc_matrix")

seurat_7895_3d <- CreateSeuratObject(counts = counts_7895_3d, project = "EMTAB7895_3d")
```


#Quality Control

```{r Quality Control, include = False}

seurat_7895_3d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_7895_3d,
  pattern = "^mt-"
)

seurat_7895_3d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_7895_3d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_7895_3d$mitoRatio <- seurat_7895_3d$percent.mito / 100
seurat_7895_3d$riboRatio <- seurat_7895_3d$percent.ribo / 100
```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_7895_3d <- seurat_7895_3d@meta.data

# Assign metadata based on orig.ident patterns
metadata_7895_3d <- metadata_7895_3d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_7895_3d@meta.data <- metadata_7895_3d
seurat_7895_3d$log10GenesPerUMI <- log10(seurat_7895_3d$nFeature_RNA) / log10(seurat_7895_3d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_7895_3d <- subset(x =seurat_7895_3d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_7895_3d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_7895_3d <- subset(seurat_7895_3d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_7895_3d <-NormalizeData(seurat_7895_3d)

seurat_7895_3d <- CellCycleScoring(seurat_7895_3d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_7895_3d <- FindVariableFeatures(seurat_7895_3d, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_7895_3d <- ScaleData(seurat_7895_3d)

```



```{r}

# Perform PCA
seurat_7895_3d <- RunPCA(seurat_7895_3d)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_7895_3d,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_7895_3d <- Seurat::SCTransform(seurat_7895_3d, method = "glmGamPoi")

seurat_7895_3d <- RunPCA(seurat_7895_3d)
seurat_7895_3d <- RunUMAP(seurat_7895_3d, dims = 1:30)


```




####SAVING OBJECT######
```{r}
saveRDS(seurat_7895_3d, "/projects/p31275/Isaac/MERTK/DATA/seurat_7895_3d.rds")

```




#----------------
#EMTAB7895
#---------------


```{r}

#---------------------
#.     MI@7D
#--------------------

counts_7895_7d <- Read10X(data.dir = "/projects/p31275/Isaac/MERTK/Atlas_Datasets/EMTAB7895/MI7D/filtered_feature_bc_matrix")

seurat_7895_7d <- CreateSeuratObject(counts = counts_7895_7d, project = "EMTAB7895_7d")
```


#Quality Control

```{r Quality Control, include = False}

seurat_7895_7d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_7895_7d,
  pattern = "^mt-"
)

seurat_7895_7d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_7895_7d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_7895_7d$mitoRatio <- seurat_7895_7d$percent.mito / 100
seurat_7895_7d$riboRatio <- seurat_7895_7d$percent.ribo / 100
```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_7895_7d <- seurat_7895_7d@meta.data

# Assign metadata based on orig.ident patterns
metadata_7895_7d <- metadata_7895_7d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_7895_7d@meta.data <- metadata_7895_7d
seurat_7895_7d$log10GenesPerUMI <- log10(seurat_7895_7d$nFeature_RNA) / log10(seurat_7895_7d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_7895_7d <- subset(x =seurat_7895_7d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_7895_7d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_7895_7d <- subset(seurat_7895_7d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Convert human cell-cycle genes to mouse title case
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
    Normalization and Variability Regression
#################################################

```{r}
#Normalize the counts
seurat_7895_7d <-NormalizeData(seurat_7895_7d)

seurat_7895_7d <- CellCycleScoring(seurat_7895_7d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
```




#################################################
                  PCA Analysis 
#################################################

```{r}

# Identify the most variable genes
seurat_7895_7d <- FindVariableFeatures(seurat_7895_7d, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat_7895_7d <- ScaleData(seurat_7895_7d)

```



```{r}

# Perform PCA
seurat_7895_7d <- RunPCA(seurat_7895_7d)
# Plot the PCA colored by cell cycle phase
plot15 <- DimPlot(seurat_7895_7d,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

plot15
```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

1. Performing the Analysis without Integration (https://satijalab.org/seurat/articles/integration_introduction#perform-integration-with-sctransform-normalized-datasets)

Twist: At this point, NormalizeData, FindVariableFeatures, ScaleData, RunPCA have already been completed 


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_7895_7d <- Seurat::SCTransform(seurat_7895_7d, method = "glmGamPoi")

seurat_7895_7d <- RunPCA(seurat_7895_7d)
seurat_7895_7d <- RunUMAP(seurat_7895_7d, dims = 1:30)


```




####SAVING OBJECT######
```{r}
saveRDS(seurat_7895_7d, "/projects/p31275/Isaac/MERTK/DATA/seurat_7895_7d.rds")

```







#-----------------------------------------------
# MERGING ATLAS SEURAT OBJECTS 



# MERGING ATLAS SEURAT OBJECTS 



# MERGING ATLAS SEURAT OBJECTS 
#----------------------------------------------

```{r Building/Loading gene_map}

library(biomaRt)
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "mmusculus_gene_ensembl")

gene_map <- getBM(
  attributes = c("mgi_symbol", "ensembl_gene_id"),
  mart = mart
)
gene_map <- gene_map[gene_map$mgi_symbol != "" & gene_map$ensembl_gene_id != "", ]
gene_map <- unique(gene_map)

write.csv(gene_map,
          file = "/projects/b1246/Isaac/gene_map_mouse.csv",
          row.names = FALSE)


```



```{r SCT Integration of all the datasets}

suppressPackageStartupMessages({
  library(Matrix)
  library(Seurat)
})

collapse_sparse_rows <- function(mat, groups) {
  if (!inherits(mat, "dgCMatrix")) mat <- as(mat, "dgCMatrix")
  groups <- as.character(groups)
  ug <- unique(groups)
  sm <- summary(mat)
  gi <- match(groups[sm$i], ug)
  res <- sparseMatrix(
    i = gi, j = sm$j, x = sm$x,
    dims = c(length(ug), ncol(mat)),
    dimnames = list(ug, colnames(mat))
  )
  methods::as(res, "dgCMatrix")
}

map_to_ensembl <- function(counts, symbol_to_ensembl) {
  old <- rownames(counts)
  new <- symbol_to_ensembl[old]
  new_ids <- ifelse(is.na(new), paste0("SYM:", old), new)
  collapse_sparse_rows(counts, new_ids)
}

ensure_gene_universe <- function(counts, target_genes) {
  rn <- rownames(counts)
  miss <- setdiff(target_genes, rn)
  if (length(miss)) {
    z <- sparseMatrix(i = integer(0), j = integer(0), x = numeric(0),
                      dims = c(length(miss), ncol(counts)),
                      dimnames = list(miss, colnames(counts)))
    counts <- rbind(counts, z)
  }
  counts[match(target_genes, rownames(counts)), , drop = FALSE]
}

safe_counts <- function(obj, assay = "RNA") GetAssayData(obj, assay = assay, slot = "counts")

# mapping prep
symbol_to_ensembl <- setNames(gene_map$ensembl_gene_id, gene_map$mgi_symbol)


```




```{r SCT Normalization}

seurat_list <- list(
  Epelman   = seurat_Epelman,
  `13264_HF`= seurat_13264_HF,
  `7895_1d` = seurat_7895_1d,
  `7895_3d` = seurat_7895_3d,
  `7895_7d` = seurat_7895_7d,
  `9816_1d` = seurat_9816_1d,
  `9816_3d` = seurat_9816_3d,
  `9816_7d` = seurat_9816_7d
)

for (nm in names(seurat_list)) {
  obj <- seurat_list[[nm]]
  obj <- RenameCells(obj, add.cell.id = nm)

  counts <- safe_counts(obj, assay = "RNA")
  rownames(counts) <- rownames(counts)  # preserve case
  counts_mapped <- map_to_ensembl(counts, symbol_to_ensembl)

  obj[["RNA.all"]] <- CreateAssayObject(counts = counts_mapped)
  DefaultAssay(obj) <- "RNA.all"
  seurat_list[[nm]] <- obj
}

all_genes <- Reduce(union, lapply(seurat_list, \(o) rownames(o[["RNA.all"]])))

for (nm in names(seurat_list)) {
  obj <- seurat_list[[nm]]
  counts <- GetAssayData(obj, assay = "RNA.all", slot = "counts")
  counts <- ensure_gene_universe(counts, all_genes)
  obj[["RNA.all"]] <- CreateAssayObject(counts = counts)
  DefaultAssay(obj) <- "RNA.all"
  seurat_list[[nm]] <- obj
}

# SCTransform normalization
seurat_list <- lapply(seurat_list, function(o) {
  DefaultAssay(o) <- "RNA.all"
  SCTransform(
    o, assay = "RNA.all", new.assay.name = "SCT",
    return.only.var.genes = FALSE,
    conserve.memory = FALSE,
    verbose = TRUE
  )
})

seurat_list <- lapply(seurat_list, function(o) { DefaultAssay(o) <- "SCT"; o })

present_in_both <- function(o) {
  rc <- rownames(GetAssayData(o, assay = "SCT", slot = "counts"))
  rd <- rownames(GetAssayData(o, assay = "SCT", slot = "data"))
  intersect(rc, rd)
}
by_obj_both <- lapply(seurat_list, present_in_both)

features_raw <- SelectIntegrationFeatures(object.list = seurat_list, nfeatures = 3000)
features_use <- Reduce(intersect, c(list(features_raw), by_obj_both))
features_use <- unique(features_use[nzchar(features_use)])
stopifnot(length(features_use) >= 2000)

seurat_list <- lapply(seurat_list, function(o) {
  DefaultAssay(o) <- "SCT"
  rn <- rownames(o[["SCT"]])
  if (anyDuplicated(rn)) {
    m <- GetAssayData(o, assay = "SCT", slot = "counts")
    keep <- !duplicated(rn)
    o[["SCT"]] <- CreateAssayObject(counts = m[keep, , drop = FALSE])
    DefaultAssay(o) <- "SCT"
  }
  VariableFeatures(o) <- features_use
  o
})

seurat_list <- PrepSCTIntegration(
  object.list     = seurat_list,
  anchor.features = features_use,
  verbose         = TRUE
)

anchors <- FindIntegrationAnchors(
  object.list          = seurat_list,
  normalization.method = "SCT",
  anchor.features      = features_use,
  dims                 = 1:30,
  verbose              = TRUE
)

seurat_atlas <- IntegrateData(
  anchorset            = anchors,
  normalization.method = "SCT",
  dims                 = 1:30,
  verbose              = TRUE
)

DefaultAssay(seurat_atlas) <- "integrated"
seurat_atlas <- seurat_atlas |>
  RunPCA(npcs = 30) |>
  RunUMAP(dims = 1:30) |>
  FindNeighbors(dims = 1:30) |>
  FindClusters(resolution = c(0.1,0.2,0.3,0.4, 0.5,0.6,0.7,0.8, 0.9))
DimPlot(seurat_atlas, reduction = "umap", label = TRUE)

```



```{r Gene Mapping}

# Load mapping
gene_map <- read.csv("/scratch/inm3056/gene_map_mouse.csv")
symbol_to_ensembl <- setNames(gene_map$ensembl_gene_id, toupper(gene_map$mgi_symbol))
ensembl_to_symbol <- setNames(gene_map$mgi_symbol, gene_map$ensembl_gene_id)

# Helper for DotPlot using gene symbols
DotPlotSymbols <- function(object, symbols, mapping, group.by) {
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  ens_to_symbol <- setNames(mapping$mgi_symbol, mapping$ensembl_gene_id)

  features_map <- symbol_to_ens[toupper(symbols)]
  features_use <- features_map[!is.na(features_map)]

  p <- DotPlot(object, features = features_use, group.by = group.by) +
    RotatedAxis() +
    scale_x_discrete(labels = ens_to_symbol[features_use])
  return(p)
}



```




####SAVING OBJECT######
```{r}
saveRDS(seurat_atlas, "/scratch/inm3056/seurat_atlas.rds")

```


#---------------------------------------------
CLUSTER ANNOTATIONS FOR THE ATLAS 
#---------------------------------------------


```{r Finding Main Cell Lineage (low resolution)}

DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.1", label = TRUE) + ggtitle("Res 0.1")
DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.2", label = TRUE) + ggtitle("Res 0.2")
DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.3", label = TRUE) + ggtitle("Res 0.3")
DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.5", label = TRUE) + ggtitle("Res 0.5")
DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.6", label = TRUE) + ggtitle("Res 0.6")
DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.7", label = TRUE) + ggtitle("Res 0.7")
DimPlot(seurat_atlas, reduction = "umap", group.by = "integrated_snn_res.0.9", label = TRUE) + ggtitle("Res 0.9")



```




```{r Myeloid vs Non-myeloid, include = FALSE}


# Set identities
Idents(seurat_atlas) <- "integrated_snn_res.0.2"

# Define marker lists (symbols)
cell_lineage_0.2 <- list(
  Myeloid  = c("Lyz2","Itgam","Adgre1","Csf1r","S100a8","S100a9"),
  Tcell    = c("Cd3e","Cd4","Cd8a"),
  Bcell    = c("Cd79a","Ms4a1"),
  Nk       = c("Ncr1","Klrb1c","Gzmb"),
  Dc       = c("Itgax","H2-Ab1","Cd74"),
  Endo     = c("Pecam1","Cdh5","Kdr"),
  Fibro    = c("Col1a1","Pdgfra"),
  Lec      = c("Prox1","Lyve1","Flt4"),
  Neut     = c("Ly6g","S100a8"),
  Eos      = c("Siglecf"),
  Baso     = c("Fcer1a","Mcpt8")
)

# Collapse to unique marker symbols
features_all <- unique(unlist(cell_lineage_0.2))

# Load gene map
gene_map <- read.csv("/projects/b1246/Isaac/gene_map_mouse.csv")

# Use helper to plot directly with symbols
p <- DotPlotSymbols(seurat_atlas,
                    symbols = features_all,
                    mapping = gene_map,
                    group.by = "integrated_snn_res.0.2") +
     ggtitle("Lineage markers across clusters (res 0.2)") +
     theme(axis.text.x.top = element_blank(),   # remove top labels
           axis.ticks.x.top = element_blank())  # remove top ticks too

p


```






```{r Cluster mapping}

Idents(seurat_atlas) <- "integrated_snn_res.0.2"
lv <- levels(seurat_atlas) 

map <- setNames(rep("other", length(lv)), lv)

map[c("0","3","4","5","9")] <- "myeloid"
map["7"] <- "B cell"
map["6"] <- "Endothelial"
map[c("1","2")] <- "Fibroblast"
map["8"] <- "Nk_T"

# apply without overwriting numeric Idents
seurat_atlas$lineage_r.02 <- plyr::mapvalues(
  x = as.character(Idents(seurat_atlas)),
  from = names(map), to = unname(map), warn_missing = FALSE
)


DimPlot(seurat_atlas, reduction="umap", group.by="lineage_r.02", label=FALSE)

# Sanity Check with helper
DefaultAssay(seurat_atlas) <- "SCT"

DotPlotSymbols(seurat_atlas,
               symbols = c("Lyz2","Itgam","Adgre1","Csf1r",   # Myeloid
                           "Cd3e","Cd4","Cd8a",               # T cell
                           "Ncr1","Klrb1c","Gzmb",            # NK
                           "Cd79a","Ms4a1",                   # B
                           "Pecam1","Cdh5","Kdr",             # Endo
                           "Col1a1","Pdgfra"),                # Fibro
               mapping = gene_map,
               group.by = "lineage_r.02") +
  ggtitle("Marker check by cell lineage")



```




####SAVING OBJECT######
```{r}
saveRDS(seurat_atlas, "/projects/b1246/Isaac/seurat_atlas.rds")

```


```{r FeaturePlotSymbols helper function}

FeaturePlotSymbols <- function(object, symbols, mapping, ...) {
  # Mapping
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  ens_to_symbol <- setNames(mapping$mgi_symbol, mapping$ensembl_gene_id)
  
  # Convert symbols -> Ensembl IDs
  features_map <- symbol_to_ens[toupper(symbols)]
  features_use <- features_map[!is.na(features_map)]
  
  # Run FeaturePlot
  p <- FeaturePlot(object, features = features_use, ...)
  
  # Patchwork output: relabel titles manually
  for (i in seq_along(p)) {
    feat <- features_use[i]
    sym  <- ens_to_symbol[feat]
    p[[i]] <- p[[i]] + ggtitle(sym)
  }
  return(p)
}



```




```{r Myeloid Sanity Check}

# Marker sets
Lymph_markers_neg <- c("Cd3e","Cd4","Cd8a","Klrb1c")    
Myeloid_markers   <- c("Itgam","Lyz2","Adgre1","Csf1r")

# Macrophage subsets 
Res_mphi   <- c("Timd4","Lyve1","Folr2","Mrc1")
Ccr2_mphi  <- c("Ccr2","Cd52","Trem2","Apoe")
MhcII_hi   <- c("Runx3","Cx3cr1","Mmp13","Il1b")

# Monocytes & Neutrophils
Monocyte   <- c("Ly6c2","Plac8","Ccr2","Cx3cr1")  
Neutrophils<- c("Ly6g","S100a8","S100a9")

# Dendritic subsets
Cdc1 <- c("Xcr1","Cd8a","Clec9a","H2-ab1","Cd24a")
Cdc2 <- c("Sirpa","Itgax","H2-ab1","Ptprc","Cd4")
Pdc  <- c("Ptprc","Itgax","H2-ab1","Siglech")

myeloid <- subset(seurat_atlas, subset = lineage_r.02 == "myeloid")
DefaultAssay(myeloid) <- "SCT"

FeaturePlotSymbols(myeloid,
                   Myeloid_markers,
                   mapping = gene_map,
                   order = TRUE, raster = TRUE)



```







```{r}

suppressPackageStartupMessages({
  library(plyr) 
  library(cowplot)
})

# Resolution
res <- 0.6

# Ensure neighbors/graph exist, cluster if needed
if (!"integrated_snn" %in% names(seurat_atlas@graphs)) {
  DefaultAssay(seurat_atlas) <- "integrated"
  seurat_atlas <- FindNeighbors(seurat_atlas, reduction = "pca", dims = 1:30)
}
seurat_atlas <- FindClusters(seurat_atlas, graph.name = "integrated_snn", resolution = res)

# Metadata column name
new_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- new_col

# Marker panels (symbols only)
marker_panels <- list(
  Mac_Resident   = c("Timd4","Lyve1","Folr2","Mrc1","Igf1","F13a1","Cd163"),
  Mac_CCR2       = c("Ccr2","Trem2","Apoe","Cd14","H2-Ab1"),
  Mac_MHCIIhi    = c("Runx3","Cx3cr1","Il1b","Maf","Klf2","H2-Aa","Cd74"),
  Monocyte       = c("Ly6c2","Csf1r","Itgam"),
  DC_cDC1        = c("Xcr1","Clec9a","Itgax","H2-Ab1","Cd74"),
  DC_cDC2        = c("Sirpa","Itgax","Cd4","H2-Ab1"),
  pDC            = c("Siglech","Irf7"),
  T_CD4          = c("Cd3e","Cd4"),
  T_CD8          = c("Cd3e","Cd8a"),
  T_Treg         = c("Cd3e","Cd4","Foxp3"),
  Bcell          = c("Cd79a","Ms4a1","Cd19"),
  NK             = c("Ncr1","Klrb1c","Gzmb"),
  LEC            = c("Prox1","Lyve1","Flt4","Nrp2","Pdpn"),
  Endo_arterial  = c("Id1","Mgll"),
  Endo_venous    = c("Plvap","Vwf"),
  Endo_general   = c("Pecam1","Cdh5","Kdr","Eng"),
  Fibro_RCF      = c("Bcl11a","Sox4","Pdgfra"),
  Fibro_Matr     = c("Atf3","Klf4","Klf2","Jun"),
  Neut           = c("Ly6g","S100a8","S100a9"),
  Eos            = c("Siglecf"),
  Baso           = c("Fcer1a","Mcpt8")
)

# Collect all unique markers
features_all <- unique(unlist(marker_panels))

# Split into halves for readability
n_feat <- length(features_all)
split_idx <- ceiling(n_feat / 2)
features_part1 <- features_all[seq_len(split_idx)]
features_part2 <- if (split_idx < n_feat) features_all[(split_idx + 1):n_feat] else character(0)

# ---- UMAP plots ----
p_lineage <- DimPlot(seurat_atlas, reduction = "umap",
                     group.by = "lineage_r.02", label = FALSE) +
  ggtitle("Broad lineages (annotated) @ res 0.2")

p_umap <- DimPlot(seurat_atlas, reduction = "umap",
                  group.by = new_col, label = TRUE) +
  ggtitle(paste0("UMAP @ res ", res))

print(p_lineage)
print(p_umap)

# ---- DotPlots separately ----
p_dot1 <- DotPlotSymbols(seurat_atlas,
                         symbols = features_part1,
                         mapping = gene_map,
                         group.by = new_col) +
  ggtitle(paste0("Markers (1/2) @ res ", res))
print(p_dot1)

if (length(features_part2) > 0) {
  p_dot2 <- DotPlotSymbols(seurat_atlas,
                           symbols = features_part2,
                           mapping = gene_map,
                           group.by = new_col) +
    ggtitle(paste0("Markers (2/2) @ res ", res))
  print(p_dot2)
}


```




```{r Resolution 0.6 cluster mappings}

res <- 0.6
res0.6_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res0.6_col

# Defining identities 
map_06 <- c(
  `4`  = "Mac_Resident/TLF+",
  `6`  = "Mac_CCR2+",
  `2`  = "Mac_MHCII^hi",
  `9`  = "Monocyte",
  `17` = "DCs",
  `16` = "NKT",
  `13` = "B cells",
  `1`  = "Neutrophils",
  # Endothelial clusters
  `23` = "Endothelial",
  `5`  = "Endothelial",
  # Fibroblast clusters
  `0`  = "Fibroblast",
  `7`  = "Fibroblast",
  `11` = "Fibroblast",
  `15` = "Fibroblast",
  `18` = "Fibroblast",
  `21` = "Fibroblast"
)

# Store mapping in a new metadata column
col_lab <- "lineage_r.06"

# Convert cluster ids to character to index map safely
cl_ids <- as.character(seurat_atlas[[res0.6_col]][,1])

# Apply mapping; anything not listed becomes "Other"
annot <- unname(map_06[cl_ids])
annot[is.na(annot)] <- "Other"

seurat_atlas[[col_lab]] <- factor(
  annot,
  levels = c(
    "Mac_Resident/TLF+","Mac_CCR2+","Mac_MHCII^hi","Monocyte",
    "DCs","NKT","B cells","Neutrophils","Endothelial","Fibroblast","Other"
  )
)


```





```{r Visualizing mappings at res 0.6}

p_lineage <- DimPlot(seurat_atlas, reduction="umap", group.by = col_lab, label=TRUE) +
  ggtitle("Broad lineages @ res 0.6")
p_clusters <- DimPlot(seurat_atlas, reduction="umap", group.by = new_col, label=TRUE) +
  ggtitle(paste0("UMAP @ res ", res))
p_lineage
p_clusters


```




```{r Res 0.8 - clustering round 2}

# Higher resolution clustering
res <- 0.8
if (!"integrated_snn" %in% names(seurat_atlas@graphs)) {
  DefaultAssay(seurat_atlas) <- "integrated"
  seurat_atlas <- FindNeighbors(seurat_atlas, reduction = "pca", dims = 1:30)
}
seurat_atlas <- FindClusters(seurat_atlas, graph.name = "integrated_snn", resolution = res)
new_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- new_col

# Marker panels
marker_panels_2 <- list(
  TLF_Mac= c("Timd4","Lyve1","Folr2","Ninj1","Gas6","Mrc1","Cd163","Igf1","Sepp1","Igfb4"),
  MhcII_hi= c("Cd14","Cst3","Lilra5","Apoe","H3f3b","C5ar1","St3gal6"),
  Ccr2_Mac= c("Ccr2","Cd52","H2-Eb1","H2-Ab1","Bcl2a1a","Coro1a","Cyp4f18","Fxdy5"),
  Prolif_C1q_TLF_Macs= c("Birc5","Mki67","C1qa"),  
  Spp1_Macs= c("Spp1","Trem2","Lipa"), 
  C1q_TLF_Macs= c("C1qa","C1qb","Mrc1","Igf1"),
  Gpnmb= c("Gpnmb","Fabp5","Ctsd","Syngr1")
)

features_all <- unique(unlist(marker_panels_2))
DefaultAssay(seurat_atlas) <- "SCT"

# UMAPs
DimPlot(seurat_atlas, reduction="umap", group.by="lineage_r.06", label=FALSE) +
  ggtitle("Broad lineages (annotated) @ res 0.8")

p_umap <- DimPlot(seurat_atlas, reduction="umap", group.by=new_col, label=TRUE) +
  ggtitle(paste0("UMAP @ res ", res))
print(p_umap)

# DotPlots (split for readability)
n_feat <- length(features_all)
split_idx <- ceiling(n_feat / 2)
features_part1 <- features_all[seq_len(split_idx)]
features_part2 <- if (split_idx < n_feat) features_all[(split_idx+1):n_feat] else character(0)

p_dot1 <- DotPlotSymbols(seurat_atlas,
                         symbols=features_part1,
                         mapping=gene_map,
                         group.by=new_col) +
  ggtitle(paste0("Markers (1/2) @ res ", res))
print(p_dot1)

if (length(features_part2) > 0) {
  p_dot2 <- DotPlotSymbols(seurat_atlas,
                           symbols=features_part2,
                           mapping=gene_map,
                           group.by=new_col) +
    ggtitle(paste0("Markers (2/2) @ res ", res))
  print(p_dot2)
}

# Monocyte markers
features_monocyte <- c("Ccr2","Ly6c2","Cx3cr1","Treml4","Plac8")
DotPlotSymbols(seurat_atlas,
               symbols=features_monocyte,
               mapping=gene_map,
               group.by=new_col) +
  ggtitle("Ly6C^hi vs Ly6C^lo monocyte markers")



```




```{r Endo + Fibroblast annotation}

# Higher resolution clustering
res <- 0.8

if (!"integrated_snn" %in% names(seurat_atlas@graphs)) {
  DefaultAssay(seurat_atlas) <- "integrated"
  seurat_atlas <- FindNeighbors(seurat_atlas, reduction = "pca", dims = 1:30)
}
seurat_atlas <- FindClusters(seurat_atlas, graph.name = "integrated_snn", resolution = res)

# New metadata column
new_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- new_col

# Marker panels
marker_panels_3 <- list(
  LECs             = c("Lyve1","Prox1","Flt4","Nrp2","Cd31","Pdpn","Reln"),
  Immunomodulatory = c("Ptprc","Cdh5"),
  Arterial_ECs     = c("Id1","Mgll"),
  Venous_ECs       = c("Plvap","Vwf"),      
  ECs              = c("Pecam1","Cdh5","Cd36","Fabp4","Eng","Esam","Tler1","Kdr"),
  RCFs             = c("Cthrc1","Acta2","Fn1","Col1a1","Postn","Tagln"),
  Matrifibrocytes  = c("Comp","Sfrp2","Cilp","Eln","Wisp2"), 
  Ccl2_fibroblasts = c("Mt2","Ccl2","Timp1","Mt1","Prg4","Angptl4"),
  Mki67_fibroblasts= c("Stmn1","H2afz","Cks2","Cenpa","Acta2","Hmgb2")
)

# Use SCT for visualization
DefaultAssay(seurat_atlas) <- "SCT"

# Load mapping (Ensembl <-> Symbol)
gene_map <- read.csv("/projects/b1246/Isaac/gene_map_mouse.csv")

# All marker features
features_all <- unique(unlist(marker_panels_3))

# Split into two halves for readability
n_feat <- length(features_all)
split_idx <- ceiling(n_feat / 2)
features_part1 <- features_all[seq_len(split_idx)]
features_part2 <- if (split_idx < n_feat) features_all[(split_idx + 1):n_feat] else character(0)

# Visualize broad lineages
DimPlot(seurat_atlas, reduction="umap", group.by="lineage_r.06", label=FALSE) +
  ggtitle("Broad lineages (annotated) @ res 0.8")

# UMAP of new clusters
p_umap <- DimPlot(seurat_atlas, reduction = "umap", group.by = new_col, label = TRUE) +
  ggtitle(paste0("UMAP @ res ", res))
print(p_umap)

# DotPlots with SYMBOLs (via helper)
p_dot1 <- DotPlotSymbols(
  seurat_atlas,
  symbols = features_part1,
  mapping = gene_map,
  group.by = new_col
) + RotatedAxis() + ggtitle(paste0("Markers (1/2) @ res ", res))
print(p_dot1)

if (length(features_part2) > 0) {
  p_dot2 <- DotPlotSymbols(
    seurat_atlas,
    symbols = features_part2,
    mapping = gene_map,
    group.by = new_col
  ) + RotatedAxis() + ggtitle(paste0("Markers (2/2) @ res ", res))
  print(p_dot2)
}



```


```{r}
## --- Setup
res <- 0.8
res_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res_col
col_lab <- "lineage_r.08"

## --- Mapping (cluster renames and merges)
map_08 <- c(
  `2`  = "MHCII Mac",
  `3`  = "CCR2+ Mac",           # renamed SPP1+
  `7`  = "TLF Mac",             # GPNMB+ under TLF
  `14` = "TLF Mac",             # C1q+ under TLF
  `9`  = "Ly6C-hi Mono",
  `22` = "Ly6C-lo Mono",
  `12` = "B Cells",
  `17` = "ARG1+ Mac",
  `24` = "Prolif TLF Mac",
  `19` = "NKT",
  `20` = "DCs",
  `1`  = "Neutrophils",
  `26` = "Arterial ECs",
  `21` = "Venous ECs",
  `5`  = "ECs",
  `6`  = "RCFs",
  `4`  = "RCFs"
)

cl_ids <- Idents(seurat_atlas)
annot  <- map_08[as.character(cl_ids)]
annot[is.na(annot)] <- "Other"
seurat_atlas@meta.data[[col_lab]] <- annot

## --- Ordered levels (collapsed categories)
lvl_order <- c(
  "Ly6C-hi Mono","Ly6C-lo Mono","Neutrophils","DCs",
  "MHCII Mac","CCR2+ Mac","TLF Mac","ARG1+ Mac","Prolif TLF Mac",
  "B Cells","NKT",
  "Arterial ECs","Venous ECs","ECs","RCFs","Other"
)
vals <- seurat_atlas@meta.data[[col_lab]]
seurat_atlas@meta.data[[col_lab]] <- factor(vals, levels = intersect(lvl_order, unique(vals)))

## --- Updated color palette (collapsed to match new labels)
pal_hc <- c(
  "Ly6C-hi Mono"   = "#D55E00",   # vermillion
  "Ly6C-lo Mono"   = "#E69F00",   # orange
  "Neutrophils"    = "#0072B2",   # blue
  "DCs"            = "#56B4E9",   # sky blue
  "MHCII Mac"      = "#009E73",   # green
  "CCR2+ Mac"      = "#851033",   # renamed
  "TLF Mac"        = "#2C7FB8",   # merged Gpnmb + C1q+
  "ARG1+ Mac"      = "#33A02C",   # green 2
  "Prolif TLF Mac" = "#A6D854",   # light green
  "B Cells"        = "#8C564B",   # brown
  "NKT"            = "#CC79A7",   # purple-pink
  "Arterial ECs"   = "#FB9A99",   # salmon
  "Venous ECs"     = "#B15928",   # sienna
  "ECs"            = "#6A3D9A",   # purple
  "RCFs"           = "#F0E442",   # yellow
  "Other"          = "#BDBDBD"    # grey
)
pal_plot <- pal_hc[levels(seurat_atlas@meta.data[[col_lab]])]

## --- Save metadata
seurat_atlas[[col_lab]] <- NULL  
seurat_atlas <- Seurat::AddMetaData(
  seurat_atlas,
  metadata = setNames(
    data.frame(
      lineage = factor(
        annot,
        levels = lvl_order
      ),
      row.names = colnames(seurat_atlas)
    ),
    col_lab
  )
)

Seurat::Idents(seurat_atlas) <- col_lab

if (is.null(seurat_atlas@misc)) seurat_atlas@misc <- list()
seurat_atlas@misc$annotation_maps <- modifyList(
  seurat_atlas@misc$annotation_maps %||% list(),
  list("r0.8" = map_08)
)
seurat_atlas@misc$palettes <- modifyList(
  seurat_atlas@misc$palettes %||% list(),
  list("lineage_r.08" = pal_hc)
)

## --- Plot
p <- DimPlot(
  seurat_atlas,
  reduction = "umap",
  group.by  = col_lab,
  label     = TRUE,
  repel     = TRUE,
  label.size= 4.2,
  pt.size   = 0.25,
  shuffle   = TRUE
) +
  ggtitle("UMAP — annotated (r = 0.8)") +
  scale_color_manual(values = pal_plot) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 9),
    plot.title   = element_text(size = 18, face = "bold")
  )
print(p)


```



```{r AddModuleScore helper function }

AddModuleScoreSymbols <- function(object, symbols, mapping, name, assay = "SCT", ...) {
  # SYMBOL -> ENSEMBL lookup
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  
  # Map symbols to Ensembl IDs
  features_ens <- symbol_to_ens[toupper(symbols)]
  features_ens <- features_ens[!is.na(features_ens)]
  
  # Keep only features that exist in the assay
  present <- rownames(GetAssayData(object[[assay]], layer = "data"))
  features_ens <- intersect(features_ens, present)
  
  if (length(features_ens) < 2) {
    stop("Not enough mapped features present in assay for module scoring")
  }
  
  AddModuleScore(object, features = list(features_ens), name = name, assay = assay, ...)
}




```




```{r Filtering Apoptotic Cells }

res <- 0.8
res_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res_col

# Defining apoptotic cell markers

apoptotic_markers <- c(
  "Bax","Bak1","Bbc3","Bcl2l11",   
  "Casp3","Casp8","Casp9",         
  "Ddit3","Atf4","Atf3",           
  "Hspa1a","Hspa1b","Jun","Fos"   
)

#Dotplot
p_apop <- DotPlotSymbols(
  seurat_atlas,
  symbols  = apoptotic_markers,
  mapping  = gene_map,
  group.by = res_col
) + RotatedAxis() + ggtitle("Apoptotic marker expression (res 0.8)")
print(p_apop)

#FeaturePlot

FeaturePlotSymbols(
  seurat_atlas,
  symbols  = c("Bax","Casp3","Ddit3"),
  mapping  = gene_map,
  reduction= "umap",
  order    = TRUE,
  raster   = TRUE
)



# Run module scoring with SYMBOLs directly
seurat_atlas <- AddModuleScoreSymbols(
  seurat_atlas,
  symbols  = apoptotic_markers,
  mapping  = gene_map,
  name     = "ApoptosisScores",
  assay    = "SCT"
)



VlnPlot(seurat_atlas,
        features = "ApoptosisScores1",
        group.by = res_col,
        pt.size = 0) +
  ggtitle("Apoptosis module score per cluster (res 0.8)")




```



```{r ReAnnotating Fibroblasts Clusters}

#Fibroblast markers

fibro_panels <- list(
  RCFs             = c("Cthrc1","Acta2","Fn1","Col1a1","Postn","Tagln"),
  Matrifibrocytes  = c("Comp","Sfrp2","Cilp","Eln","Wisp2"), 
  Ccl2_fibroblasts = c("Mt2","Ccl2","Timp1","Mt1","Prg4","Angptl4"),
  Mki67_fibroblasts= c("Stmn1","H2afz","Cks2","Cenpa","Acta2","Hmgb2")
)

#AddModuleScore
for (nm in names(fibro_panels)) {
  seurat_atlas <- AddModuleScoreSymbols(
    seurat_atlas,
    symbols  = fibro_panels[[nm]],
    mapping  = gene_map,
    name     = paste0("Fib_", nm),
    assay    = "SCT"
  )
}

#Violin Plots
fibro_features <- c("Fib_RCFs1","Fib_Matrifibrocytes1",
                    "Fib_Ccl2_fibroblasts1","Fib_Mki67_fibroblasts1")

for (feat in fibro_features) {
  print(
    VlnPlot(seurat_atlas,
            features = feat,
            group.by = res_col,
            pt.size  = 0) +
      ggtitle(paste0("Fibroblast program: ", feat))
  )
}

# FeaturePlots 
FeaturePlot(seurat_atlas,
            features = c("Fib_RCFs1","Fib_Matrifibrocytes1"),
            order    = TRUE, raster = TRUE) + 
  ggtitle("Fibroblast subsets: RCFs & Matrifibrocytes")

FeaturePlot(seurat_atlas,
            features = c("Fib_Ccl2_fibroblasts1","Fib_Mki67_fibroblasts1"),
            order    = TRUE, raster = TRUE) + 
  ggtitle("Fibroblast subsets: Ccl2+ & Mki67+")



```




```{r Differential expression cluster 11,8,23}

# Make sure SCT assay is active
DefaultAssay(seurat_atlas) <- "SCT"

# Prepare SCT models for differential expression
seurat_atlas <- PrepSCTFindMarkers(seurat_atlas)

# Now rerun FindMarkers
markers_11 <- FindMarkers(seurat_atlas, ident.1 = 11, min.pct = 0.25, logfc.threshold = 0.25)
markers_23 <- FindMarkers(seurat_atlas, ident.1 = 23, min.pct = 0.25, logfc.threshold = 0.25)
markers_8  <- FindMarkers(seurat_atlas, ident.1 = 8,  min.pct = 0.25, logfc.threshold = 0.25)


annotate_markers <- function(markers_df, gene_map) {
  markers_df$ensembl_gene_id <- rownames(markers_df)
  out <- merge(markers_df, gene_map, by = "ensembl_gene_id", all.x = TRUE)
  colnames(out)[colnames(out) == "mgi_symbol"] <- "gene_symbol"  # standardize name
  return(out)
}


markers_11_annot <- annotate_markers(markers_11, gene_map)
markers_23_annot <- annotate_markers(markers_23, gene_map)
markers_8_annot  <- annotate_markers(markers_8,  gene_map)

# Preview results
head(markers_11_annot[, c("gene_symbol", "avg_log2FC", "p_val_adj")])
head(markers_23_annot[, c("gene_symbol", "avg_log2FC", "p_val_adj")])
head(markers_8_annot[,  c("gene_symbol", "avg_log2FC", "p_val_adj")])


```





```{r FeaturePlot visualization}

Idents(seurat_atlas) <- "RNA_snn_res.0.8"

FeaturePlot(seurat_atlas, features = c("Acta2", "S100a4", "Fap"),
            label = TRUE)

VlnPlot(seurat_atlas,
        features = c("Acta2", "S100a4", "Fap"),
        group.by = "seurat_clusters", pt.size = 0)


FeaturePlot(seurat_atlas, features = c("Cd68", "Mrc1"),
            label = TRUE)



p_umap <- DimPlot(seurat_atlas, reduction="umap", group.by=new_col, label=TRUE) +
  ggtitle(paste0("UMAP @ res ", res))
print(p_umap)

```





```{r further investigation of cluster 11,23,8}

# Myofibroblast activation panel
FeaturePlot(seurat_atlas, features = c("S100a4", "Fap", "Postn"), label = TRUE)

# VSMC panel
FeaturePlot(seurat_atlas, features = c("Tagln", "Myh11", "Cnn1", "Pdgfrb"), label = TRUE)

# Macrophage panel
FeaturePlot(seurat_atlas, features = c("Cd68", "Mrc1"), label = TRUE)



```





```{r Cluster 8 refinement}

VlnPlot(seurat_atlas, features = c("nCount_RNA","nFeature_RNA","percent.mt"), group.by = "seurat_clusters")


```






```{r Reannotation of UMAP}

## make sure identities match resolution 0.8 clustering
res <- 0.8
res_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res_col

map_08 <- c(
  # Macs & monos 
  `2`  = "MHCII Mac",
  `3`  = "CCR2+ Mac",           
  `7`  = "TLF Mac",             
  `14` = "TLF Mac",             
  `9`  = "Ly6C-hi Mono",
  `22` = "Ly6C-lo Mono",
  `12` = "B Cells",
  `17` = "ARG1+ Mac",
  `24` = "Prolif TLF Mac",     
  `19` = "NKT",
  `20` = "DCs",
  `1`  = "Neutrophils",
  
  # Endothelial 
  `26` = "Arterial ECs",        
  `21` = "Venous ECs",
  `5`  = "ECs",
  
  # Fibroblast lineage
  `16` = "Fibroblast progenitors",    
  `0`  = "Matrifibrocytes",
  `15` = "Ccl2 fibroblasts",
  `13` = "Ccl2 fibroblasts",
  `4`  = "Transitional fibroblasts", 
  `6`  = "VSMCs",                
  `10` = "RCFs",                      
  `25` = "Prolif RCFs",              
  
  # Stromal branches
  `11` = "Transitional MMT",     
  `8`  = "Cardiomyocytes",
  `23` = "Fibroblasts"          
)

## Apply mapping
col_lab <- "lineage_r.08"
cl_ids  <- Idents(seurat_atlas)
annot   <- map_08[as.character(cl_ids)]
annot[is.na(annot)] <- "Unassigned stromal"
seurat_atlas@meta.data[[col_lab]] <- annot

## Ordered levels
lvl_order <- c(
  # monos/macrophages
  "Ly6C-hi Mono","Ly6C-lo Mono","Neutrophils","DCs",
  "MHCII Mac","CCR2+ Mac","TLF Mac","ARG1+ Mac","Prolif TLF Mac",
  "B Cells","NKT",
  # endothelium
  "Arterial ECs","Venous ECs","ECs",
  # fibroblast programs
  "Fibroblast progenitors",
  "RCFs","Transitional fibroblasts","Prolif RCFs",
  "Matrifibrocytes",
  "Ccl2 fibroblasts",
  # stromal
  "Transitional MMT","Fibroblasts","VSMCs",
  "Cardiomyocytes",
  "Unassigned"
)
vals <- seurat_atlas@meta.data[[col_lab]]
seurat_atlas@meta.data[[col_lab]] <- factor(vals, levels = intersect(lvl_order, unique(vals)))

## Palette
pal_hc <- c(
  # monos/macrophages
  "Ly6C-hi Mono"   = "#D55E00",
  "Ly6C-lo Mono"   = "#E69F00",
  "Neutrophils"    = "#0072B2",
  "DCs"            = "#56B4E9",
  "MHCII Mac"      = "#009E73",
  "CCR2+ Mac"      = "#851033",
  "TLF Mac"        = "#2C7FB8",
  "ARG1+ Mac"      = "#33A02C",
  "Prolif TLF Mac" = "#A6D854",
  "B Cells"        = "#8C564B",
  "NKT"            = "#CC79A7",
  # endothelium
  "Arterial ECs"   = "#FB9A99",
  "Venous ECs"     = "#B15928",
  "ECs"            = "#6A3D9A",
  # fibroblasts
  "Fibroblast progenitors" = "#D13F21",
  "RCFs"                   = "#F0E442",
  "Transitional fibroblasts" = "#FFD700",
  "Prolif RCFs"            = "#FFB347",
  "Matrifibrocytes"        = "#FF7F00",
  "Ccl2 fibroblasts"       = "#800080",
  # stromal refinements
  "Transitional MMT"       = "#20B2AA",   # teal
  "Fibroblasts"            = "#A0522D",   # brown
  "VSMCs"                  = "#6421D1",   # deep violet
  "Cardiomyocytes"= "#7A9134"
)

pal_plot <- pal_hc[levels(seurat_atlas@meta.data[[col_lab]])]

p <- DimPlot(
  seurat_atlas,
  reduction = "umap",         
  group.by  = col_lab,
  label     = TRUE,
  repel     = TRUE,
  label.size= 4,
  pt.size   = 0.25,
  shuffle   = TRUE
) +
  ggtitle("UMAP — annotated (r = 0.8)") +
  scale_color_manual(values = pal_plot) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 9),
    plot.title   = element_text(size = 18, face = "bold")
  )

print(p)


```




```{r}

## --- Macrophage/monocyte identities (match annotation above)
mac_ids <- c("Ly6C-hi Mono","Ly6C-lo Mono",
             "MHCII Mac","CCR2+ Mac","TLF Mac",
             "ARG1+ Mac","Prolif TLF Mac")

## --- Extract cells belonging to macrophage/monocyte groups
mac_cells <- rownames(
  seurat_atlas@meta.data[
    seurat_atlas@meta.data[[col_lab]] %in% mac_ids, ]
)

## --- Subset object directly by those cells
mac_obj <- subset(seurat_atlas, cells = mac_cells)

## --- Clean factor levels
mac_obj[[col_lab]] <- droplevels(mac_obj[[col_lab]])

## --- Macrophage palette
pal_mac <- pal_hc[mac_ids]

## --- Plot macrophage-only UMAP
p_mac <- DimPlot(
  mac_obj,
  reduction = "umap",
  group.by  = col_lab,
  label     = TRUE,
  repel     = TRUE,
  label.size= 4,
  pt.size   = 0.3,
  shuffle   = TRUE
) +
  ggtitle("UMAP — macrophage & monocyte subsets (r = 0.8)") +
  scale_color_manual(values = pal_mac) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 9),
    plot.title   = element_text(size = 16, face = "bold")
  )

print(p_mac)



```



####SAVING OBJECT######
```{r}
saveRDS(seurat_atlas, "/projects/b1246/Isaac/seurat_atlas.rds")

```


###########################################

# CREATING THE SINGLE R COMPATIBLE OBJECT

###########################################

```{r SINGLE R ATLAS CREATION}

## --- Setup
Atlas <- seurat_atlas
DefaultAssay(Atlas) <- "RNA"

## --- Extract raw counts from Seurat v5 Assay5
expr_counts <- Atlas[["RNA"]]$counts

## --- Ensure unique gene names
expr_counts <- expr_counts[!duplicated(rownames(expr_counts)), , drop = FALSE]

## --- Align labels with counts
cells_counts <- colnames(expr_counts)              # cells in counts
cells_meta   <- rownames(Atlas@meta.data)          # cells in metadata

# Subset metadata to same cells and order
labels <- as.character(Atlas@meta.data[cells_counts, "lineage_r.08"])

## --- Keep only valid cells
keep <- !is.na(labels) & labels != ""
expr_counts <- expr_counts[, keep, drop = FALSE]
labels <- labels[keep]

## --- Build SingleCellExperiment
library(SingleCellExperiment)
sce_reference <- SingleCellExperiment(assays = list(counts = expr_counts))
colData(sce_reference)$label <- labels

## --- Filter: keep labels with >=20 cells
cell_counts <- table(labels)
good_labels <- names(cell_counts[cell_counts >= 20])
sce_reference <- sce_reference[, labels %in% good_labels]
labels <- colData(sce_reference)$label

## --- Aggregate to pseudobulk by label
library(scuttle)
sce_reference_pb <- aggregateAcrossCells(
  sce_reference,
  ids = labels
)

## --- Normalize (adds logcounts)
library(scater)
sce_reference_pb <- logNormCounts(sce_reference_pb)

## --- Sanity checks
dim(sce_reference_pb)              # genes × pseudobulk samples
colnames(sce_reference_pb)         # pseudobulk labels kept
assayNames(sce_reference_pb)       # "counts", "logcounts"
table(labels)                      # cell counts per label (post-filter)

## --- Barplot (after filtering)
barplot(table(labels), las = 2, cex.names = 0.7,
        main = "Cells per pseudobulk label (≥20 filter)", col = "steelblue")



```

#Saving reference

```{r}
saveRDS(sce_reference_pb,
        file = "/scratch/inm3056/atlas_reference.rds")
```




#############################################
ANNOTATING MERTK ADULT SAMPLES
ANNOTATING MERTK ADULT SAMPLES 
ANNOTATING MERTK ADULT SAMPLES 
#############################################



Annotating the MERTK adult conditions using the cardiac reference atlas
```{r Seurat Harmony --> Annotate with Atlas}

## --- Subset to Adult
MERTK_adult <- subset(seurat_harmony, subset = age_group == "Adult")

## --- Dependencies
library(SingleCellExperiment)
library(Matrix)
library(SingleR)

## --- 1) Extract KO + WT cell identities
cells_KO <- colnames(MERTK_adult)[MERTK_adult$genotype == "KO"]
cells_WT <- colnames(MERTK_adult)[MERTK_adult$genotype == "WT"]

## --- 2) Extract KO + WT matrices from layers
gene_names <- rownames(MERTK_adult[["RNA"]])
counts_KO  <- MERTK_adult[["RNA"]]@layers[["counts.KO_Adult"]]
counts_WT  <- MERTK_adult[["RNA"]]@layers[["counts.WT_Adult"]]

# Attach gene names and matching cell barcodes
rownames(counts_KO) <- gene_names
rownames(counts_WT) <- gene_names
colnames(counts_KO) <- cells_KO
colnames(counts_WT) <- cells_WT

# Combine
expr_counts <- cbind(counts_KO, counts_WT)
expr_counts <- expr_counts[!duplicated(rownames(expr_counts)), , drop = FALSE]
cat("Counts matrix:", nrow(expr_counts), "genes ×", ncol(expr_counts), "cells\n")

## --- 3) Align with metadata
# Metadata already uses cell barcodes like "AAACCTGAGACTTTCG-1"
common <- intersect(colnames(expr_counts), rownames(MERTK_adult@meta.data))
length(common)

# Subset and ensure order
expr_counts <- expr_counts[, common, drop = FALSE]
labels_meta <- MERTK_adult@meta.data[common, , drop = FALSE]
expr_counts <- as(expr_counts, "dgCMatrix")
stopifnot(all(colnames(expr_counts) == rownames(labels_meta)))

## --- 4) Build SingleCellExperiment
MERTK_adult_sce <- SingleCellExperiment(
  assays  = list(counts = expr_counts),
  colData = labels_meta
)
cat("Built SCE with:", nrow(MERTK_adult_sce), "genes ×", ncol(MERTK_adult_sce), "cells\n")


## --- 5) Harmonize gene identifiers
gene_map <- gene_map_mouse
gene_map <- gene_map[!duplicated(gene_map$ensembl_gene_id), ]

convert_ids <- function(ids) {
  mapped <- gene_map$mgi_symbol[match(ids, gene_map$ensembl_gene_id)]
  mapped[is.na(mapped)] <- ids[is.na(mapped)]
  mapped
}

rownames(atlas_reference) <- convert_ids(rownames(atlas_reference))
rownames(MERTK_adult_sce) <- convert_ids(rownames(MERTK_adult_sce))

atlas_reference <- atlas_reference[!duplicated(rownames(atlas_reference)) & !is.na(rownames(atlas_reference)), ]
MERTK_adult_sce <- MERTK_adult_sce[!duplicated(rownames(MERTK_adult_sce)) & !is.na(rownames(MERTK_adult_sce)), ]

## --- 6) Check overlap
common_genes <- intersect(rownames(MERTK_adult_sce), rownames(atlas_reference))
cat("Overlapping genes:", length(common_genes), "\n")

## --- 7) Restrict to common genes
MERTK_adult_sce <- MERTK_adult_sce[common_genes, ]
atlas_reference <- atlas_reference[common_genes, ]

#Adding logcounts
library(scuttle)
MERTK_adult_sce <- logNormCounts(MERTK_adult_sce)


## --- 8) Run SingleR
pred_MERTK_adult <- SingleR(
  test   = MERTK_adult_sce,
  ref    = atlas_reference,
  labels = colnames(atlas_reference)
)

## --- 9) Add SingleR labels back into Seurat object
MERTK_adult$SingleR_label <- NA
MERTK_adult$SingleR_label[common] <- pred_MERTK_adult$labels

## --- 10) Visualize
library(Seurat)
library(ggplot2)

p <- DimPlot(
  MERTK_adult,
  reduction = "umap",
  group.by  = "SingleR_label",
  label     = TRUE,
  repel     = TRUE,
  label.size = 5,
  pt.size   = 0.5
) +
  ggtitle("Atlas Reference Annotation — Adult Only") +
  theme_minimal(base_size = 16) +
  theme(
    plot.title      = element_text(size = 22, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.text     = element_text(size = 14),
    legend.title    = element_blank(),
    axis.title      = element_text(size = 16),
    axis.text       = element_text(size = 12)
  )

p_atlas_umap <- p

print(p_atlas_umap)
```




```{r}
saveRDS(MERTK_adult,
        file = "/projects/b1246/Isaac/Mertk_adult.rds")
```




```{r}
ggsave(
  filename = "MERTK_Adult_AtlasAnnotation_UMAP.pdf",
  path = "/projects/p31275/Isaac/MERTK/Figures",
  width = 14, height = 8, dpi = 300
)
```




Annotating the Adult samples (KO and WT) with Immgen
```{r Seurat Harmony --> Annotate with ImmGen}

library(scuttle)
MERTK_adult_sce <- logNormCounts(MERTK_adult_sce)

library(celldex)
immgen_ref <- celldex::ImmGenData()
immgen_ref

common_genes <- intersect(rownames(MERTK_adult_sce), rownames(immgen_ref))
length(common_genes)

##Restricting both datasets to shared genes
MERTK_adult_sce <- MERTK_adult_sce[common_genes, ]
immgen_ref <- immgen_ref[common_genes, ]

##Run SingleR using ImmGen reference
pred_immgen <- SingleR(
  test = MERTK_adult_sce,
  ref = immgen_ref,
  labels = immgen_ref$label.main,     
  assay.type.test = "logcounts",      
  assay.type.ref = "logcounts"        
)

##Inspect results
table(pred_immgen$labels)

##Add results back to Seurat object
MERTK_adult$ImmGen_label <- pred_immgen$labels

##Visualization
library(ggplot2)
DimPlot(
  MERTK_adult,
  reduction = "umap",
  group.by = "ImmGen_label",
  label = TRUE,
  repel = TRUE,
  label.size = 4,
  pt.size = 0.4
) +
  ggtitle("ImmGen Reference Annotation — Adult Only") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 12),
    legend.title = element_blank()
  )

##Save results
ggsave(
  filename = "MERTK_Adult_ImmGenAnnotation_UMAP.pdf",
  path = "/projects/b1246/Isaac/MERTK/Figures",
  width = 14, height = 8, dpi = 300
)



```


```{r}
saveRDS(MERTK_adult,
        file = "/projects/b1246/Isaac/Mertk_adult.rds")
```



```{r}
saveRDS(MERTK_adult_sce,
        file = "/scratch/inm3056/mertk_adult_sce.rds")
```











##########################
ATLAS VS IMMGEN REFERENCE
#########################



```{r ATLAS vs IMMGEN Reference Annotation}

library(dplyr)
library(pheatmap)

## --- Filter cells with valid ImmGen annotations
valid <- !(is.na(MERTK_adult$ImmGen_label) | MERTK_adult$ImmGen_label == "")
pred_immgenMERTK_adult_filtered <- subset(MERTK_adult, cells = colnames(MERTK_adult)[valid])

## --- Step 1: Mapping dictionaries
atlas_to_immgen <- list(
  "Macrophages"      = c("CCR2+ Mac","ARG1+ Mac","MHCII Mac","TLF Mac"),
  "Monocytes"        = c("Ly6C-hi Mono","Ly6C-lo Mono"),
  "Neutrophils"      = c("Neutrophils"),
  "Fibroblasts"      = c("RCFs","Transitional fibroblasts","Matrifibrocytes",
                         "Ccl2 fibroblasts","Fibroblasts","Fibroblast progenitors"),
  "Endothelial cells"= c("ECs","Venous ECs"),
  "Lymphoid"         = c("B Cells","T Cells","NKT","DCs"),
  "Stromal cells"    = c("VSMCs","Transitional MMT","Stromal (perivascular?)"),
  "Cardiomyocytes"   = c("Cardiomyocytes")
)

## Mapping each label to its ImmGen equivalent
map_label <- function(x, map) {
  for (key in names(map)) {
    if (x %in% map[[key]]) return(key)
  }
  return("Other")
}

atlas_broad  <- sapply(MERTK_adult_filtered$SingleR_label, map_label, map = atlas_to_immgen)

## --- Step 3: Normalize ImmGen labels to same naming scheme
immgen_broad <- MERTK_adult_filtered$ImmGen_label
immgen_broad[grepl("B cell", immgen_broad, ignore.case = TRUE)] <- "Lymphoid"
immgen_broad[grepl("T cell|NK|NKT|DC|ILC|Tgd", immgen_broad, ignore.case = TRUE)] <- "Lymphoid"
immgen_broad[grepl("Macroph", immgen_broad, ignore.case = TRUE)] <- "Macrophages"
immgen_broad[grepl("Monocyte", immgen_broad, ignore.case = TRUE)] <- "Monocytes"
immgen_broad[grepl("Neutroph", immgen_broad, ignore.case = TRUE)] <- "Neutrophils"
immgen_broad[grepl("Fibro", immgen_broad, ignore.case = TRUE)] <- "Fibroblasts"
immgen_broad[grepl("Endoth", immgen_broad, ignore.case = TRUE)] <- "Endothelial cells"
immgen_broad[grepl("Strom", immgen_broad, ignore.case = TRUE)] <- "Stromal cells"
immgen_broad[grepl("Cardio|Myo", immgen_broad, ignore.case = TRUE)] <- "Cardiomyocytes"

##Confusion matrix
mat <- prop.table(table(Atlas = atlas_broad, ImmGen = immgen_broad), 1)

## Reordering the matrix
mat <- mat[rowSums(mat) > 0, colSums(mat) > 0, drop = FALSE]
order <- c("Macrophages","Monocytes","Neutrophils","Fibroblasts",
           "Endothelial cells","Stromal cells","Lymphoid","Cardiomyocytes")
mat <- mat[intersect(order, rownames(mat)), intersect(order, colnames(mat)), drop = FALSE]



##HEATMAP COMPARISON



pheatmap(
  mat,
  color = colorRampPalette(c("white","firebrick3"))(100),
  main = "Broad Lineage Agreement — Atlas vs. ImmGen",
  display_numbers = TRUE,
  cluster_rows = FALSE, cluster_cols = FALSE,
  fontsize = 12, angle_col = 45
)



summary_df <- as.data.frame(mat) %>%
  group_by(Atlas) %>%
  summarize(
    TopMatch  = ImmGen[which.max(Freq)],
    Agreement = round(max(Freq) * 100, 1)
  ) %>%
  arrange(desc(Agreement))
print(summary_df)


```



###### IMMGEN vs ATLAS ANNOTATION DISCORDANCE ######


A/ Sensitivity, precision, F1 score 
```{r ImmGen Atlas Discordance}

# Precision per atlas cell type
calculate_precision <- function(cell_type) {
  
  immgen_calls_mac <- sum(MERTK_adult_filtered$ImmGen_label=="Macrophages")
  
  #Identifying how many cells are actually this cell type
  atlas_is_type <- MERTK_adult_filtered$SingleR_label == cell_type
  true_positives <- sum(immgen_calls_mac & atlas_is_type)
  precision <- true_positives/ immgen_calls_mac
  return(precision)
}
#Creating Validation metrics table
validation_metrics <- data.frame(
  Cell_Type = c("CCR2+ Mac", "ARG1+ Mac", "MHCII Mac", "TLF Mac", "Ly6c-hi Mono", "Ly6C-lo Mono", "B Cells", "VSMCs"),
  Sensitivity= NA,
  Precision = NA,
  F1_Score = NA
)
#Filling in the metrics
for (i in 1:nrow(validation_metrics)) {
  cell_type <- validation_metrics$Cell_Type[i]
  atlas_cells <- MERTK_adult_filtered$SingleR_label == cell_type
  
  if (grepl("Mac", cell_type)) {
    expected_label <- "Macrophages"
  } else if (grepl("Mono", cell_type)) {
    expected_label <- "Monocytes"
  } else {
      expected_label <- cell_type
  }
    
    #Calculation sensitivity: Of my true cell type, what percentage did ImmGen correctly identify?
    total_true <- sum(atlas_cells)
    correctly_called <- sum(atlas_cells & 
                              MERTK_adult_filtered$ImmGen_label == expected_label)
    validation_metrics$Sensitivity[i] <- correctly_called / total_true
    
    #Calculating Precision: When ImmGen makes a prediction, how often is it correct?
    immgen_calls <- MERTK_adult_filtered$ImmGen_label == expected_label
    true_positives <- sum(atlas_cells & immgen_calls)
    validation_metrics$Precision[i] <- true_positives / sum(immgen_calls)
    #F1 score 
    P <- validation_metrics$Precision[i]
    R <- validation_metrics$Sensitivity[i]
    validation_metrics$F1_Score[i] <- 2 * (P * R) / (P + R)  
  }
print(validation_metrics)
```







B/ Imgenn Labels Atlas receives
```{r ATLAS and IMMGEN Comparison Follow-up}

#Identify Atlas SingleR annotations
table(MERTK_adult_filtered$SingleR_label)

# ImmGen labels assigned to atlas macrophages
mac_cells <- MERTK_adult_filtered$SingleR_label %in% 
  c("CCR2+ Mac","ARG1+ Mac","MHCII Mac","TLF Mac")
table(MERTK_adult_filtered$ImmGen_label[mac_cells])


# Which atlas cell types is ImmGen calling "Macrophages"?
immgen_mac_cells <- MERTK_adult_filtered$ImmGen_label == "Macrophages"
table(MERTK_adult_filtered$SingleR_label[immgen_mac_cells])

# Are monocytes being called macrophages by ImmGen?
mono_cells <- MERTK_adult_filtered$SingleR_label %in% 
  c("Ly6C-hi Mono", "Ly6C-lo Mono")
table(MERTK_adult_filtered$ImmGen_label[mono_cells])


subtypes <- c("CCR2+ Mac", "ARG1+ Mac", "MHCII Mac", "TLF Mac")
for (subtype in subtypes) {
  cells <- MERTK_adult_filtered$SingleR_label == subtype
  cat("\n", subtype, ":\n")
  print(table(MERTK_adult_filtered$ImmGen_label[cells]))
}



```








C/ Figures 


Plot 1 - Labels Imgenn calls macrophages 
```{r}

library(ggplot2)
library(dplyr)
library(forcats)
library(scales)

# Extracting composition of ImmGen's macrophages 

immgen_mac_cells <- MERTK_adult_filtered$ImmGen_label == "Macrophages"
total_immgen_macs <- sum(immgen_mac_cells)  

composition_raw <- table(MERTK_adult_filtered$SingleR_label[immgen_mac_cells])

composition_df <- as.data.frame(composition_raw, stringsAsFactors = FALSE)
colnames(composition_df) <- c("Cell_Type", "Count") 
composition_df <- composition_df %>%
  mutate(Pct = Count / total_immgen_macs * 100) %>%
  arrange(desc(Count))

#collapsing low-frequency occurences into "other" category and show prevalent misclassifications
TOP_N <- 8
top_types <- composition_df %>% slice_head(n = TOP_N)
other_types <- composition_df %>% slice_tail(n = nrow(composition_df) - TOP_N)

other_row <- data.frame(
  Cell_Type = factor("Other"),
  Count = sum(other_types$Count),
  Pct   = sum(other_types$Pct)
)

plot_df <- bind_rows(top_types, other_row) %>%
  mutate(
    Cell_Type = as.character(Cell_Type),
# Broad category per label
Category = case_when(
  grepl("Mac",                   Cell_Type) ~ "Macrophage",
  grepl("Mono",                  Cell_Type) ~ "Monocyte",
  Cell_Type == "B Cells"                    ~ "B Cell",        
  Cell_Type == "VSMCs"                      ~ "VSMC",          
  Cell_Type %in% c("NKT")                  ~ "NKT Cell",      
  grepl("Fibro|MMT|RCF|Matrifi", Cell_Type) ~ "Fibroblast",
  grepl("EC|Venous|Endo",        Cell_Type) ~ "Endothelial",
  Cell_Type == "Cardiomyocytes"             ~ "Cardiomyocyte",
  Cell_Type == "Other"                      ~ "Other",
  TRUE                                      ~ "Other Immune"   
),
    # forcing macrophage subtypes into one display label 
    Display_Label = case_when(
      grepl("Mac", Cell_Type) ~ "Atlas Macrophages\n(CCR2+, ARG1+, MHCII, TLF)",
      TRUE ~ Cell_Type
    )
  )

# collapsing atlas mac subtypes into one bar
plot_df_collapsed <- plot_df %>%
  group_by(Display_Label, Category) %>%
  summarise(Count = sum(Count), Pct = sum(Pct), .groups = "drop") %>%
  arrange(desc(Count)) %>%
  mutate(
    Display_Label = fct_reorder(Display_Label, Count),
    # Bar labels
    Bar_Label = paste0(Count, "\n(", round(Pct, 1), "%)")
  )


category_colors <- c(
  "Macrophage"    = "#4D9E6F",   # muted green — correct
  "Monocyte"      = "#E07B39",   # orange
  "B Cell"        = "#3A6BAD",   # blue — lymphoid
  "VSMC"          = "#C0392B",   # red — non-immune
  "NKT Cell"      = "#C75BAD",   # magenta — lymphoid
  "Fibroblast"    = "#8B6BB1",   # purple — stromal
  "Endothelial"   = "#2B9CB5",   # teal
  "Cardiomyocyte" = "#8B1A1A",   # dark red
  "Other Immune"  = "#999999",
  "Other"         = "#BBBBBB"
)

#Plotting 
p <- ggplot(plot_df_collapsed,
            aes(x = Display_Label,
                y = Pct,
                fill = Category)) +

  geom_col(width = 0.65, colour = "black", linewidth = 0.4) +  

  geom_text(aes(label = Bar_Label),
            hjust  = -0.08,
            size   = 3.5,
            colour = "black",
            fontface = "plain",
            family = "Helvetica") +

  coord_flip(clip = "off",
             ylim = c(0, max(plot_df_collapsed$Pct) * 1.32)) +

  scale_fill_manual(values = category_colors, guide = "none") +

  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    breaks = seq(0, 40, by = 10),
    expand = expansion(mult = c(0, 0))
  ) +

  labs(
    title    = NULL,          
    subtitle = NULL,
    x = NULL,
    y = "% of ImmGen \"Macrophages\""
  ) +

  theme_classic(base_size = 13, base_family = "Helvetica") +
  theme(
    plot.background  = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),

    # Axis lines
    axis.line.x = element_line(colour = "black", linewidth = 0.8),
    axis.line.y = element_line(colour = "black", linewidth = 0.8),

    axis.ticks.x = element_line(colour = "black", linewidth = 0.6),
    axis.ticks.y = element_blank(),

    axis.ticks.length = unit(4, "pt"),

    # Axis text
    axis.text.y  = element_text(colour = "black", size = 12, face = "bold",
                                hjust = 1, margin = margin(r = 4)),
    axis.text.x  = element_text(colour = "black", size = 11),
    axis.title.x = element_text(colour = "black", size = 12, face = "bold",
                                margin = margin(t = 8)),

    # No gridlines inside plot
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    plot.margin = margin(15, 70, 12, 12)
  )

print(p)


ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/ImmGen_Macrophage_Composition.png",
       plot = p, width = 7, height = 5, dpi = 300, bg = "white")  



```


# Plot 2 - Marker Validation, side by side umaps

Single-R label Umaps
```{r}

library(ggplot2)
library(Seurat)
library(patchwork)
library(dplyr)


# Extracting Umap coordinates and metadata
umap_coords <- as.data.frame(Embeddings(MERTK_adult_filtered, reduction = "umap"))
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")

umap_data <- cbind(umap_coords, 
                   SingleR_label = MERTK_adult_filtered$SingleR_label,
                   ImmGen_label  = MERTK_adult_filtered$ImmGen_label)

# Cell categories UMAP

# Panel A: B Cells
umap_data_bcells <- umap_data %>%
  mutate(
    Highlight = case_when(
      SingleR_label == "B Cells" ~ "B Cells",
      TRUE ~ "Other"
    ),
    Highlight = factor(Highlight, levels = c("Other", "B Cells"))  
  )

# Panel B: VSMCs  
umap_data_vsmcs <- umap_data %>%
  mutate(
    Highlight = case_when(
      SingleR_label == "VSMCs" ~ "VSMCs",
      TRUE ~ "Other"
    ),
    Highlight = factor(Highlight, levels = c("Other", "VSMCs"))
  )

# Panel C: Ly6C-hi Monocytes
umap_data_mono <- umap_data %>%
  mutate(
    Highlight = case_when(
      SingleR_label == "Ly6C-hi Mono" ~ "Ly6C-hi Mono",
      TRUE ~ "Other"
    ),
    Highlight = factor(Highlight, levels = c("Other", "Ly6C-hi Mono"))
  )

# Color palettes
color_bcells <- c("Other" = "#E5E5E5", "B Cells"      = "#3A6BAD")  
color_vsmcs  <- c("Other" = "#E5E5E5", "VSMCs"        = "#C0392B")  
color_mono   <- c("Other" = "#E5E5E5", "Ly6C-hi Mono" = "#E07B39")  

#Theme for all panels
umap_theme <- theme_classic(base_size = 11, base_family = "Helvetica") +
  theme(
    plot.background  = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),
    
    # axes
    axis.line   = element_line(colour = "black", linewidth = 0.6),
    axis.ticks  = element_line(colour = "black", linewidth = 0.4),
    axis.text   = element_text(colour = "black", size = 9),
    axis.title  = element_text(colour = "black", size = 10, face = "bold"),
    
    # Title
    plot.title = element_text(colour = "black", size = 11, face = "bold", hjust = 0.5),
    
    # Legend
    legend.position = "none",  # remove legend since cell type is in title
    
    plot.margin = margin(10, 10, 10, 10)
  )

# Panel A: B Cells 
p_bcells <- ggplot(umap_data_bcells, aes(x = UMAP_1, y = UMAP_2, color = Highlight)) +
  geom_point(size = 0.3, alpha = 0.8) +
  scale_color_manual(values = color_bcells) +
  labs(title = "B Cells", x = "UMAP 1", y = "UMAP 2") +
  umap_theme

#Panel B: VSMCs
p_vsmcs <- ggplot(umap_data_vsmcs, aes(x = UMAP_1, y = UMAP_2, color = Highlight)) +
  geom_point(size = 0.3, alpha = 0.8) +
  scale_color_manual(values = color_vsmcs) +
  labs(title = "VSMCs", x = "UMAP 1", y = "UMAP 2") +
  umap_theme

# Panel C: Ly6C-hi Monocytes
p_mono <- ggplot(umap_data_mono, aes(x = UMAP_1, y = UMAP_2, color = Highlight)) +
  geom_point(size = 0.3, alpha = 0.8) +
  scale_color_manual(values = color_mono) +
  labs(title = "Ly6C-hi Monocytes", x = "UMAP 1", y = "UMAP 2") +
  umap_theme

# Combine into one row
p_combined <- p_bcells | p_vsmcs | p_mono

print(p_combined)

#Save 
ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/UMAP_Atlas_Annotations.png",
       plot = p_combined, width = 12, height = 4, dpi = 300, bg = "white")

ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/UMAP_Atlas_Annotations.pdf",
       plot = p_combined, width = 12, height = 4, device = cairo_pdf)

#Cell counts
cat("\n──Cell counts for highlighted populations ──\n")
cat(sprintf("B Cells:        %d cells\n", sum(umap_data$SingleR_label == "B Cells")))
cat(sprintf("VSMCs:          %d cells\n", sum(umap_data$SingleR_label == "VSMCs")))
cat(sprintf("Ly6C-hi Mono:   %d cells\n", sum(umap_data$SingleR_label == "Ly6C-hi Mono")))


```

ImmGen labels Umaps
```{r}

library(ggplot2)
library(Seurat)
library(patchwork)
library(dplyr)


#Extracting umap coordinates
umap_coords <- as.data.frame(Embeddings(MERTK_adult_filtered, reduction = "umap"))
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")

umap_data <- cbind(umap_coords, 
                   SingleR_label = MERTK_adult_filtered$SingleR_label,
                   ImmGen_label  = MERTK_adult_filtered$ImmGen_label)


# Panel A: B Cells misclassified as Macrophages
umap_data_bcells <- umap_data %>%
  mutate(
    Highlight = case_when(
      SingleR_label == "B Cells" & ImmGen_label == "Macrophages" ~ "B Cells → 'Macrophages'",
      SingleR_label == "B Cells" & ImmGen_label != "Macrophages" ~ "B Cells",
      TRUE ~ "Other"
    ),
    Highlight = factor(Highlight, 
                      levels = c("Other", "B Cells", "B Cells → 'Macrophages'"))
  )

# Panel B: VSMCs misclassified as Macrophages
umap_data_vsmcs <- umap_data %>%
  mutate(
    Highlight = case_when(
      SingleR_label == "VSMCs" & ImmGen_label == "Macrophages" ~ "VSMCs → 'Macrophages'",
      SingleR_label == "VSMCs" & ImmGen_label != "Macrophages" ~ "VSMCs",
      TRUE ~ "Other"
    ),
    Highlight = factor(Highlight,
                      levels = c("Other", "VSMCs", "VSMCs → 'Macrophages'"))
  )

# Panel C: Ly6C-hi Monocytes misclassified as Macrophages
umap_data_mono <- umap_data %>%
  mutate(
    Highlight = case_when(
      SingleR_label == "Ly6C-hi Mono" & ImmGen_label == "Macrophages" ~ "Ly6C-hi Mono → 'Macrophages'",
      SingleR_label == "Ly6C-hi Mono" & ImmGen_label != "Macrophages" ~ "Ly6C-hi Mono",
      TRUE ~ "Other"
    ),
    Highlight = factor(Highlight,
                      levels = c("Other", "Ly6C-hi Mono", "Ly6C-hi Mono → 'Macrophages'"))
  )

#Color palettes (Red for flagged errors)

color_bcells <- c(
  "Other" = "#E5E5E5",
  "B Cells" = "#7AA3CC",          
  "B Cells → 'Macrophages'" = "#D32F2F"    
)

color_vsmcs <- c(
  "Other" = "#E5E5E5",
  "VSMCs (correct)" = "#D97B6F",           
  "VSMCs → 'Macrophages'" = "#D32F2F"      
)

color_mono <- c(
  "Other" = "#E5E5E5",
  "Ly6C-hi Mono (correct)" = "#F4AA73",     
  "Ly6C-hi Mono → 'Macrophages'" = "#D32F2F"  
)

# Base theme for each umap
umap_theme <- theme_classic(base_size = 11, base_family = "Helvetica") +
  theme(
    plot.background  = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),
    
    axis.line   = element_line(colour = "black", linewidth = 0.6),
    axis.ticks  = element_line(colour = "black", linewidth = 0.4),
    axis.text   = element_text(colour = "black", size = 9),
    axis.title  = element_text(colour = "black", size = 10, face = "bold"),
    
    plot.title = element_text(colour = "black", size = 11, face = "bold", hjust = 0.5),
    
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm"),
    
    plot.margin = margin(10, 5, 10, 10)
  )

#Panel A: B Cells
p_bcells <- ggplot(umap_data_bcells, aes(x = UMAP_1, y = UMAP_2, color = Highlight)) +
  geom_point(size = 0.3, alpha = 0.8) +
  scale_color_manual(values = color_bcells) +
  labs(title = "B Cells (ImmGen Labels)", x = "UMAP 1", y = "UMAP 2") +
  umap_theme

#Panel B: VSMCs
p_vsmcs <- ggplot(umap_data_vsmcs, aes(x = UMAP_1, y = UMAP_2, color = Highlight)) +
  geom_point(size = 0.3, alpha = 0.8) +
  scale_color_manual(values = color_vsmcs) +
  labs(title = "VSMCs (ImmGen Labels)", x = "UMAP 1", y = "UMAP 2") +
  umap_theme

#Panel C: Ly6C-hi Monocytes
p_mono <- ggplot(umap_data_mono, aes(x = UMAP_1, y = UMAP_2, color = Highlight)) +
  geom_point(size = 0.3, alpha = 0.8) +
  scale_color_manual(values = color_mono) +
  labs(title = "Ly6C-hi Mono (ImmGen Labels)", x = "UMAP 1", y = "UMAP 2") +
  umap_theme

#Combine
p_combined <- p_bcells | p_vsmcs | p_mono

print(p_combined)

#Save
ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/UMAP_ImmGen_Misclassification.png",
       plot = p_combined, width = 14, height = 4, dpi = 300, bg = "white")

ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/UMAP_ImmGen_Misclassification.pdf",
       plot = p_combined, width = 14, height = 4, device = cairo_pdf)

#Quantify the misclassification
cat("\n── ImmGen Misclassification Statistics ──\n\n")

# B Cells
bcell_total <- sum(umap_data$SingleR_label == "B Cells")
bcell_miscalled <- sum(umap_data$SingleR_label == "B Cells" & 
                       umap_data$ImmGen_label == "Macrophages")
cat(sprintf("B Cells:\n"))
cat(sprintf("  Total: %d cells\n", bcell_total))
cat(sprintf("  Called 'Macrophages' by ImmGen: %d (%.1f%%)\n\n", 
            bcell_miscalled, 100 * bcell_miscalled / bcell_total))

# VSMCs
vsmc_total <- sum(umap_data$SingleR_label == "VSMCs")
vsmc_miscalled <- sum(umap_data$SingleR_label == "VSMCs" & 
                      umap_data$ImmGen_label == "Macrophages")
cat(sprintf("VSMCs:\n"))
cat(sprintf("  Total: %d cells\n", vsmc_total))
cat(sprintf("  Called 'Macrophages' by ImmGen: %d (%.1f%%)\n\n",
            vsmc_miscalled, 100 * vsmc_miscalled / vsmc_total))

# Ly6C-hi Monocytes
mono_total <- sum(umap_data$SingleR_label == "Ly6C-hi Mono")
mono_miscalled <- sum(umap_data$SingleR_label == "Ly6C-hi Mono" & 
                      umap_data$ImmGen_label == "Macrophages")
cat(sprintf("Ly6C-hi Monocytes:\n"))
cat(sprintf("  Total: %d cells\n", mono_total))
cat(sprintf("  Called 'Macrophages' by ImmGen: %d (%.1f%%)\n",
            mono_miscalled, 100 * mono_miscalled / mono_total))






```








Plot 3 - Marker Expression 


B cells: CD79A, CD79B, CD19, MS4A1          
Ly6C-hi monocytes:Ly6c2, Ccr2, Csf1r, Itgam 
VSMCs: Acta2, Myh11

```{r}

library(ggplot2)
library(Seurat)
library(patchwork)
library(dplyr)
library(viridis)


#Extracting UMAP coordinates
umap_coords <- as.data.frame(Embeddings(MERTK_adult_filtered, reduction = "umap"))
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")


# B Cells: Cd79a
bcell_marker <- "Cd79a"
if (!bcell_marker %in% rownames(MERTK_adult_filtered)) {
  # Try alternative capitalizations
  bcell_marker <- grep("^cd79a$|^CD79A$|^Cd79a$", 
                       rownames(MERTK_adult_filtered), 
                       value = TRUE, ignore.case = TRUE)[1]
}

# VSMCs: Acta2 
vsmc_marker <- "Acta2"
if (!vsmc_marker %in% rownames(MERTK_adult_filtered)) {
  vsmc_marker <- grep("^acta2$|^ACTA2$|^Acta2$",
                      rownames(MERTK_adult_filtered),
                      value = TRUE, ignore.case = TRUE)[1]
}

# Ly6C-hi Monocytes: Ly6c2 
mono_marker <- "Ly6c2"
if (!mono_marker %in% rownames(MERTK_adult_filtered)) {
  mono_marker <- grep("^ly6c2$|^LY6C2$|^Ly6c2$",
                      rownames(MERTK_adult_filtered),
                      value = TRUE, ignore.case = TRUE)[1]
}

# Extracting normalized expression
bcell_expr <- FetchData(MERTK_adult_filtered, vars = bcell_marker, slot = "data")[,1]
vsmc_expr  <- FetchData(MERTK_adult_filtered, vars = vsmc_marker,  slot = "data")[,1]
mono_expr  <- FetchData(MERTK_adult_filtered, vars = mono_marker,  slot = "data")[,1]

# Combining with UMAP coordinates
umap_data_bcells <- cbind(umap_coords, Expression = bcell_expr)
umap_data_vsmcs  <- cbind(umap_coords, Expression = vsmc_expr)
umap_data_mono   <- cbind(umap_coords, Expression = mono_expr)

#Color scale: viridis for expression

expression_limits <- c(0, NA)  

#Base theme for each plot 
umap_theme <- theme_classic(base_size = 11, base_family = "Helvetica") +
  theme(
    plot.background  = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),
    
    axis.line   = element_line(colour = "black", linewidth = 0.6),
    axis.ticks  = element_line(colour = "black", linewidth = 0.4),
    axis.text   = element_text(colour = "black", size = 9),
    axis.title  = element_text(colour = "black", size = 10, face = "bold"),
    
    plot.title = element_text(colour = "black", size = 11, face = "bold", hjust = 0.5),
    
    # Legend for expression scale
    legend.position = "right",
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 8),
    legend.key.height = unit(0.8, "cm"),
    legend.key.width = unit(0.3, "cm"),
    
    plot.margin = margin(10, 5, 10, 10)
  )

#Panel A: B Cell Marker (Cd79a)
p_bcells <- ggplot(umap_data_bcells, aes(x = UMAP_1, y = UMAP_2, color = Expression)) +
  geom_point(size = 0.3, alpha = 0.9) +
  scale_color_viridis_c(option = "plasma", 
                        limits = expression_limits,
                        name = "Expression") +
  labs(title = paste0("B Cells: ", bcell_marker), 
       x = "UMAP 1", 
       y = "UMAP 2") +
  umap_theme

#Panel B: VSMC Marker (Acta2) 
p_vsmcs <- ggplot(umap_data_vsmcs, aes(x = UMAP_1, y = UMAP_2, color = Expression)) +
  geom_point(size = 0.3, alpha = 0.9) +
  scale_color_viridis_c(option = "plasma",
                        limits = expression_limits,
                        name = "Expression") +
  labs(title = paste0("VSMCs: ", vsmc_marker),
       x = "UMAP 1",
       y = "UMAP 2") +
  umap_theme

#Panel C: Monocyte Marker (Ly6c2)
p_mono <- ggplot(umap_data_mono, aes(x = UMAP_1, y = UMAP_2, color = Expression)) +
  geom_point(size = 0.3, alpha = 0.9) +
  scale_color_viridis_c(option = "plasma",
                        limits = expression_limits,
                        name = "Expression") +
  labs(title = paste0("Ly6C-hi Mono: ", mono_marker),
       x = "UMAP 1",
       y = "UMAP 2") +
  umap_theme

#Combine
p_combined <- p_bcells | p_vsmcs | p_mono

print(p_combined)

#Saving
ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/UMAP_Marker_Expression.png",
       plot = p_combined, width = 14, height = 4, dpi = 300, bg = "white")

ggsave("/projects/b1246/Isaac/MERTK/Figures/Atlas/Classification/UMAP_Marker_Expression.pdf",
       plot = p_combined, width = 14, height = 4, device = cairo_pdf)

#Marker Expression Statistics
cell_types <- MERTK_adult_filtered$SingleR_label
immgen_labels <- MERTK_adult_filtered$ImmGen_label

# B Cells
bcells_idx <- cell_types == "B Cells"
bcells_miscalled_idx <- bcells_idx & immgen_labels == "Macrophages"
cat(sprintf("%s expression in B Cells:\n", bcell_marker))
cat(sprintf("  Mean in true B cells: %.2f\n", mean(bcell_expr[bcells_idx])))
cat(sprintf("  Mean in B cells ImmGen calls 'Macrophages': %.2f\n", 
            mean(bcell_expr[bcells_miscalled_idx])))
cat(sprintf("  %% of B cells expressing %s (>0): %.1f%%\n\n",
            bcell_marker, 100 * sum(bcell_expr[bcells_idx] > 0) / sum(bcells_idx)))

# VSMCs
vsmcs_idx <- cell_types == "VSMCs"
vsmcs_miscalled_idx <- vsmcs_idx & immgen_labels == "Macrophages"
cat(sprintf("%s expression in VSMCs:\n", vsmc_marker))
cat(sprintf("  Mean in true VSMCs: %.2f\n", mean(vsmc_expr[vsmcs_idx])))
cat(sprintf("  Mean in VSMCs ImmGen calls 'Macrophages': %.2f\n",
            mean(vsmc_expr[vsmcs_miscalled_idx])))
cat(sprintf("  %% of VSMCs expressing %s (>0): %.1f%%\n\n",
            vsmc_marker, 100 * sum(vsmc_expr[vsmcs_idx] > 0) / sum(vsmcs_idx)))

# Ly6C-hi Monocytes
mono_idx <- cell_types == "Ly6C-hi Mono"
mono_miscalled_idx <- mono_idx & immgen_labels == "Macrophages"
cat(sprintf("%s expression in Ly6C-hi Monocytes:\n", mono_marker))
cat(sprintf("  Mean in true Ly6C-hi Mono: %.2f\n", mean(mono_expr[mono_idx])))
cat(sprintf("  Mean in Ly6C-hi Mono ImmGen calls 'Macrophages': %.2f\n",
            mean(mono_expr[mono_miscalled_idx])))
cat(sprintf("  %% of Ly6C-hi Mono expressing %s (>0): %.1f%%\n",
            mono_marker, 100 * sum(mono_expr[mono_idx] > 0) / sum(mono_idx)))

#Comparing true macrophages
cat("\n── Control: Expression in TRUE Macrophages ──\n\n")
true_macs_idx <- grepl("Mac", cell_types)

cat(sprintf("%s in true Macrophages: %.2f (expected low)\n", 
            bcell_marker, mean(bcell_expr[true_macs_idx])))
cat(sprintf("%s in true Macrophages: %.2f (expected low)\n",
            vsmc_marker, mean(vsmc_expr[true_macs_idx])))
cat(sprintf("%s in true Macrophages: %.2f (expected low)\n",
            mono_marker, mean(mono_expr[true_macs_idx])))


```















```{r ATLAS vs Immgen: Comparison figures}


################################################################################
# EXTENDED VALIDATION - ALL PRIORITY MISCLASSIFICATIONS
# Includes: Monocytes (both subtypes), B cells, VSMCs, Cardiomyocytes, 
#           Fibroblasts, AND NKT cells
################################################################################

library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)

## Define all marker panels
marker_panels <- list(
  Mac_Resident   = c("Timd4","Lyve1","Folr2","Mrc1","Igf1","F13a1","Cd163"),
  Mac_CCR2       = c("Ccr2","Trem2","Apoe","Cd14","H2-Ab1"),
  Mac_MHCIIhi    = c("Runx3","Cx3cr1","Il1b","Maf","Klf2","H2-Aa","Cd74"),
  Monocyte       = c("Ly6c2","Csf1r","Itgam","Ccr2"),
  Bcell          = c("Cd79a","Ms4a1","Cd19","Cd79b"),
  NK_NKT         = c("Ncr1","Klrb1c","Cd3e","Gzmb"),
  T_markers      = c("Cd3e","Cd4","Cd8a"),
  Endo_general   = c("Pecam1","Cdh5","Kdr","Eng"),
  Fibro_general  = c("Cthrc1","Acta2","Fn1","Col1a1","Postn","Tagln","Pdgfra"),
  Cardio         = c("Tnnt2","Myh6","Myl2","Ttn"),
  VSMC           = c("Acta2","Myh11","Tagln","Cnn1"),
  Neut           = c("Ly6g","S100a8","S100a9")
)

## Master output directory
master_dir <- "/projects/b1246/Isaac/MERTK/Figures/ALL_Misclassifications"
dir.create(master_dir, showWarnings = FALSE, recursive = TRUE)

## Initialize summary table
all_misclass_summary <- data.frame()

## ============================================================================
## 1. LY6C-HI MONOCYTES 
## ============================================================================

cat("\n=== VALIDATING: Ly6C-hi Monocytes ===\n")

atlas_ly6chi <- MERTK_adult_filtered$SingleR_label == "Ly6C-hi Mono"
immgen_mac <- MERTK_adult_filtered$ImmGen_label == "Macrophages"
misclass_ly6chi <- atlas_ly6chi & immgen_mac

cat("Ly6C-hi Mono misclassified:", sum(misclass_ly6chi), "\n")

MERTK_adult_filtered$ly6chi_status <- "Other"
MERTK_adult_filtered$ly6chi_status[atlas_ly6chi] <- "Atlas: Ly6C-hi Mono"
MERTK_adult_filtered$ly6chi_status[misclass_ly6chi] <- "ImmGen: 'Macrophages'"

# Markers
mono_pos <- intersect(marker_panels$Monocyte, rownames(MERTK_adult_filtered))
mac_neg <- intersect(c(marker_panels$Mac_Resident, marker_panels$Mac_CCR2), 
                     rownames(MERTK_adult_filtered))
all_markers_ly6chi <- unique(c(mono_pos, mac_neg))

# Subset
cells_ly6chi <- MERTK_adult_filtered$ly6chi_status != "Other"
ly6chi_subset <- subset(MERTK_adult_filtered, cells = colnames(MERTK_adult_filtered)[cells_ly6chi])

dir_ly6chi <- file.path(master_dir, "Ly6Chi_Monocytes")
dir.create(dir_ly6chi, showWarnings = FALSE, recursive = TRUE)

# Dot plot
if (length(all_markers_ly6chi) > 0) {
  p_dot <- DotPlot(ly6chi_subset, features = all_markers_ly6chi, 
                   group.by = "ly6chi_status", cols = c("lightgrey", "red"), dot.scale = 10) +
    coord_flip() + ggtitle("Ly6C-hi Monocytes: Marker Expression") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave("Ly6Chi_Mono_dotplot.pdf", plot = p_dot, path = dir_ly6chi, 
         width = 10, height = 8, dpi = 300)
}

# Violin plots
if (length(mono_pos) >= 2) {
  violins <- list()
  for (marker in head(mono_pos, 4)) {
    p <- VlnPlot(ly6chi_subset, features = marker, group.by = "ly6chi_status",
                 pt.size = 0, cols = c("#E41A1C", "#377EB8")) +
      ggtitle(marker) + theme(legend.position = "none", axis.title.x = element_blank())
    violins[[marker]] <- p
  }
  
  p_violin <- wrap_plots(violins, ncol = 2) +
    plot_annotation(title = "Ly6C-hi Monocyte Canonical Marker Expression")
  
  ggsave("Ly6Chi_Mono_violins.pdf", plot = p_violin, path = dir_ly6chi,
         width = 12, height = 8, dpi = 300)
}

# Export cell IDs
write.csv(
  data.frame(cell_id = colnames(MERTK_adult_filtered)[misclass_ly6chi],
             atlas_label = "Ly6C-hi Mono", immgen_label = "Macrophage"),
  file = file.path(dir_ly6chi, "misclassified_Ly6Chi_ids.csv"), row.names = FALSE
)

all_misclass_summary <- rbind(all_misclass_summary, data.frame(
  Cell_Type = "Ly6C-hi Monocytes", N_Misclassified = sum(misclass_ly6chi),
  Pct_of_ImmGen_Macs = round(sum(misclass_ly6chi)/sum(immgen_mac)*100, 1),
  Classification = "Lineage Error", Category = "Myeloid"
))



## ============================================================================
## 2. LY6C-LO MONOCYTES
## ============================================================================

cat("\n=== VALIDATING: Ly6C-lo Monocytes ===\n")

atlas_ly6clo <- MERTK_adult_filtered$SingleR_label == "Ly6C-lo Mono"
misclass_ly6clo <- atlas_ly6clo & immgen_mac

cat("Ly6C-lo Mono misclassified:", sum(misclass_ly6clo), "\n")

MERTK_adult_filtered$ly6clo_status <- "Other"
MERTK_adult_filtered$ly6clo_status[atlas_ly6clo] <- "Atlas: Ly6C-lo Mono"
MERTK_adult_filtered$ly6clo_status[misclass_ly6clo] <- "ImmGen: 'Macrophages'"

# Subset
cells_ly6clo <- MERTK_adult_filtered$ly6clo_status != "Other"
ly6clo_subset <- subset(MERTK_adult_filtered, cells = colnames(MERTK_adult_filtered)[cells_ly6clo])

dir_ly6clo <- file.path(master_dir, "Ly6Clo_Monocytes")
dir.create(dir_ly6clo, showWarnings = FALSE, recursive = TRUE)

# Dot plot
if (length(all_markers_ly6chi) > 0) {  # Use same markers as Ly6C-hi
  p_dot <- DotPlot(ly6clo_subset, features = all_markers_ly6chi, 
                   group.by = "ly6clo_status", cols = c("lightgrey", "red"), dot.scale = 10) +
    coord_flip() + ggtitle("Ly6C-lo Monocytes: Marker Expression") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave("Ly6Clo_Mono_dotplot.pdf", plot = p_dot, path = dir_ly6clo,
         width = 10, height = 8, dpi = 300)
}

# Export
write.csv(
  data.frame(cell_id = colnames(MERTK_adult_filtered)[misclass_ly6clo],
             atlas_label = "Ly6C-lo Mono", immgen_label = "Macrophage"),
  file = file.path(dir_ly6clo, "misclassified_Ly6Clo_ids.csv"), row.names = FALSE
)

all_misclass_summary <- rbind(all_misclass_summary, data.frame(
  Cell_Type = "Ly6C-lo Monocytes", N_Misclassified = sum(misclass_ly6clo),
  Pct_of_ImmGen_Macs = round(sum(misclass_ly6clo)/sum(immgen_mac)*100, 1),
  Classification = "Lineage Error", Category = "Myeloid"
))


## ============================================================================
## 3. B CELLS 
## ============================================================================

cat("\n=== VALIDATING: B Cells ===\n")

atlas_bcells <- MERTK_adult_filtered$SingleR_label == "B Cells"
misclass_bcells <- atlas_bcells & immgen_mac

cat("B Cells misclassified:", sum(misclass_bcells), "\n")

MERTK_adult_filtered$bcell_status <- "Other"
MERTK_adult_filtered$bcell_status[atlas_bcells] <- "Atlas: B Cells"
MERTK_adult_filtered$bcell_status[misclass_bcells] <- "ImmGen: 'Macrophages'"

# Markers
bcell_pos <- intersect(marker_panels$Bcell, rownames(MERTK_adult_filtered))
all_markers_bcell <- unique(c(bcell_pos, mac_neg))

# Subset
cells_bcell <- MERTK_adult_filtered$bcell_status != "Other"
bcell_subset <- subset(MERTK_adult_filtered, cells = colnames(MERTK_adult_filtered)[cells_bcell])

dir_bcell <- file.path(master_dir, "B_Cells")
dir.create(dir_bcell, showWarnings = FALSE, recursive = TRUE)

# Dot plot
if (length(all_markers_bcell) > 0) {
  p_dot <- DotPlot(bcell_subset, features = all_markers_bcell,
                   group.by = "bcell_status", cols = c("lightgrey", "blue"), dot.scale = 10) +
    coord_flip() + ggtitle("B Cells: Marker Expression") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave("BCells_dotplot.pdf", plot = p_dot, path = dir_bcell,
         width = 10, height = 8, dpi = 300)
}

# Violin plots
if (length(bcell_pos) >= 2) {
  violins <- list()
  for (marker in bcell_pos) {
    p <- VlnPlot(bcell_subset, features = marker, group.by = "bcell_status",
                 pt.size = 0, cols = c("#1F78B4", "#E41A1C")) +
      ggtitle(marker) + theme(legend.position = "none", axis.title.x = element_blank())
    violins[[marker]] <- p
  }
  
  p_violin <- wrap_plots(violins, ncol = 2) +
    plot_annotation(title = "B Cell Canonical Marker Expression")
  
  ggsave("BCells_violins.pdf", plot = p_violin, path = dir_bcell,
         width = 12, height = 8, dpi = 300)
}

# Export
write.csv(
  data.frame(cell_id = colnames(MERTK_adult_filtered)[misclass_bcells],
             atlas_label = "B Cells", immgen_label = "Macrophage"),
  file = file.path(dir_bcell, "misclassified_BCells_ids.csv"), row.names = FALSE
)

all_misclass_summary <- rbind(all_misclass_summary, data.frame(
  Cell_Type = "B Cells", N_Misclassified = sum(misclass_bcells),
  Pct_of_ImmGen_Macs = round(sum(misclass_bcells)/sum(immgen_mac)*100, 1),
  Classification = "Cross-Lineage", Category = "Lymphoid"
))


## ============================================================================
## 4. NKT CELLS 
## ============================================================================

cat("\n=== VALIDATING: NKT Cells ===\n")

atlas_nkt <- MERTK_adult_filtered$SingleR_label == "NKT"
misclass_nkt <- atlas_nkt & immgen_mac

cat("NKT cells misclassified:", sum(misclass_nkt), "\n")

MERTK_adult_filtered$nkt_status <- "Other"
MERTK_adult_filtered$nkt_status[atlas_nkt] <- "Atlas: NKT"
MERTK_adult_filtered$nkt_status[misclass_nkt] <- "ImmGen: 'Macrophages'"

# Markers
nkt_pos <- intersect(c(marker_panels$NK_NKT, marker_panels$T_markers), 
                     rownames(MERTK_adult_filtered))
all_markers_nkt <- unique(c(nkt_pos, mac_neg))

# Subset
cells_nkt <- MERTK_adult_filtered$nkt_status != "Other"
nkt_subset <- subset(MERTK_adult_filtered, cells = colnames(MERTK_adult_filtered)[cells_nkt])

dir_nkt <- file.path(master_dir, "NKT_Cells")
dir.create(dir_nkt, showWarnings = FALSE, recursive = TRUE)

# Dot plot
if (length(all_markers_nkt) > 0) {
  p_dot <- DotPlot(nkt_subset, features = all_markers_nkt,
                   group.by = "nkt_status", cols = c("lightgrey", "purple"), dot.scale = 10) +
    coord_flip() + ggtitle("NKT Cells: Marker Expression") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave("NKT_dotplot.pdf", plot = p_dot, path = dir_nkt,
         width = 10, height = 8, dpi = 300)
}

# Violin plots
if (length(nkt_pos) >= 2) {
  violins <- list()
  for (marker in head(nkt_pos, 4)) {
    p <- VlnPlot(nkt_subset, features = marker, group.by = "nkt_status",
                 pt.size = 0, cols = c("#9467BD", "#E41A1C")) +
      ggtitle(marker) + theme(legend.position = "none", axis.title.x = element_blank())
    violins[[marker]] <- p
  }
  
  p_violin <- wrap_plots(violins, ncol = 2) +
    plot_annotation(title = "NKT Cell Canonical Marker Expression")
  
  ggsave("NKT_violins.pdf", plot = p_violin, path = dir_nkt,
         width = 12, height = 8, dpi = 300)
}

# Export
write.csv(
  data.frame(cell_id = colnames(MERTK_adult_filtered)[misclass_nkt],
             atlas_label = "NKT", immgen_label = "Macrophage"),
  file = file.path(dir_nkt, "misclassified_NKT_ids.csv"), row.names = FALSE
)

all_misclass_summary <- rbind(all_misclass_summary, data.frame(
  Cell_Type = "NKT Cells", N_Misclassified = sum(misclass_nkt),
  Pct_of_ImmGen_Macs = round(sum(misclass_nkt)/sum(immgen_mac)*100, 1),
  Classification = "Cross-Lineage", Category = "Lymphoid"
))


## ============================================================================
## 5. VSMCs
## ============================================================================

cat("\n=== VALIDATING: VSMCs ===\n")

atlas_vsmc <- MERTK_adult_filtered$SingleR_label %in% c("VSMCs", "Transitional MMT")
misclass_vsmc <- atlas_vsmc & immgen_mac

cat("VSMCs misclassified:", sum(misclass_vsmc), "\n")

MERTK_adult_filtered$vsmc_status <- "Other"
MERTK_adult_filtered$vsmc_status[atlas_vsmc] <- "Atlas: VSMCs"
MERTK_adult_filtered$vsmc_status[misclass_vsmc] <- "ImmGen: 'Macrophages'"

# Markers
vsmc_pos <- intersect(marker_panels$VSMC, rownames(MERTK_adult_filtered))
all_markers_vsmc <- unique(c(vsmc_pos, mac_neg))

# Subset
cells_vsmc <- MERTK_adult_filtered$vsmc_status != "Other"
vsmc_subset <- subset(MERTK_adult_filtered, cells = colnames(MERTK_adult_filtered)[cells_vsmc])

dir_vsmc <- file.path(master_dir, "VSMCs")
dir.create(dir_vsmc, showWarnings = FALSE, recursive = TRUE)

# Dot plot
if (length(all_markers_vsmc) > 0) {
  p_dot <- DotPlot(vsmc_subset, features = all_markers_vsmc,
                   group.by = "vsmc_status", cols = c("lightgrey", "red"), dot.scale = 10) +
    coord_flip() + ggtitle("VSMCs: Marker Expression") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave("VSMCs_dotplot.pdf", plot = p_dot, path = dir_vsmc,
         width = 10, height = 8, dpi = 300)
}

# Export
write.csv(
  data.frame(cell_id = colnames(MERTK_adult_filtered)[misclass_vsmc],
             atlas_label = MERTK_adult_filtered$SingleR_label[misclass_vsmc],
             immgen_label = "Macrophage"),
  file = file.path(dir_vsmc, "misclassified_VSMCs_ids.csv"), row.names = FALSE
)

all_misclass_summary <- rbind(all_misclass_summary, data.frame(
  Cell_Type = "VSMCs", N_Misclassified = sum(misclass_vsmc),
  Pct_of_ImmGen_Macs = round(sum(misclass_vsmc)/sum(immgen_mac)*100, 1),
  Classification = "Non-Immune", Category = "Stromal"
))


## ============================================================================
## 6. CARDIOMYOCYTES 
## ============================================================================

cat("\n=== VALIDATING: Cardiomyocytes ===\n")

atlas_cardio <- MERTK_adult_filtered$SingleR_label == "Cardiomyocytes"
misclass_cardio <- atlas_cardio & immgen_mac

cat("Cardiomyocytes misclassified:", sum(misclass_cardio), "\n")

MERTK_adult_filtered$cardio_status <- "Other"
MERTK_adult_filtered$cardio_status[atlas_cardio] <- "Atlas: Cardiomyocytes"
MERTK_adult_filtered$cardio_status[misclass_cardio] <- "ImmGen: 'Macrophages'"

# Markers
cardio_pos <- intersect(marker_panels$Cardio, rownames(MERTK_adult_filtered))
all_markers_cardio <- unique(c(cardio_pos, mac_neg))

# Subset
cells_cardio <- MERTK_adult_filtered$cardio_status != "Other"
cardio_subset <- subset(MERTK_adult_filtered, cells = colnames(MERTK_adult_filtered)[cells_cardio])

dir_cardio <- file.path(master_dir, "Cardiomyocytes")
dir.create(dir_cardio, showWarnings = FALSE, recursive = TRUE)

# Dot plot
if (length(all_markers_cardio) > 0) {
  p_dot <- DotPlot(cardio_subset, features = all_markers_cardio,
                   group.by = "cardio_status", cols = c("lightgrey", "red"), dot.scale = 10) +
    coord_flip() + ggtitle("Cardiomyocytes: Marker Expression") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave("Cardiomyocytes_dotplot.pdf", plot = p_dot, path = dir_cardio,
         width = 10, height = 8, dpi = 300)
}

# Violin plots
if (length(cardio_pos) >= 2) {
  violins <- list()
  for (marker in cardio_pos) {
    p <- VlnPlot(cardio_subset, features = marker, group.by = "cardio_status",
                 pt.size = 0, cols = c("#E41A1C", "#377EB8")) +
      ggtitle(marker) + theme(legend.position = "none", axis.title.x = element_blank())
    violins[[marker]] <- p
  }
  
  p_violin <- wrap_plots(violins, ncol = 2) +
    plot_annotation(title = "Cardiomyocyte Canonical Marker Expression")
  
  ggsave("Cardiomyocytes_violins.pdf", plot = p_violin, path = dir_cardio,
         width = 12, height = 8, dpi = 300)
}

# Export
write.csv(
  data.frame(cell_id = colnames(MERTK_adult_filtered)[misclass_cardio],
             atlas_label = "Cardiomyocytes", immgen_label = "Macrophage"),
  file = file.path(dir_cardio, "misclassified_Cardiomyocytes_ids.csv"), row.names = FALSE
)

all_misclass_summary <- rbind(all_misclass_summary, data.frame(
  Cell_Type = "Cardiomyocytes", N_Misclassified = sum(misclass_cardio),
  Pct_of_ImmGen_Macs = round(sum(misclass_cardio)/sum(immgen_mac)*100, 1),
  Classification = "Non-Immune", Category = "Parenchymal"
))



## ============================================================================
## MASTER SUMMARY
## ============================================================================

# Sort by number of misclassified
all_misclass_summary <- all_misclass_summary %>%
  arrange(desc(N_Misclassified))

print("\n========================================")
print("MASTER SUMMARY OF ALL VALIDATIONS")
print("========================================")
print(all_misclass_summary)

# Save
write.csv(
  all_misclass_summary,
  file = file.path(master_dir, "ALL_misclassifications_summary.csv"),
  row.names = FALSE
)

## Summary plot
p_summary <- ggplot(all_misclass_summary, 
                    aes(x = reorder(Cell_Type, N_Misclassified), 
                        y = N_Misclassified, fill = Classification)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = paste0(N_Misclassified, " (", Pct_of_ImmGen_Macs, "%)")), 
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Lineage Error" = "#e74c3c",      # Red - same lineage but different stage
      "Cross-Lineage" = "#f39c12",      # Orange - different immune lineage
      "Non-Immune" = "#9b59b6"          # Purple - non-immune tissue
    ),
    name = "Error Type"
  ) +
  labs(
    title = "Reference-Based Annotation Errors in Cardiac Tissue",
    subtitle = "ImmGen reference misclassification of non-macrophage cell types",
    x = "Cell Type",
    y = "Number of Misclassified Cells"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30"),
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    panel.grid.major.y = element_blank()
  ) +
  ylim(0, max(all_misclass_summary$N_Misclassified) * 1.15)

ggsave(
  "ALL_misclassifications_summary_plot.pdf",
  plot = p_summary,
  path = master_dir,
  width = 12, height = 8, dpi = 300
)


cat("Total misclassified cells:", sum(all_misclass_summary$N_Misclassified), "\n")
cat("Percentage of ImmGen 'macrophage' calls:", 
    round(sum(all_misclass_summary$N_Misclassified)/sum(immgen_mac)*100, 1), "%\n")
cat("All outputs saved to:", master_dir, "\n")



```






Relationship between ImmGen precision and atlas cell-type prevalence.
```{r Precision-Prevalence Relationship}
################################################################################
# PRECISION-PREVALENCE ANALYSIS
# Precision vs. prevalence plots
################################################################################

library(dplyr)
library(ggplot2)

## --- Calculate metrics ---
total_cells <- ncol(MERTK_adult_filtered)
mac_subtypes <- c("CCR2+ Mac", "ARG1+ Mac", "MHCII Mac", "TLF Mac")

prevalence_data <- data.frame()

for (subtype in mac_subtypes) {
  atlas_is_type <- MERTK_adult_filtered$SingleR_label == subtype
  immgen_calls_mac <- MERTK_adult_filtered$ImmGen_label == "Macrophages"
  
  n_true <- sum(atlas_is_type)
  true_positives <- sum(atlas_is_type & immgen_calls_mac)
  
  prevalence <- (n_true / total_cells) * 100
  precision <- (true_positives / sum(immgen_calls_mac)) * 100
  
  prevalence_data <- rbind(prevalence_data, data.frame(
    Subtype = subtype,
    Cells = n_true,
    Prevalence = prevalence,
    Precision = precision
  ))
}

prevalence_data <- prevalence_data %>% arrange(desc(Prevalence))

## --- Output directory ---
output_dir <- "/projects/b1246/Isaac/MERTK/Figures/Precision_Prevalence_Analysis"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

## --- Statistical analysis ---
cor_test <- cor.test(prevalence_data$Prevalence, prevalence_data$Precision)
model <- lm(Precision ~ Prevalence, data = prevalence_data)

cat("\nCorrelation: r =", round(cor_test$estimate, 3), 
    ", p =", format.pval(cor_test$p.value, digits = 3), "\n")
cat("R² =", round(summary(model)$r.squared, 3), "\n\n")

print(prevalence_data)

write.csv(prevalence_data, 
          file.path(output_dir, "precision_prevalence_data.csv"), 
          row.names = FALSE)

## ============================================================================
## PLOT 1: SCATTER PLOT WITH REGRESSION
## ============================================================================

p1 <- ggplot(prevalence_data, aes(x = Prevalence, y = Precision)) +
  geom_smooth(method = "lm", se = TRUE, color = "#2c3e50", 
              fill = "#95a5a6", alpha = 0.3, linewidth = 1) +
  geom_point(aes(size = Cells), color = "#e74c3c", alpha = 0.8) +
  geom_text(aes(label = Subtype), vjust = -1.5, size = 4, fontface = "bold") +
  
  annotate("text", 
           x = max(prevalence_data$Prevalence) * 0.65, 
           y = min(prevalence_data$Precision) * 2,
           label = sprintf("r = %.3f\nR² = %.3f\np = %.3f", 
                          cor_test$estimate, 
                          summary(model)$r.squared, 
                          cor_test$p.value),
           size = 4.5, hjust = 0, fontface = "bold", color = "#2c3e50") +
  
  labs(
    title = "Reference Annotation Precision Depends on Cell Type Prevalence",
    x = "Prevalence in Cardiac Tissue (%)",
    y = "Annotation Precision (%)",
    size = "Cells"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  ) +
  
  scale_size_continuous(range = c(6, 15)) +
  scale_y_continuous(breaks = seq(0, 100, 10), limits = c(0, max(prevalence_data$Precision) * 1.15))

print(p1)

ggsave("precision_prevalence_scatter.pdf", plot = p1, path = output_dir,
       width = 9, height = 6, dpi = 300)

## ============================================================================
## PLOT 2: DUAL-AXIS COMPARISON
## ============================================================================

prevalence_data$Subtype <- factor(prevalence_data$Subtype, 
                                   levels = prevalence_data$Subtype)

scale_factor <- max(prevalence_data$Prevalence) / max(prevalence_data$Precision)

p2 <- ggplot(prevalence_data, aes(x = Subtype)) +
  
  # Prevalence bars
  geom_bar(aes(y = Prevalence), stat = "identity", 
           fill = "#3498db", alpha = 0.7, width = 0.6) +
  
  # Precision line and points
  geom_line(aes(y = Precision * scale_factor, group = 1), 
            color = "#e74c3c", linewidth = 1.8) +
  geom_point(aes(y = Precision * scale_factor), 
             color = "#e74c3c", size = 5) +
  
  scale_y_continuous(
    name = "Prevalence (%)",
    sec.axis = sec_axis(~./scale_factor, name = "Precision (%)")
  ) +
  
  labs(
    title = "Annotation Precision Degrades with Cell Type Rarity",
    x = "Macrophage Subtype"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11, face = "bold"),
    axis.title.y.left = element_text(color = "#3498db", face = "bold", size = 12),
    axis.title.y.right = element_text(color = "#e74c3c", face = "bold", size = 12),
    axis.text.y.left = element_text(color = "#3498db", size = 10),
    axis.text.y.right = element_text(color = "#e74c3c", size = 10),
    axis.title.x = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  )

print(p2)

ggsave("precision_prevalence_dual_axis.pdf", plot = p2, path = output_dir,
       width = 9, height = 6, dpi = 300)

## --- Summary ---
cat("\n=== ANALYSIS COMPLETE ===\n")
cat("Outputs saved to:", output_dir, "\n")
cat("  - precision_prevalence_data.csv\n")
cat("  - precision_prevalence_scatter.pdf\n")
cat("  - precision_prevalence_dual_axis.pdf\n\n")

cat("Key finding: Precision ranges from", 
    round(min(prevalence_data$Precision), 1), "% to",
    round(max(prevalence_data$Precision), 1), "%\n")
cat("as prevalence ranges from", 
    round(min(prevalence_data$Prevalence), 2), "% to",
    round(max(prevalence_data$Prevalence), 2), "%\n")








```



































#########################################
PSEUDOTIME ANALYSIS 
PSEUDOTIME ANALYSIS 
PSEUDOTIME ANALYSIS 
#######################################



```{r Separating Adult KO vs WT conditions}

MERTK_adult <- mertk_adult
table(MERTK_adult$group)

# Subset into KO and WT
MERTK_KO <- subset(MERTK_adult, subset = group == "KO_Adult")
MERTK_WT <- subset(MERTK_adult, subset = group == "WT_Adult")

# Sanity check
cat("KO cells:", ncol(MERTK_KO), "\n")
cat("WT cells:", ncol(MERTK_WT), "\n")

table(MERTK_KO$group)
table(MERTK_WT$group)
```


```{r}
saveRDS(MERTK_KO,
        file = "/scratch/inm3056/MERTK_KO.rds")
```

```{r}
saveRDS(MERTK_WT,
        file = "/scratch/inm3056/MERTK_WT.rds")
```



################################
PSEUDOTIME _ MAC/MONO SUBSETTING 
################################




```{r Subsetting Macrophages and Monocytes}

library(SeuratWrappers)

cds_WT <- as.cell_data_set(MERTK_WT)
cds_KO <- as.cell_data_set(MERTK_KO)

myeloid_labels <- c(
  "ARG1+ Mac", "CCR2+ Mac", "MHCII Mac", "TLF Mac",
  "Ly6C-hi Mono", "Ly6C-lo Mono"
)

cds_WT <- cds_WT[, cds_WT$SingleR_label %in% myeloid_labels]
cds_KO <- cds_KO[, cds_KO$SingleR_label %in% myeloid_labels]

table(cds_WT$SingleR_label)
table(cds_KO$SingleR_label)



```
```{r}

cds_WT <- preprocess_cds(cds_WT, num_dim = 30)
cds_WT <- reduce_dimension(cds_WT, reduction_method = "UMAP")
cds_WT <- cluster_cells(cds_WT)

cds_KO <- preprocess_cds(cds_KO, num_dim = 30)
cds_KO <- reduce_dimension(cds_KO, reduction_method = "UMAP")
cds_KO <- cluster_cells(cds_KO)
cds_WT <- learn_graph(cds_WT)
cds_KO <- learn_graph(cds_KO)

WT_root <- colnames(cds_WT)[cds_WT$SingleR_label == "Ly6C-hi Mono"]
KO_root <- colnames(cds_KO)[cds_KO$SingleR_label == "Ly6C-hi Mono"]

cds_WT <- order_cells(cds_WT, root_cells = WT_root)
cds_KO <- order_cells(cds_KO, root_cells = KO_root)


```

```{r Plotting by pseudotime}

library(ggplot2)
library(ggrepel)
library(dplyr)

### ---- WT BASE PLOT ----
wt_plot_base <- plot_cells(
  cds_WT,
  color_cells_by = "pseudotime",
  label_cell_groups = FALSE,
  label_leaves = TRUE,
  label_branch_points = TRUE,
  graph_label_size = 4,
  cell_size = 1.0,
  alpha = 0.85,
  show_trajectory_graph = TRUE
)

### ---- Extract coordinates ----
wt_coords <- as.data.frame(reducedDims(cds_WT)$UMAP)
wt_coords$label <- cds_WT$SingleR_label
colnames(wt_coords)[1:2] <- c("UMAP_1", "UMAP_2")

### ---- Compute centroids ----
wt_centroids <- wt_coords %>%
  group_by(label) %>%
  summarise(
    x = median(UMAP_1),
    y = median(UMAP_2)
  )

wt_pseudotime_umap <- wt_plot_base +
  geom_text_repel(
    data = wt_centroids,
    aes(x = x, y = y, label = label),
    size = 6,
    fontface = "bold",
    color = "black",
    point.padding = 0.6,
    box.padding = 0.7,
    segment.color = "gray40",
    max.overlaps = Inf
  ) +
  scale_color_viridis_c(option = "plasma", name = "Pseudotime") +
  theme_bw(base_size = 18) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

wt_pseudotime_umap


### ---- KO BASE PLOT ----
ko_plot_base <- plot_cells(
  cds_KO,
  color_cells_by = "pseudotime",
  label_cell_groups = FALSE,
  label_leaves = TRUE,
  label_branch_points = TRUE,
  graph_label_size = 4,
  cell_size = 1.0,
  alpha = 0.85,
  show_trajectory_graph = TRUE
)

### ---- Extract coordinates ----
ko_coords <- as.data.frame(reducedDims(cds_KO)$UMAP)
ko_coords$label <- cds_KO$SingleR_label
colnames(ko_coords)[1:2] <- c("UMAP_1", "UMAP_2")

### ---- Compute centroids ----
ko_centroids <- ko_coords %>%
  group_by(label) %>%
  summarise(
    x = median(UMAP_1),
    y = median(UMAP_2)
  )

### ---- Final KO pseudotime UMAP ----
ko_pseudotime_umap <- ko_plot_base +
  geom_text_repel(
    data = ko_centroids,
    aes(x = x, y = y, label = label),
    size = 6,
    fontface = "bold",
    color = "black",
    point.padding = 0.6,
    box.padding = 0.7,
    segment.color = "gray40",
    max.overlaps = Inf
  ) +
  scale_color_viridis_c(option = "plasma", name = "Pseudotime") +
  theme_bw(base_size = 18) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ko_pseudotime_umap

```



```{r Pseudotime Gene Differential Expression}

genes_all <- c(
  # Monocyte → macrophage
  "Ly6c2","Ccr2","Lgals3","Cd14",
  
  # M1 / inflammatory
  "Il1b","Tnf","Nos2","Nfkbia","S100a8","S100a9",
  
  # M2 / reparative
  "Arg1","Retnla","Chil3","Mrc1",
  
  # Efferocytosis / MerTK pathway
  "Mertk","Axl","Tyro3","Gas6","Mfge8","Cd36","Lrp1",
  
  # Antigen presentation
  "H2-Ab1","H2-Aa","Cd74","Ciita",
  
  # Tissue resident macrophages
  "Timd4","Cd163","Vsig4","Marco",
  
  # Repair/fibrosis associated
  "Spp1","Fn1"
)

rowData(cds_WT)$gene_short_name <- rownames(cds_WT)
rowData(cds_KO)$gene_short_name <- rownames(cds_KO)

valid_genes <- genes_all[genes_all %in% rownames(cds_WT)]
valid_genes


library(ggplot2)
library(dplyr)
library(patchwork)

plot_three_genes <- function(cds, genelist) {
  
  plist <- list()
  
  for (g in genelist) {
    df <- data.frame(
      pseudotime = pseudotime(cds),
      expr = exprs(cds)[g,],
      label = cds$SingleR_label
    )
    
    df$expr_log <- log10(df$expr + 1e-6)
    
    p <- ggplot(df, aes(pseudotime, expr_log, color = label)) +
      geom_point(size = 0.5, alpha = 0.35) +
      geom_smooth(se = FALSE, size = 1.2) +
      labs(title = g, y = "log10(expr + 1e-6)", x = "Pseudotime") +
      theme_bw(base_size = 14) +
      theme(
        plot.title = element_text(face = "bold", size = 18),
        legend.position = "right",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10)
      )
    
    plist[[g]] <- p
  }
  
  wrap_plots(plist, ncol = 1)
}

gene_groups <- split(valid_genes, ceiling(seq_along(valid_genes)/3))
gene_groups

WT_plots <- lapply(gene_groups, function(g) plot_three_genes(cds_WT, g))
WT_plots[[1]]   # show the first 3-gene plot

KO_plots <- lapply(gene_groups, function(g) plot_three_genes(cds_KO, g))
KO_plots[[1]]   # first KO plot





```

















##############################################

              CELLCHAT 
              
##############################################

```{r Cell Chat Data Preprocessing}
## Load the required libraries
library(CellChat)
library(patchwork)
options(stringsAsFactors = FALSE)

# Part I: Data input & processing and initialization of CellChat object

MERTK_WT_cellchat <- createCellChat(object = MERTK_WT, meta = meta, group.by = "SingleR_label")
MERTK_KO_cellchat <- createCellChat(object = MERTK_KO, meta = meta, group.by = "SingleR_label")

```





#############################################################
#FIGURES 
#FIGURES 
#FIGURES 
############################################################

```{r Figure 1: Cardiac Atlas Creation}


library(ggplot2)
library(grid)
library(gridExtra)

# Atlas integration workflow diagram
create_accurate_workflow <- function() {
  
  # Stage 1: Per-dataset processing
  # Top row: 4 datasets (equally spaced)
  datasets <- data.frame(
    x = c(1.5, 3, 4.5, 6),
    y = rep(8.5, 4),
    width = rep(1.2, 4),
    label = c("E-MTAB-13264\n(HFrEF)", 
              "E-MTAB-9816\n(MI: 1D, 3D, 7D)", 
              "GSE119355\n(MI)", 
              "E-MTAB-7895\n(MI: 1D, 3D, 7D)"),
    type = "dataset",
    stringsAsFactors = FALSE
  )
  
  # Middle row: Processing pipeline (equally spaced)
  stage1_process <- data.frame(
    x = c(2.25, 4, 5.75),
    y = rep(7.5, 3),
    width = rep(1.4, 3),
    label = c("QC filtering &\nNormalization", 
              "Initial SCTransform\n(per dataset)",
              "Save 8 individual\nobjects"),
    type = c("qc", "qc", "save"),
    stringsAsFactors = FALSE
  )
  
  # Stage 2: Integration preparation (6 boxes, equally spaced)
  stage2_data <- data.frame(
    x = seq(1, 7, length.out = 6),
    y = rep(5.8, 6),
    width = rep(1, 6),
    label = c("Load 8\nobjects", 
              "Add cell IDs\n(dataset prefix)",
              "Map genes:\nSymbol → Ensembl\n(biomaRt)",
              "Enforce common\ngene universe\n(all datasets)",
              "Re-run SCT\n(full genes)",
              "PrepSCT\nIntegration"),
    type = "prep",
    stringsAsFactors = FALSE
  )
  
  # Stage 3: Integration (5 boxes, equally spaced)
  stage3_data <- data.frame(
    x = seq(1.2, 6.8, length.out = 5),
    y = rep(4.3, 5),
    width = rep(1.05, 5),
    label = c("Select 3000\nintegration\nfeatures",
              "Find integration\nanchors\n(30 PCs)",
              "IntegrateData\n(SCT method)",
              "RunPCA\n(30 PCs)",
              "RunUMAP +\nClustering\n(res 0.1-0.9)"),
    type = "integration",
    stringsAsFactors = FALSE
  )
  
  # Combine all elements
  all_boxes <- rbind(
    cbind(datasets, stage = 1),
    cbind(stage1_process, stage = 1),
    cbind(stage2_data, stage = 2),
    cbind(stage3_data, stage = 3)
  )
  
  # Arrows for Stage 1
  arrows_datasets <- data.frame(
    x1 = c(1.5, 3, 4.5, 6),
    y1 = rep(8.3, 4),
    x2 = c(1.5, 3, 4.5, 6),
    y2 = rep(7.8, 4)
  )
  
  arrows1_process <- data.frame(
    x1 = c(2.95, 4.7),
    y1 = rep(7.5, 2),
    x2 = c(3.3, 5.05),
    y2 = rep(7.5, 2)
  )
  
  # Between stage 1 and 2
  arrows_s1_s2 <- data.frame(
    x1 = 4, y1 = 7.15, x2 = 4, y2 = 6.15
  )
  
  # Stage 2 arrows
  arrows2 <- data.frame(
    x1 = stage2_data$x[1:5] + 0.5,
    y1 = rep(5.8, 5),
    x2 = stage2_data$x[2:6] - 0.5,
    y2 = rep(5.8, 5)
  )
  
  # Between stage 2 and 3
  arrows_s2_s3 <- data.frame(
    x1 = 4, y1 = 5.45, x2 = 4, y2 = 4.65
  )
  
  # Stage 3 arrows
  arrows3 <- data.frame(
    x1 = stage3_data$x[1:4] + 0.525,
    y1 = rep(4.3, 4),
    x2 = stage3_data$x[2:5] - 0.525,
    y2 = rep(4.3, 4)
  )
  
  # Color scheme
  colors <- c(
    "dataset" = "#2563eb",
    "qc" = "#059669",
    "save" = "#7c3aed",
    "prep" = "#d97706",
    "integration" = "#dc2626"
  )
  
  # Create plot
  p <- ggplot() +
    # Stage backgrounds
    annotate("rect", xmin = 0.5, xmax = 7.5, ymin = 7.05, ymax = 8.85,
             fill = "#e0f2fe", alpha = 0.4, color = "#0284c7", size = 0.8) +
    annotate("rect", xmin = 0.5, xmax = 7.5, ymin = 5.45, ymax = 6.15,
             fill = "#fef3c7", alpha = 0.4, color = "#d97706", size = 0.8) +
    annotate("rect", xmin = 0.5, xmax = 7.5, ymin = 3.95, ymax = 4.65,
             fill = "#fee2e2", alpha = 0.4, color = "#dc2626", size = 0.8) +
    
    # Stage labels
    annotate("text", x = 0.55, y = 8.75, 
             label = "Stage 1: Per-dataset QC, normalization & SCTransform",
             hjust = 0, fontface = "bold", size = 3.2, color = "gray20") +
    annotate("text", x = 0.55, y = 6.1, 
             label = "Stage 2: Standardization for integration (gene mapping, common universe, re-SCT)",
             hjust = 0, fontface = "bold", size = 3.2, color = "gray20") +
    annotate("text", x = 0.55, y = 4.6, 
             label = "Stage 3: SCT-based integration (Seurat v5)",
             hjust = 0, fontface = "bold", size = 3.2, color = "gray20") +
    
    # Annotation for Stage 1
    annotate("text", x = 3.75, y = 8, 
             label = "↓ Each dataset processed individually ↓",
             hjust = 0.5, fontface = "italic", size = 2.6, color = "gray40") +
    
    # Draw boxes
    geom_tile(data = all_boxes,
              aes(x = x, y = y, width = width, height = 0.35, fill = type),
              color = "white", size = 1.2, alpha = 0.95) +
    
    # Add labels
    geom_text(data = all_boxes,
              aes(x = x, y = y, label = label),
              size = 2.3, color = "white", fontface = "bold", 
              lineheight = 0.8, hjust = 0.5, vjust = 0.5) +
    
    # Dataset arrows
    geom_segment(data = arrows_datasets,
                 aes(x = x1, y = y1, xend = x2, yend = y2),
                 arrow = arrow(length = unit(0.08, "cm"), type = "closed"),
                 color = "gray50", size = 0.4) +
    
    # Stage 1 process arrows
    geom_segment(data = arrows1_process,
                 aes(x = x1, y = y1, xend = x2, yend = y2),
                 arrow = arrow(length = unit(0.1, "cm"), type = "closed"),
                 color = "gray40", size = 0.5) +
    
    # Between stages arrows
    geom_segment(data = arrows_s1_s2,
                 aes(x = x1, y = y1, xend = x2, yend = y2),
                 arrow = arrow(length = unit(0.12, "cm"), type = "closed"),
                 color = "gray20", size = 0.8) +
    geom_segment(data = arrows_s2_s3,
                 aes(x = x1, y = y1, xend = x2, yend = y2),
                 arrow = arrow(length = unit(0.12, "cm"), type = "closed"),
                 color = "gray20", size = 0.8) +
    
    # Stage 2 arrows
    geom_segment(data = arrows2,
                 aes(x = x1, y = y1, xend = x2, yend = y2),
                 arrow = arrow(length = unit(0.1, "cm"), type = "closed"),
                 color = "gray40", size = 0.5) +
    
    # Stage 3 arrows
    geom_segment(data = arrows3,
                 aes(x = x1, y = y1, xend = x2, yend = y2),
                 arrow = arrow(length = unit(0.1, "cm"), type = "closed"),
                 color = "gray40", size = 0.5) +
    
    scale_fill_manual(values = colors) +
    coord_cartesian(xlim = c(0.3, 7.7), ylim = c(3.8, 9)) +
    theme_void() +
    theme(
      legend.position = "none",
      plot.background = element_rect(fill = "white", color = NA),
      plot.margin = margin(10, 10, 10, 10)
    )
  
  return(p)
}

# Add methods annotation
create_figure_with_methods <- function() {
  p_main <- create_accurate_workflow()
  
  # Key methods details
  methods_text <- textGrob(
    paste(
      "Stage 1 QC: nUMI ≥500, nGene ≥250, log10(nGene/nUMI) >0.80, mitoRatio <30%, genes expressed in ≥10 cells",
      "  Normalization: NormalizeData → FindVariableFeatures (2000) → ScaleData → CellCycleScoring → PCA → SCTransform (glmGamPoi) → RunUMAP (30 dims)",
      "Stage 2: Cell IDs harmonized with dataset prefix → MGI symbol to Ensembl via biomaRt → Common gene universe enforced across all 8 datasets → Re-run SCTransform (return.only.var.genes=FALSE)",
      "Stage 3: SelectIntegrationFeatures (3000) → VariableFeatures assigned → PrepSCTIntegration → FindIntegrationAnchors (30 PCs, SCT) → IntegrateData → PCA/UMAP/FindNeighbors → FindClusters (res 0.1-0.9)",
      sep = "\n"
    ),
    x = 0.02, hjust = 0, just = "left",
    gp = gpar(fontsize = 7, fontfamily = "sans", lineheight = 1.3, col = "gray30")
  )
  
  # Combine
  layout <- rbind(
    c(1, 1, 1),
    c(1, 1, 1),
    c(1, 1, 1),
    c(2, 2, 2)
  )
  
  g <- grid.arrange(
    p_main,
    methods_text,
    layout_matrix = layout,
    heights = c(2.5, 2.5, 2.5, 1.5),
    top = textGrob("Cardiac Reference Atlas Integration Workflow", 
                   gp = gpar(fontsize = 13, fontface = "bold"))
  )
  
  return(g)
}

# Generate and save
fig <- create_figure_with_methods()

ggsave("cardiac_atlas_workflow_accurate.pdf", 
       plot = fig, 
       width = 9, 
       height = 6.5, 
       dpi = 600,
       device = cairo_pdf)

ggsave("cardiac_atlas_workflow_accurate.png", 
       plot = fig, 
       width = 9, 
       height = 6.5, 
       dpi = 600)

ggsave("cardiac_atlas_workflow_accurate.tiff", 
       plot = fig, 
       width = 9, 
       height = 6.5, 
       dpi = 600,
       compression = "lzw")

# Version without methods text
p_clean <- create_accurate_workflow() +
  ggtitle("Cardiac Reference Atlas Integration Workflow") +
  theme(plot.title = element_text(hjust = 0.5, size = 13, face = "bold",
                                   margin = margin(b = 10)))

ggsave("cardiac_atlas_workflow_clean.pdf", 
       plot = p_clean, 
       width = 9, 
       height = 4.5, 
       dpi = 600,
       device = cairo_pdf)

ggsave("cardiac_atlas_workflow_clean.png", 
       plot = p_clean, 
       width = 9, 
       height = 4.5, 
       dpi = 600)


```










