---
title: "Cardiac Single-Cell Atlas Construction and Annotation"
author: "Meite I, Pesce L, Feinstein M, Thorp EB, Lantz C"
date: "2026-03-26"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Packages}

library(Seurat)
library(harmony)
library(SingleCellExperiment)
library(batchelor)
library(dplyr)
library(stringr)
library(ggplot2)
library(patchwork)
library(gridExtra)
library(sctransform)
library(SingleR)

```


```{r Helper Functions definition}

# Base UMAP theme
umap_pub_theme <- function() {
  theme_classic(base_size = 11) +
    theme(
      plot.title       = element_blank(),
      axis.line        = element_blank(),
      axis.ticks       = element_blank(),
      axis.text        = element_blank(),
      axis.title       = element_blank(),
      panel.border     = element_blank(),
      panel.background = element_blank(),
      plot.background  = element_blank(),
      legend.text      = element_text(size = 8, face = "bold"),
      legend.title     = element_blank(),
      legend.key.size  = unit(0.35, "cm"),
      plot.margin      = margin(5, 5, 5, 5)
    )
}

# UMAP axis arrows
add_umap_arrows <- function(umap_df) {
  x_range   <- diff(range(umap_df$UMAP_1))
  y_range   <- diff(range(umap_df$UMAP_2))
  arrow_x   <- 3.0
  arrow_y   <- arrow_x * (y_range / x_range)
  list(
    annotate("segment",
             x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+0.2+arrow_x,
             y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+0.2,
             arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
             color = "black", linewidth = 0.8),
    annotate("segment",
             x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+0.2,
             y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+0.2+arrow_y,
             arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
             color = "black", linewidth = 0.8),
    annotate("text",
             x = min(umap_df$UMAP_1)+0.2+arrow_x/2,
             y = min(umap_df$UMAP_2)-0.5,
             label = "UMAP-1", size = 3, color = "black",
             hjust = 0.5, fontface = "bold"),
    annotate("text",
             x = min(umap_df$UMAP_1)-0.5,
             y = min(umap_df$UMAP_2)+0.2+arrow_y/2,
             label = "UMAP-2", size = 3, color = "black",
             hjust = 0.5, angle = 90, fontface = "bold")
  )
}

# Bar chart theme
bar_pub_theme <- function() {
  theme_classic(base_size = 11) +
    theme(
      plot.title       = element_blank(),
      axis.text.x      = element_text(angle = 45, hjust = 1, vjust = 1,
                                       size = 9, color = "black", face = "bold"),
      axis.text.y      = element_text(size = 9, color = "black", face = "bold"),
      axis.title       = element_text(size = 10, color = "black", face = "bold"),
      axis.line        = element_line(color = "black", linewidth = 0.6),
      axis.ticks       = element_line(color = "black", linewidth = 0.4),
      panel.border     = element_blank(),
      panel.grid       = element_blank(),
      legend.text      = element_text(size = 9, face = "bold"),
      legend.title     = element_blank(),
      legend.key.size  = unit(0.4, "cm"),
      plot.margin      = margin(10, 10, 5, 10)
    )
}

```


#################################################
ATLAS CREATION 
################################################

#NORMALIZATION OF DATASET W/ ACCESSION E-MTAB-13264


```{r Normaizing Dataset ACCESSION: E-MTAB-13264}


seurat_13264 <- readRDS("/projects/b1246/Isaac/MERTK/Atlas_Datasets/E-MTAB-13264.rds")

#Checking metadata
table(seurat_13264$condition)
#Selecting HFrEF condition
seurat_13264_HF <- subset(seurat_13264, subset = condition == "Mice-HFrEF")

#saving original clusters:
seurat_13264_HF$orig_cluster <- Idents(seurat_13264_HF)


```



#Quality Control

```{r Quality Control, include = False}

# Quality Control Setup
DefaultAssay(seurat_13264_HF) <- "RNA"

# Human patterns due multi-species atlas: use MT-, RPS, RPL 
seurat_13264_HF[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_13264_HF,
  pattern = "^MT-"
)

seurat_13264_HF[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_13264_HF,
  pattern = "^RP[SL]"
)

# Convert to ratios
seurat_13264_HF$mitoRatio <- seurat_13264_HF$percent.mito / 100
seurat_13264_HF$riboRatio <- seurat_13264_HF$percent.ribo / 100


```




#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_13264_HF <- seurat_13264_HF@meta.data

# Assign metadata based on orig.ident patterns
metadata_13264_HF <- metadata_13264_HF %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )

# Assign back to Seurat object
seurat_13264_HF@meta.data <- metadata_13264_HF
seurat_13264_HF$log10GenesPerUMI <- log10(seurat_13264_HF$nGene) / log10(seurat_13264_HF$nUMI)

```



```{r Filtered_seurat, include = False}

seurat_13264_HF <- subset(x = seurat_13264_HF, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Get raw counts slot
counts <- GetAssayData(seurat_13264_HF, assay = "RNA", slot = "counts")

# Filter genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_13264_HF <- subset(seurat_13264_HF, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```





#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################



```{r SCTransform, include = FALSE}

library(SeuratData)
library(glmGamPoi)

seurat_13264_HF <- CellCycleScoring(seurat_13264_HF, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

seurat_13264_HF <- Seurat::SCTransform(seurat_13264_HF, method = "glmGamPoi")

seurat_13264_HF <- RunPCA(seurat_13264_HF)
seurat_13264_HF <- RunUMAP( seurat_13264_HF, dims = 1:30)


```




#NORMALIZATION OF DATASET W/ ACCESSION EMTAB9816

```{r}

#---------------------
#.     MI@1D
#--------------------

counts_9816_1d <- Read10X(data.dir= "/projects/b1246/Isaac/MERTK/Atlas_Datasets/EMTAB9816/MI1D")

seurat_9816_1d <- CreateSeuratObject(counts = counts_9816_1d, project = "EMTAB9816_MI1D")


```


#Quality Control

```{r Quality Control, include = False}

seurat_9816_1d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_9816_1d,
  pattern = "^mt-"
)

seurat_9816_1d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_9816_1d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_9816_1d$mitoRatio <- seurat_9816_1d$percent.mito / 100
seurat_9816_1d$riboRatio <- seurat_9816_1d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_9816_1d <- seurat_9816_1d@meta.data

# Assign metadata based on orig.ident patterns
metadata_9816_1d <- metadata_9816_1d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_9816_1d@meta.data <- metadata_9816_1d
seurat_9816_1d$log10GenesPerUMI <- log10(seurat_9816_1d$nFeature_RNA) / log10(seurat_9816_1d$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_9816_1d <- subset(x = seurat_9816_1d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_9816_1d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_9816_1d <- subset(seurat_9816_1d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################



```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_9816_1d <- CellCycleScoring(seurat_9816_1d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)


seurat_9816_1d <- Seurat::SCTransform(seurat_9816_1d, method = "glmGamPoi")

seurat_9816_1d <- RunPCA(seurat_9816_1d)
seurat_9816_1d <- RunUMAP( seurat_9816_1d, dims = 1:30)

```




#-----------------
#EMTAB9816 MI3D
#-----------------



```{r}

#---------------------
#.     MI@3D
#--------------------
count_9816_3d <- Read10X(data.dir= "/projects/b1246/Isaac/MERTK/Atlas_Datasets/EMTAB9816/MI3D")

seurat_9816_3d <- CreateSeuratObject(counts = count_9816_3d, project = "EMTAB9816_MI3D")

```


#Quality Control

```{r Quality Control, include = False}

seurat_9816_3d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_9816_3d,
  pattern = "^mt-"
)

seurat_9816_3d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_9816_3d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_9816_3d$mitoRatio <- seurat_9816_3d$percent.mito / 100
seurat_9816_3d$riboRatio <- seurat_9816_3d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_9816_3d <- seurat_9816_3d@meta.data

# Assign metadata based on orig.ident patterns
metadata_9816_3d <- metadata_9816_3d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_9816_3d@meta.data <- metadata_9816_3d
seurat_9816_3d$log10GenesPerUMI <- log10(seurat_9816_3d$nFeature_RNA) / log10(seurat_9816_3d$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_9816_3d <- subset(x = seurat_9816_3d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_9816_3d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_9816_3d <- subset(seurat_9816_3d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```




#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_9816_3d <- CellCycleScoring(seurat_9816_3d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

seurat_9816_3d <- Seurat::SCTransform(seurat_9816_3d, method = "glmGamPoi")

seurat_9816_3d <- RunPCA(seurat_9816_3d)
seurat_9816_3d <- RunUMAP( seurat_9816_3d, dims = 1:30)


```






#-----------------
#EMTAB9816 MI7D
#-----------------


```{r}

#---------------------
#.     MI@7D
#--------------------
count_9816_7d <- Read10X(data.dir= "/projects/b1246/Isaac/MERTK/Atlas_Datasets/EMTAB9816/MI7D")

seurat_9816_7d <- CreateSeuratObject(counts = count_9816_7d, project = "EMTAB9816_MI7D")

```


#Quality Control

```{r Quality Control, include = False}

seurat_9816_7d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_9816_7d,
  pattern = "^mt-"
)

seurat_9816_7d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_9816_7d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_9816_7d$mitoRatio <- seurat_9816_7d$percent.mito / 100
seurat_9816_7d$riboRatio <- seurat_9816_7d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_9816_7d <- seurat_9816_7d@meta.data

# Assign metadata based on orig.ident patterns
metadata_9816_7d <- metadata_9816_7d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_9816_7d@meta.data <- metadata_9816_7d
seurat_9816_7d$log10GenesPerUMI <- log10(seurat_9816_7d$nFeature_RNA) / log10(seurat_9816_7d$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_9816_7d <- subset(x = seurat_9816_7d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_9816_7d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_9816_7d <- subset(seurat_9816_7d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```




#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################

```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_9816_7d <- CellCycleScoring(seurat_9816_7d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)


seurat_9816_7d <- Seurat::SCTransform(seurat_9816_7d, method = "glmGamPoi")

seurat_9816_7d <- RunPCA(seurat_9816_7d)
seurat_9816_7d <- RunUMAP( seurat_9816_7d, dims = 1:30)


```



#NORMALIZATION OF DATASET W/ ACCESSION GSE119355


```{r}

counts_Epelman <- Read10X(data.dir = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/GSE119355")

seurat_Epelman <- CreateSeuratObject(counts = counts_Epelman, project = "GSE119355")



```


#Quality Control

```{r Quality Control, include = False}

seurat_Epelman[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_Epelman,
  pattern = "^mt-"
)

seurat_Epelman[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_Epelman,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_Epelman$mitoRatio <- seurat_Epelman$percent.mito / 100
seurat_Epelman$riboRatio <- seurat_Epelman$percent.ribo / 100
```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_Epelman <- seurat_Epelman@meta.data

# Assign metadata based on orig.ident patterns
metadata_Epelman <- metadata_Epelman %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_Epelman@meta.data <- metadata_Epelman
seurat_Epelman$log10GenesPerUMI <- log10(seurat_Epelman$nFeature_RNA) / log10(seurat_Epelman$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_Epelman <- subset(x =seurat_Epelman, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_Epelman, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_Epelman <- subset(seurat_Epelman, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_Epelman <- CellCycleScoring(seurat_Epelman, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

seurat_Epelman <- Seurat::SCTransform(seurat_Epelman, method = "glmGamPoi")

seurat_Epelman <- RunPCA(seurat_Epelman)
seurat_Epelman <- RunUMAP(seurat_Epelman, dims = 1:30)


```









#NORMALIZATION OF DATASET W/ ACCESSION EMTAB7895


```{r}

#---------------------
#.     MI@1D
#--------------------

counts_7895_1d <- Read10X(data.dir = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/EMTAB7895/MI1D/filtered_feature_bc_matrix")

seurat_7895_1d <- CreateSeuratObject(counts = counts_7895_1d, project = "EMTAB7895_1d")
```


#Quality Control

```{r Quality Control, include = False}

seurat_7895_1d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_7895_1d,
  pattern = "^mt-"
)

seurat_7895_1d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_7895_1d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_7895_1d$mitoRatio <- seurat_7895_1d$percent.mito / 100
seurat_7895_1d$riboRatio <- seurat_7895_1d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_7895_1d <- seurat_7895_1d@meta.data

# Assign metadata based on orig.ident patterns
metadata_7895_1d <- metadata_7895_1d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_7895_1d@meta.data <- metadata_7895_1d
seurat_7895_1d$log10GenesPerUMI <- log10(seurat_7895_1d$nFeature_RNA) / log10(seurat_7895_1d$nCount_RNA)

```



```{r Filtered_seurat, include = False}

seurat_7895_1d <- subset(x =seurat_7895_1d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_7895_1d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_7895_1d <- subset(seurat_7895_1d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_7895_1d <- CellCycleScoring(seurat_7895_1d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

seurat_7895_1d <- Seurat::SCTransform(seurat_7895_1d, method = "glmGamPoi")

seurat_7895_1d <- RunPCA(seurat_7895_1d)
seurat_7895_1d <- RunUMAP(seurat_7895_1d, dims = 1:30)


```



#--------------------
#EMTAB7895 3D
#--------------------


```{r}

#---------------------
#.     MI@3D
#--------------------

counts_7895_3d <- Read10X(data.dir = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/EMTAB7895/MI3D/filtered_feature_bc_matrix")

seurat_7895_3d <- CreateSeuratObject(counts = counts_7895_3d, project = "EMTAB7895_3d")

```


#Quality Control

```{r Quality Control, include = False}

seurat_7895_3d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_7895_3d,
  pattern = "^mt-"
)

seurat_7895_3d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_7895_3d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_7895_3d$mitoRatio <- seurat_7895_3d$percent.mito / 100
seurat_7895_3d$riboRatio <- seurat_7895_3d$percent.ribo / 100

```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_7895_3d <- seurat_7895_3d@meta.data

# Assign metadata based on orig.ident patterns
metadata_7895_3d <- metadata_7895_3d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_7895_3d@meta.data <- metadata_7895_3d
seurat_7895_3d$log10GenesPerUMI <- log10(seurat_7895_3d$nFeature_RNA) / log10(seurat_7895_3d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_7895_3d <- subset(x =seurat_7895_3d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_7895_3d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_7895_3d <- subset(seurat_7895_3d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_7895_3d <- CellCycleScoring(seurat_7895_3d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

seurat_7895_3d <- Seurat::SCTransform(seurat_7895_3d, method = "glmGamPoi")

seurat_7895_3d <- RunPCA(seurat_7895_3d)
seurat_7895_3d <- RunUMAP(seurat_7895_3d, dims = 1:30)


```




#----------------
#EMTAB7895 7D
#---------------


```{r}

#---------------------
#.     MI@7D
#--------------------

counts_7895_7d <- Read10X(data.dir = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/EMTAB7895/MI7D/filtered_feature_bc_matrix")

seurat_7895_7d <- CreateSeuratObject(counts = counts_7895_7d, project = "EMTAB7895_7d")

```


#Quality Control

```{r Quality Control, include = False}

seurat_7895_7d[["percent.mito"]] <- PercentageFeatureSet(
  object = seurat_7895_7d,
  pattern = "^mt-"
)

seurat_7895_7d[["percent.ribo"]] <- PercentageFeatureSet(
  object = seurat_7895_7d,
  pattern = "^Rp[sl]"
)

# Convert to ratios
seurat_7895_7d$mitoRatio <- seurat_7895_7d$percent.mito / 100
seurat_7895_7d$riboRatio <- seurat_7895_7d$percent.ribo / 100


```



#Meta Data Annotation
```{r Meta Data Annotation, include=FALSE}

# Extract metadata
metadata_7895_7d <- seurat_7895_7d@meta.data

# Assign metadata based on orig.ident patterns
metadata_7895_7d <- metadata_7895_7d %>%
  mutate(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
seurat_7895_7d@meta.data <- metadata_7895_7d
seurat_7895_7d$log10GenesPerUMI <- log10(seurat_7895_7d$nFeature_RNA) / log10(seurat_7895_7d$nCount_RNA)
```



```{r Filtered_seurat, include = False}

seurat_7895_7d <- subset(x =seurat_7895_7d, 
                          subset = (nUMI >= 500) & 
                                   (nGene >= 250) & 
                                   (log10GenesPerUMI > 0.80) & 
                                   (mitoRatio < 0.30))

# Collapse layers into one count matrix
counts <- GetAssayData(seurat_7895_7d, assay = "RNA", layer = "counts")

# Keep genes expressed in ≥10 cells
keep_genes <- rowSums(counts > 0) >= 10
seurat_7895_7d <- subset(seurat_7895_7d, features = rownames(counts)[keep_genes])

```


#################################################
        Cell Cycle Variability  
#################################################

```{r}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

#Converting gens to lower case (Ex: Xcr1, Naaa)
s_genes <- str_to_title(s.genes)
g2m_genes <- str_to_title(g2m.genes)

```


#################################################
     SCT TRANSFORM & DATA INTEGRATION 
#################################################


```{r SCTransform, include = FALSE}

library(SeuratData)

seurat_7895_7d <- CellCycleScoring(seurat_7895_7d, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

seurat_7895_7d <- Seurat::SCTransform(seurat_7895_7d, method = "glmGamPoi")

seurat_7895_7d <- RunPCA(seurat_7895_7d)
seurat_7895_7d <- RunUMAP(seurat_7895_7d, dims = 1:30)


```



```{r}
saveRDS(seurat_13264_HF, "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_13264_HF.rds")
saveRDS(seurat_9816_1d,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_9816_1d.rds")
saveRDS(seurat_9816_3d,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_9816_3d.rds")
saveRDS(seurat_9816_7d,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_9816_7d.rds")
saveRDS(seurat_Epelman,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_Epelman.rds")
saveRDS(seurat_7895_1d,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_7895_1d.rds")
saveRDS(seurat_7895_3d,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_7895_3d.rds")
saveRDS(seurat_7895_7d,  "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_7895_7d.rds")

```




# ── ATLAS INTEGRATION ──────────────────────────────────────────


```{r Building/Loading gene_map}

library(biomaRt)
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "mmusculus_gene_ensembl")

gene_map <- getBM(
  attributes = c("mgi_symbol", "ensembl_gene_id"),
  mart = mart
)
gene_map <- gene_map[gene_map$mgi_symbol != "" & gene_map$ensembl_gene_id != "", ]
gene_map <- unique(gene_map)

write.csv(gene_map,
          file = "/projects/b1246/Isaac/gene_map_mouse.csv",
          row.names = FALSE)


```



```{r SCT Integration of all the datasets}}

suppressPackageStartupMessages({
  library(Matrix)
  library(Seurat)
})

collapse_sparse_rows <- function(mat, groups) {
  if (!inherits(mat, "dgCMatrix")) mat <- as(mat, "dgCMatrix")
  groups <- as.character(groups)
  ug <- unique(groups)
  sm <- summary(mat)
  gi <- match(groups[sm$i], ug)
  res <- sparseMatrix(
    i = gi, j = sm$j, x = sm$x,
    dims = c(length(ug), ncol(mat)),
    dimnames = list(ug, colnames(mat))
  )
  methods::as(res, "dgCMatrix")
}

map_to_ensembl <- function(counts, symbol_to_ensembl) {
  old <- rownames(counts)
  new <- symbol_to_ensembl[old]
  new_ids <- ifelse(is.na(new), paste0("SYM:", old), new)
  collapse_sparse_rows(counts, new_ids)
}

ensure_gene_universe <- function(counts, target_genes) {
  rn <- rownames(counts)
  miss <- setdiff(target_genes, rn)
  if (length(miss)) {
    z <- sparseMatrix(i = integer(0), j = integer(0), x = numeric(0),
                      dims = c(length(miss), ncol(counts)),
                      dimnames = list(miss, colnames(counts)))
    counts <- rbind(counts, z)
  }
  counts[match(target_genes, rownames(counts)), , drop = FALSE]
}

safe_counts <- function(obj, assay = "RNA") GetAssayData(obj, assay = assay, slot = "counts")

# mapping prep
symbol_to_ensembl <- setNames(gene_map$ensembl_gene_id, gene_map$mgi_symbol)


```




```{r SCT Normalization}

seurat_list <- list(
  Epelman   = seurat_Epelman,
  `13264_HF`= seurat_13264_HF,
  `7895_1d` = seurat_7895_1d,
  `7895_3d` = seurat_7895_3d,
  `7895_7d` = seurat_7895_7d,
  `9816_1d` = seurat_9816_1d,
  `9816_3d` = seurat_9816_3d,
  `9816_7d` = seurat_9816_7d
)

for (nm in names(seurat_list)) {
  obj <- seurat_list[[nm]]
  obj <- RenameCells(obj, add.cell.id = nm)

  counts <- safe_counts(obj, assay = "RNA")
  counts_mapped <- map_to_ensembl(counts, symbol_to_ensembl)

  obj[["RNA.all"]] <- CreateAssayObject(counts = counts_mapped)
  DefaultAssay(obj) <- "RNA.all"
  seurat_list[[nm]] <- obj
}

all_genes <- Reduce(union, lapply(seurat_list, \(o) rownames(o[["RNA.all"]])))

for (nm in names(seurat_list)) {
  obj <- seurat_list[[nm]]
  counts <- GetAssayData(obj, assay = "RNA.all", slot = "counts")
  counts <- ensure_gene_universe(counts, all_genes)
  obj[["RNA.all"]] <- CreateAssayObject(counts = counts)
  DefaultAssay(obj) <- "RNA.all"
  seurat_list[[nm]] <- obj
}

# SCTransform normalization
seurat_list <- lapply(seurat_list, function(o) {
  DefaultAssay(o) <- "RNA.all"
  SCTransform(
    o, assay = "RNA.all", new.assay.name = "SCT",
    return.only.var.genes = FALSE,
    conserve.memory = FALSE,
    verbose = TRUE
  )
})

seurat_list <- lapply(seurat_list, function(o) { DefaultAssay(o) <- "SCT"; o })

present_in_both <- function(o) {
  rc <- rownames(GetAssayData(o, assay = "SCT", slot = "counts"))
  rd <- rownames(GetAssayData(o, assay = "SCT", slot = "data"))
  intersect(rc, rd)
}
by_obj_both <- lapply(seurat_list, present_in_both)

features_raw <- SelectIntegrationFeatures(object.list = seurat_list, nfeatures = 3000)
features_use <- Reduce(intersect, c(list(features_raw), by_obj_both))
features_use <- unique(features_use[nzchar(features_use)])
stopifnot(length(features_use) >= 2000)

seurat_list <- lapply(seurat_list, function(o) {
  DefaultAssay(o) <- "SCT"
  rn <- rownames(o[["SCT"]])
  if (anyDuplicated(rn)) {
    m <- GetAssayData(o, assay = "SCT", slot = "counts")
    keep <- !duplicated(rn)
    o[["SCT"]] <- CreateAssayObject(counts = m[keep, , drop = FALSE])
    DefaultAssay(o) <- "SCT"
  }
  VariableFeatures(o) <- features_use
  o
})

seurat_list <- PrepSCTIntegration(
  object.list     = seurat_list,
  anchor.features = features_use,
  verbose         = TRUE
)

anchors <- FindIntegrationAnchors(
  object.list          = seurat_list,
  normalization.method = "SCT",
  anchor.features      = features_use,
  dims                 = 1:30,
  verbose              = TRUE
)

seurat_atlas <- IntegrateData(
  anchorset            = anchors,
  normalization.method = "SCT",
  dims                 = 1:30,
  verbose              = TRUE
)

DefaultAssay(seurat_atlas) <- "integrated"
seurat_atlas <- seurat_atlas |>
  RunPCA(npcs = 30) |>
  RunUMAP(dims = 1:30) |>
  FindNeighbors(dims = 1:30) |>
  FindClusters(resolution = c(0.1,0.2,0.3,0.4, 0.5,0.6,0.7,0.8, 0.9))
DimPlot(seurat_atlas, reduction = "umap", label = TRUE)

```



```{r Gene Mapping}

# Load mapping
gene_map <- read.csv("/projects/b1246/Isaac/gene_map_mouse.csv")
symbol_to_ensembl <- setNames(gene_map$ensembl_gene_id, toupper(gene_map$mgi_symbol))
ensembl_to_symbol <- setNames(gene_map$mgi_symbol, gene_map$ensembl_gene_id)

# Helper for DotPlot using gene symbols
DotPlotSymbols <- function(object, symbols, mapping, group.by) {
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  ens_to_symbol <- setNames(mapping$mgi_symbol, mapping$ensembl_gene_id)

  features_map <- symbol_to_ens[toupper(symbols)]
  features_use <- features_map[!is.na(features_map)]

  p <- DotPlot(object, features = features_use, group.by = group.by) +
    RotatedAxis() +
    scale_x_discrete(labels = ens_to_symbol[features_use])
  return(p)
}



```




```{r}


# ── Remap to study-level groups ───────────────────────────────
umap_df_atlas$dataset_group <- dplyr::case_when(
  umap_df_atlas$dataset %in% c("EMTAB7895_1d",
                                "EMTAB7895_3d",
                                "EMTAB7895_7d")   ~ "E-MTAB-7895",
  umap_df_atlas$dataset %in% c("EMTAB9816_MI1D",
                                "EMTAB9816_MI3D",
                                "EMTAB9816_MI7D") ~ "E-MTAB-9816",
  umap_df_atlas$dataset == "GSE119355"            ~ "GSE119355",
  umap_df_atlas$dataset %in% c("Mice-HFrEF-n1",
                                "Mice-HFrEF-n2",
                                "Mice-HFrEF-n3",
                                "Mice-HFrEF-n4")  ~ "E-MTAB-13264"
)

# ── 4 fully saturated distinct colors ────────────────────────
group_colors <- c(
  "E-MTAB-7895"  = "#1F77B4",   # vivid blue
  "E-MTAB-9816"  = "#D62728",   # vivid red
  "GSE119355"  = "#2CA02C",   # vivid green
  "E-MTAB-13264"      = "#9467BD"    # vivid purple
)

# ── Plot ──────────────────────────────────────────────────────
p_dataset_umap <- ggplot(
  umap_df_atlas[sample(nrow(umap_df_atlas)), ],
  aes(x = UMAP_1, y = UMAP_2, color = dataset_group)) +
  geom_point(size = 0.3, alpha = 0.7, stroke = 0) +
  scale_color_manual(
    values = group_colors,
    name   = NULL,
    guide  = guide_legend(
      override.aes = list(size = 5, alpha = 1, shape = 16),
      ncol = 1
    )
  ) +
  umap_theme("right")

p_dataset_umap <- add_umap_arrows(p_dataset_umap, umap_df_atlas)
p_dataset_umap

save_figure("Fig1B_dataset_origin_UMAP",
            p_dataset_umap,
            file.path(atlas_fig_dir, "07_Final_Annotation"),
            200, 140)

```



####SAVING OBJECT######
```{r}
saveRDS(seurat_atlas, "/projects/b1246/Isaac/MERTK/Atlas_Datasets/seurat_atlas.rds")

```


#---------------------------------------------
ATLAS BIOLOGICAL ANNOTATION + VALIDATION
#---------------------------------------------


```{r Directory creation}


library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
library(ggrepel)

atlas_fig_dir <- "/projects/b1246/Isaac/MERTK/Atlas_Datasets/Figures"
dirs <- c(
  file.path(atlas_fig_dir, "01_Resolution_Selection"),
  file.path(atlas_fig_dir, "02_Broad_Lineage"),
  file.path(atlas_fig_dir, "03_Myeloid"),
  file.path(atlas_fig_dir, "04_Lymphoid"),
  file.path(atlas_fig_dir, "05_Endothelial"),
  file.path(atlas_fig_dir, "06_Fibroblast_Stromal"),
  file.path(atlas_fig_dir, "07_Final_Annotation"),
  file.path(atlas_fig_dir, "08_QC")
)
for (d in dirs) dir.create(d, recursive = TRUE, showWarnings = FALSE)

```


```{r Helper Functions, include=FALSE}

DotPlotSymbols <- function(object, symbols, mapping, group.by,
                            dot.scale = 8, cols = c("grey90", "#C0392B")) {
  
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  ens_to_symbol <- setNames(mapping$mgi_symbol, mapping$ensembl_gene_id)
  features_map  <- symbol_to_ens[toupper(symbols)]
  features_use  <- features_map[!is.na(features_map)]
  missing       <- symbols[is.na(features_map)]
  if (length(missing) > 0)
    message("Not found in gene_map: ", paste(missing, collapse = ", "))

  #Extract expression matrix manually
  groups     <- object@meta.data[[group.by]]
  group_lvls <- levels(factor(groups))

  expr_data <- GetAssayData(object, assay = DefaultAssay(object), layer = "data")
  expr_data  <- expr_data[features_use, , drop = FALSE]

  #Compute avg expression and pct expressed per group 
  results <- lapply(group_lvls, function(grp) {
    cells    <- which(groups == grp)
    mat      <- expr_data[, cells, drop = FALSE]
    avg_exp  <- rowMeans(mat)
    pct_exp  <- rowMeans(mat > 0) * 100
    data.frame(
      features.plot = features_use,
      id            = grp,
      avg.exp       = avg_exp,
      pct.exp       = pct_exp,
      stringsAsFactors = FALSE
    )
  })

  plot_df <- do.call(rbind, results)

  #Scale avg expression across groups per gene ─
  plot_df <- plot_df %>%
    dplyr::group_by(features.plot) %>%
    dplyr::mutate(avg.exp.scaled = as.numeric(scale(avg.exp))) %>%
    dplyr::ungroup()

  # Cap scaled values at -2.5 / 2.5 
  plot_df$avg.exp.scaled <- pmax(pmin(plot_df$avg.exp.scaled, 2.5), -2.5)

  #Replace Ensembl IDs with gene symbols 
  plot_df$gene_symbol <- factor(
    ens_to_symbol[plot_df$features.plot],
    levels = ens_to_symbol[features_use]
  )

  plot_df$id <- factor(plot_df$id, levels = group_lvls)

  # Build plot
  ggplot(plot_df,
         aes(x = gene_symbol, y = id,
             size = pct.exp, color = avg.exp.scaled)) +
    geom_point() +
    scale_color_gradient(low = cols[1], high = cols[2],
                         name = "Avg\nexpression\n(scaled)") +
    scale_size_continuous(range = c(0, dot.scale),
                          name  = "% expressed",
                          limits = c(0, 100)) +
    theme_classic(base_size = 10) +
    theme(
      axis.text.x      = element_text(angle = 45, hjust = 1, vjust = 1,
                                       size = 9, face = "italic"),
      axis.text.y      = element_text(size = 9, face = "bold"),
      axis.title       = element_blank(),
      panel.border     = element_rect(fill = NA, color = "black",
                                       linewidth = 0.5),
      panel.grid.major = element_line(color = "grey90", linewidth = 0.3),
      legend.text      = element_text(size = 8),
      legend.title     = element_text(size = 9, face = "bold"),
      plot.margin      = margin(5, 5, 5, 5)
    )
}



FeaturePlotSymbols <- function(object, symbols, mapping, ncol = 2, order = TRUE,
                                raster = FALSE, pt.size = 0.5,
                                min.cutoff = NA, max.cutoff = NA,
                                layer = "data", ...) {
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  ens_to_symbol <- setNames(mapping$mgi_symbol, mapping$ensembl_gene_id)
  features_map  <- symbol_to_ens[toupper(symbols)]
  features_use  <- features_map[!is.na(features_map)]
  missing       <- symbols[is.na(features_map)]
  
  if (length(missing) > 0)
    message("Not found in gene_map: ", paste(missing, collapse = ", "))
  
  p <- FeaturePlot(object, features = features_use, pt.size = pt.size,
                   order = order, ncol = ncol, min.cutoff = min.cutoff,
                   max.cutoff = max.cutoff, cols = c("grey90", "#C0392B"), ...)
  
  p <- p & theme_classic(base_size = 10) &
    theme(plot.title       = element_text(size = 10, face = "bold.italic"),
          axis.line        = element_blank(),
          axis.ticks       = element_blank(),
          axis.text        = element_blank(),
          axis.title       = element_blank(),
          legend.text      = element_text(size = 7),
          legend.title     = element_blank(),
          plot.margin      = margin(3, 3, 3, 3))
  
  for (i in seq_along(features_use)) {
    p[[i]] <- p[[i]] + ggtitle(ens_to_symbol[features_use[i]])
  }
  
  return(p)
}

AddModuleScoreSymbols <- function(object, symbols, mapping,
                                   name, assay = "SCT", ...) {
  symbol_to_ens <- setNames(mapping$ensembl_gene_id, toupper(mapping$mgi_symbol))
  features_ens  <- symbol_to_ens[toupper(symbols)]
  features_ens  <- features_ens[!is.na(features_ens)]
  present       <- rownames(GetAssayData(object[[assay]], layer = "data"))
  features_ens  <- intersect(features_ens, present)
  if (length(features_ens) < 2)
    stop("Not enough mapped features present in assay for module scoring")
  AddModuleScore(object, features = list(features_ens),
                 name = name, assay = assay, ...)
}
```




```{r Resolution Selection}

DefaultAssay(seurat_atlas) <- "integrated"

umap_df <- as.data.frame(Embeddings(seurat_atlas, "umap"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")

res_to_show <- c(0.1, 0.2, 0.3, 0.6, 0.8, 0.9)

p_res_panels <- lapply(res_to_show, function(r) {
  col    <- paste0("integrated_snn_res.", r)
  df     <- umap_df
  df$cl  <- as.character(seurat_atlas[[col]][, 1])
  n_cl   <- length(unique(df$cl))

  ggplot(df, aes(x = UMAP_1, y = UMAP_2, color = cl)) +
    geom_point(size = 0.15, alpha = 0.6, stroke = 0) +
    annotate("text",
             x = min(df$UMAP_1), y = max(df$UMAP_2),
             label = paste0("res = ", r, "\n", n_cl, " clusters"),
             hjust = 0, vjust = 1, size = 3, fontface = "bold", color = "black") +
    theme_classic(base_size = 9) +
    theme(axis.line = element_blank(), axis.ticks = element_blank(),
          axis.text = element_blank(), axis.title = element_blank(),
          panel.border = element_blank(), panel.background = element_blank(),
          plot.background = element_blank(), legend.position = "none",
          plot.margin = margin(3, 3, 3, 3))
})

p_res_grid <- wrap_plots(p_res_panels, ncol = 3) +
  plot_annotation(theme = theme(plot.title = element_text(face = "bold",
                                                           hjust = 0.5, size = 12)))
p_res_grid

ggsave("Resolution_comparison.pdf", plot = p_res_grid,
       path = file.path(atlas_fig_dir, "01_Resolution_Selection"),
       device = "pdf", width = 220, height = 150, units = "mm")

```


```{r Broad Lineage Dotplot at res 0.2, include = FALSE}

Idents(seurat_atlas) <- "integrated_snn_res.0.2"
DefaultAssay(seurat_atlas) <- "SCT"

# Ordered so each lineage's markers are grouped together
lineage_markers_02 <- c(
  # Myeloid
  "Lyz2", "Itgam", "Adgre1", "Csf1r", "S100a8", "S100a9",
  # T cell
  "Cd3e", "Cd4", "Cd8a",
  # B cell
  "Cd79a", "Ms4a1",
  # NK
  "Ncr1", "Klrb1c", "Gzmb",
  # DC
  "Itgax", "H2-Ab1", "Cd74", "Siglech",
  # Endothelial
  "Pecam1", "Cdh5", "Kdr",
  # Fibroblast
  "Col1a1", "Pdgfra", "Postn",
  # Neutrophil
  "Ly6g"
)

gene_map <- read.csv("/projects/b1246/Isaac/gene_map_mouse.csv")

p_dot_02 <- DotPlotSymbols(seurat_atlas,
                             symbols  = lineage_markers_02,
                             mapping  = gene_map,
                             group.by = "integrated_snn_res.0.2")
p_dot_02

ggsave(
  filename = "Res02_lineage_dotplot.tiff",
  plot     = p_dot_02,
  path     = file.path(atlas_fig_dir, "01_Resolution_Selection"),
  device   = "tiff",
  width    = 260,
  height   = 120,
  units    = "mm",
  dpi      = 600,                # 600 for mixed raster/vector; 300 is bare minimum
  compression = "lzw",           # lossless, keeps file size manageable
  bg       = "white"             # explicit white background, not transparent
)

```




```{r Broad lineage UMAP + marker check after cluster mapping}

Idents(seurat_atlas) <- "integrated_snn_res.0.2"
DefaultAssay(seurat_atlas) <- "SCT"

# Apply broad lineage mapping
lv  <- levels(seurat_atlas)
map <- setNames(rep("Other", length(lv)), lv)
map[c("0","3","4","5","9")] <- "Myeloid"
map["7"] <- "B cell"
map["6"] <- "Endothelial"
map[c("1","2")] <- "Fibroblast"
map["8"] <- "NKT cell"

seurat_atlas$lineage_r.02 <- plyr::mapvalues(
  x = as.character(Idents(seurat_atlas)),
  from = names(map), to = unname(map), warn_missing = FALSE
)

lineage_colors_02 <- c(
  "Myeloid"      = "#C0392B",
  "B cell"  = "#2980B9",
  "NKT cell"  = "#8E44AD",
  "Endothelial"  = "#27AE60",
  "Fibroblast"   = "#E67E22",
  "Other"        = "grey80"
)

umap_df$lineage_02 <- seurat_atlas$lineage_r.02

centroids_02 <- umap_df %>%
  filter(lineage_02 != "Other") %>%
  group_by(lineage_02) %>%
  summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")

p_broad_02 <- ggplot(
  umap_df[order(umap_df$lineage_02 == "Other", decreasing = TRUE), ],
  aes(x = UMAP_1, y = UMAP_2, color = lineage_02)) +
  geom_point(size = 0.25, alpha = 0.8, stroke = 0) +
  scale_color_manual(values = lineage_colors_02,
                     guide = guide_legend(
                       override.aes = list(size = 4, alpha = 1))) +
  geom_text_repel(data = centroids_02,
                  aes(x = UMAP_1, y = UMAP_2, label = lineage_02),
                  color = "black", size = 3.2, fontface = "bold",
                  box.padding = 0.4, segment.color = "grey50",
                  segment.size = 0.3, max.overlaps = 20, seed = 42) +
  theme_classic(base_size = 11) +
  theme(axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.border = element_blank(), panel.background = element_blank(),
        plot.background = element_blank(), legend.position = "right",
        legend.text = element_text(size = 9, face = "bold"),
        legend.title = element_blank(), plot.margin = margin(5, 5, 5, 5)) +
  # Axis arrows
  annotate("segment",
           x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+3.2,
           y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+0.2,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("segment",
           x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+0.2,
           y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+3.75,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("text", x = min(umap_df$UMAP_1)+1.7, y = min(umap_df$UMAP_2)-0.5,
           label = "UMAP-1", size = 3, color = "black",
           hjust = 0.5, fontface = "bold") +
  annotate("text", x = min(umap_df$UMAP_1)-0.5, y = min(umap_df$UMAP_2)+2.0,
           label = "UMAP-2", size = 3, color = "black",
           hjust = 0.5, angle = 90, fontface = "bold")

p_broad_02

# Sanity check DotPlot alongside the UMAP
p_sanity_02 <- DotPlotSymbols(seurat_atlas,
  symbols  = c("Lyz2","Itgam","Adgre1","Csf1r",
               "Cd3e","Cd4","Cd8a",
               "Ncr1","Klrb1c","Gzmb",
               "Cd79a","Ms4a1",
               "Pecam1","Cdh5","Kdr",
               "Col1a1","Pdgfra"),
  mapping  = gene_map,
  group.by = "lineage_r.02")

p_broad_02
p_sanity_02

ggsave(
  filename = "Broad_lineage_02_UMAP.tiff",
  plot     = p_broad_02,
  path     = file.path(atlas_fig_dir, "02_Broad_Lineage"),
  device   = "tiff",
  width    = 160,
  height   = 130,
  units    = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

ggsave(
  filename = "Broad_lineage_02_sanity_dotplot.tiff",
  plot     = p_sanity_02,
  path     = file.path(atlas_fig_dir, "02_Broad_Lineage"),
  device   = "tiff",
  width    = 240,
  height   = 100,
  units    = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)


```




```{r Myeloid Sanity Check FeaturePlots}


myeloid <- subset(seurat_atlas, subset = lineage_r.02 == "Myeloid")
DefaultAssay(myeloid) <- "SCT"

# Positive myeloid markers
p_myeloid_pos <- FeaturePlotSymbols(
  myeloid,
  symbols    = c("Itgam", "Lyz2", "Adgre1", "Csf1r"),
  mapping    = gene_map,
  ncol       = 4,
  order      = TRUE,
  raster     = FALSE,
  pt.size    = 0.6,
  min.cutoff = 0,
  max.cutoff = "q95"
)

# Negative controls 
p_myeloid_neg <- FeaturePlotSymbols(
  myeloid,
  symbols    = c("Cd79a", "Pecam1", "Col1a1", "Cd3e"),
  mapping    = gene_map,
  ncol       = 4,
  order      = TRUE,
  raster     = FALSE,
  pt.size    = 0.6,
  min.cutoff = 0,
  max.cutoff = "q95"
)

p_myeloid_check <- (p_myeloid_pos / p_myeloid_neg) +
  plot_annotation(
    title    = "Myeloid subset: positive  vs negative controls",
    theme    = theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 12),
      plot.subtitle = element_text(hjust = 0.5, size = 9, color = "grey40")
    )
  )

p_myeloid_check

ggsave(
  filename = "Myeloid_sanity_FeaturePlots.tiff",
  plot     = p_myeloid_check,
  path     = file.path(atlas_fig_dir, "03_Myeloid"),
  device   = "tiff",
  width    = 260,
  height   = 160,
  units    = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```




```{r Res 0.6 full marker DotPlots}

res <- 0.6
new_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- new_col
DefaultAssay(seurat_atlas) <- "SCT"

# Immune markers Dotplot
immune_markers_06 <- c(
  "Timd4","Lyve1","Folr2","Mrc1","Igf1","Cd163",  # TLF Mac
  "Ccr2","Trem2","Apoe","Cd14",                    # CCR2 Mac
  "Runx3","Cx3cr1","H2-Ab1","Cd74",                # MHCII Mac
  "Ly6c2","Csf1r","Itgam",                         # Monocyte
  "Xcr1","Clec9a","Siglech","Irf7",                # DCs
  "Cd3e","Ncr1","Klrb1c","Gzmb",                   # NK/T
  "Cd79a","Ms4a1","Cd19",                          # B
  "Ly6g","S100a8","S100a9"                         # Neutrophil
)

p_immune_06 <- DotPlotSymbols(seurat_atlas,
                               symbols  = immune_markers_06,
                               mapping  = gene_map,
                               group.by = new_col)

# Stromal markers Dotplot
stromal_markers_06 <- c(
  "Pecam1","Cdh5","Kdr","Eng",           # General EC
  "Id1","Mgll",                           # Arterial EC
  "Plvap","Vwf",                          # Venous EC
  "Col1a1","Pdgfra","Postn","Fn1",       # General Fibro
  "Cthrc1","Acta2","Tagln",               # RCFs
  "Comp","Sfrp2","Cilp",                  # Matrifibrocytes
  "Ccl2","Mt1","Timp1"                    # Ccl2 fibroblasts
)

p_stromal_06 <- DotPlotSymbols(seurat_atlas,
                                symbols  = stromal_markers_06,
                                mapping  = gene_map,
                                group.by = new_col)
p_immune_06
p_stromal_06


ggsave(
  filename = "Res06_immune_markers_dotplot.tiff",
  plot     = p_immune_06,
  path     = file.path(atlas_fig_dir, "01_Resolution_Selection"),
  device   = "tiff",
  width    = 280,
  height   = 140,
  units    = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

ggsave(
  filename = "Res06_stromal_markers_dotplot.tiff",
  plot     = p_stromal_06,
  path     = file.path(atlas_fig_dir, "01_Resolution_Selection"),
  device   = "tiff",
  width    = 280,
  height   = 140,
  units    = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)



```




```{r Res 0.6 UMAP with lineage labels}

res <- 0.6
res0.6_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res0.6_col

# Defining identities 
map_06 <- c(
  `4`  = "Mac_Resident/TLF+",
  `6`  = "Mac_CCR2+",
  `2`  = "Mac_MHCII^hi",
  `9`  = "Monocyte",
  `17` = "DCs",
  `16` = "NKT",
  `13` = "B cells",
  `1`  = "Neutrophils",
  # Endothelial reported as 23 + 5 → give each the same label
  `23` = "Endothelial",
  `5`  = "Endothelial",
  # Fibroblasts reported as 0,7,11,15,18,21
  `0`  = "Fibroblast",
  `7`  = "Fibroblast",
  `11` = "Fibroblast",
  `15` = "Fibroblast",
  `18` = "Fibroblast",
  `21` = "Fibroblast"
)

# 2) Create a new metadata column (don’t overwrite the old one)
col_lab <- "lineage_r.06"

# Convert cluster ids to character to index map safely
cl_ids <- as.character(seurat_atlas[[res0.6_col]][,1])

# Apply mapping; anything not listed becomes "Other"
annot <- unname(map_06[cl_ids])
annot[is.na(annot)] <- "Other"

seurat_atlas[[col_lab]] <- factor(
  annot,
  levels = c(
    "Mac_Resident/TLF+","Mac_CCR2+","Mac_MHCII^hi","Monocyte",
    "DCs","NKT","B cells","Neutrophils","Endothelial","Fibroblast","Other"
  )
)


```




```{r Res 0.6 UMAP with lineage labels}

library(ggrepel)
lineage_colors_06 <- c(
  "Mac_Resident/TLF+" = "#C0392B",
  "Mac_CCR2+"         = "#E67E22",
  "Mac_MHCII^hi"      = "#27AE60",
  "Monocyte"          = "#F39C12",
  "DCs"               = "#8E44AD",
  "NKT"               = "#CC79A7",
  "B cells"           = "#2980B9",
  "Neutrophils"       = "#0072B2",
  "Endothelial"       = "#1ABC9C",
  "Fibroblast"        = "#D35400",
  "Other"             = "grey80"
)

umap_df$lineage_06 <- as.character(seurat_atlas$lineage_r.06)

centroids_06 <- umap_df %>%
  filter(lineage_06 != "Other") %>%
  group_by(lineage_06) %>%
  summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")

p_broad_06 <- ggplot(
  umap_df[order(umap_df$lineage_06 == "Other", decreasing = TRUE), ],
  aes(x = UMAP_1, y = UMAP_2, color = lineage_06)) +
  geom_point(size = 0.25, alpha = 0.8, stroke = 0) +
  scale_color_manual(values = lineage_colors_06,
                     guide = guide_legend(
                       override.aes = list(size = 4, alpha = 1))) +
  geom_text_repel(data = centroids_06,
                  aes(x = UMAP_1, y = UMAP_2, label = lineage_06),
                  color = "black", size = 3, fontface = "bold",
                  box.padding = 0.4, segment.color = "grey50",
                  segment.size = 0.3, max.overlaps = 20, seed = 42) +
  theme_classic(base_size = 11) +
  theme(axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.border = element_blank(), panel.background = element_blank(),
        plot.background = element_blank(), legend.position = "right",
        legend.text = element_text(size = 9, face = "bold"),
        legend.title = element_blank(), plot.margin = margin(5, 5, 5, 5)) +
  annotate("segment",
           x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+3.2,
           y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+0.2,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("segment",
           x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+0.2,
           y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+3.75,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("text", x = min(umap_df$UMAP_1)+1.7, y = min(umap_df$UMAP_2)-0.5,
           label = "UMAP-1", size = 3, color = "black",
           hjust = 0.5, fontface = "bold") +
  annotate("text", x = min(umap_df$UMAP_1)-0.5, y = min(umap_df$UMAP_2)+2.0,
           label = "UMAP-2", size = 3, color = "black",
           hjust = 0.5, angle = 90, fontface = "bold")

p_broad_06 

ggsave(
  filename    = "Broad_lineage_06_UMAP.tiff",
  plot        = p_broad_06,
  path        = file.path(atlas_fig_dir, "02_Broad_Lineage"),
  device      = "tiff",
  width       = 180, height = 130, units = "mm",
  dpi = 600, compression = "lzw", bg = "white"
)


```




```{r Res 0.8 myeloid markers + monocyte discrimination}


res <- 0.8
new_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas)      <- new_col
DefaultAssay(seurat_atlas) <- "SCT"

#TLF Mac 
p_dot_TLF <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("Timd4","Lyve1","Folr2","Ninj1",
                                           "Gas6","Mrc1","Cd163","Igf1",
                                           "Spp1","Igfb4"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("TLF Mac markers")

#C1q+ TLF Mac
p_dot_C1q <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("C1qa","C1qb","Mrc1","Igf1"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("C1q+ TLF Mac markers")

#CCR2+ Mac 
p_dot_CCR2 <- DotPlotSymbols(seurat_atlas,
                               symbols  = c("Ccr2","Cd52","H2-Eb1","H2-Ab1",
                                            "Bcl2a1a","Coro1a","Cyp4f18","Fxdy5"),
                               mapping  = gene_map,
                               group.by = new_col) +
  ggtitle("CCR2+ Mac markers")

#MHCII Mac 
p_dot_MHCII <- DotPlotSymbols(seurat_atlas,
                                symbols  = c("Cd14","Cst3","Lilra5","Apoe",
                                             "H3f3b","C5ar1","St3gal6"),
                                mapping  = gene_map,
                                group.by = new_col) +
  ggtitle("MHCII Mac markers")

#SPP1+ Mac
p_dot_Spp1 <- DotPlotSymbols(seurat_atlas,
                               symbols  = c("Spp1","Trem2","Lipa"),
                               mapping  = gene_map,
                               group.by = new_col) +
  ggtitle("SPP1+ Mac markers")

#Prolif TLF Mac
p_dot_Prolif <- DotPlotSymbols(seurat_atlas,
                                symbols  = c("Birc5","Mki67","C1qa"),
                                mapping  = gene_map,
                                group.by = new_col) +
  ggtitle("Prolif TLF Mac markers")

#GPNMB Mac 
p_dot_Gpnmb <- DotPlotSymbols(seurat_atlas,
                                symbols  = c("Gpnmb","Fabp5","Ctsd","Syngr1"),
                                mapping  = gene_map,
                                group.by = new_col) +
  ggtitle("GPNMB Mac markers")

#Monocytes
p_dot_Mono <- DotPlotSymbols(seurat_atlas,
                               symbols  = c("Ccr2","Ly6c2","Cx3cr1",
                                            "Treml4","Plac8"),
                               mapping  = gene_map,
                               group.by = new_col) +
  ggtitle("Monocyte markers: Ly6C-hi vs Ly6C-lo")


p_dot_TLF
p_dot_C1q
p_dot_CCR2
p_dot_MHCII
p_dot_Spp1
p_dot_Prolif
p_dot_Gpnmb
p_dot_Mono

save_plots <- list(
  "Res08_TLF_Mac"        = p_dot_TLF,
  "Res08_C1q_TLF_Mac"    = p_dot_C1q,
  "Res08_CCR2_Mac"       = p_dot_CCR2,
  "Res08_MHCII_Mac"      = p_dot_MHCII,
  "Res08_Spp1_Mac"       = p_dot_Spp1,
  "Res08_Prolif_TLF_Mac" = p_dot_Prolif,
  "Res08_Gpnmb_Mac"      = p_dot_Gpnmb,
  "Res08_Monocytes"      = p_dot_Mono
)

for (nm in names(save_plots)) {
  ggsave(
    filename    = paste0(nm, ".tiff"),
    plot        = save_plots[[nm]],
    path        = file.path(atlas_fig_dir, "03_Myeloid"),
    device      = "tiff",
    width       = 220,
    height      = 130,
    units       = "mm",
    dpi         = 600,
    compression = "lzw",
    bg          = "white"
  )
}



#monocyte markers
features_monocyte <- c("Ccr2","Ly6c2","Plac8")

# Publication DotPlot replacing the old bare DotPlotSymbols call
p_mono_dot <- DotPlotSymbols(seurat_atlas,
                               symbols  = features_monocyte,
                               mapping  = gene_map,
                               group.by = new_col)
p_mono_dot


ggsave(
  filename    = "Res08_monocyte_markers_dotplot.tiff",
  plot        = p_mono_dot,
  path        = file.path(atlas_fig_dir, "03_Myeloid"),
  device      = "tiff",
  width       = 180,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
       

# VlnPlot: Ly6C-hi vs Ly6C-lo discrimination 
mono_obj <- subset(seurat_atlas,
                    subset = lineage_r.08 %in% c("Ly6C-hi Mono","Ly6C-lo Mono"))
DefaultAssay(mono_obj) <- "SCT"

mono_markers_ens <- setNames(gene_map$ensembl_gene_id,
                              toupper(gene_map$mgi_symbol))
mono_ens <- mono_markers_ens[toupper(features_monocyte)]
mono_ens <- mono_ens[!is.na(mono_ens)]
ens_to_sym <- setNames(gene_map$mgi_symbol, gene_map$ensembl_gene_id)

p_mono_vln <- VlnPlot(mono_obj,
                       features = mono_ens,
                       group.by = "lineage_r.08",
                       pt.size  = 0,
                       ncol     = 5,
                       cols     = c("#D55E00","#E69F00")) &
  theme_classic(base_size = 10) &
  theme(axis.title.x   = element_blank(),
        axis.text.x    = element_text(angle = 45, hjust = 1, face = "bold"),
        plot.title     = element_text(size = 9, face = "bold.italic"),
        axis.line      = element_line(linewidth = 0.5),
        axis.ticks     = element_line(linewidth = 0.4),
        legend.position = "none")

for (i in seq_along(mono_ens)) {
  p_mono_vln[[i]] <- p_mono_vln[[i]] + ggtitle(ens_to_sym[mono_ens[i]])
}

p_mono_vln

ggsave(
  filename    = "Monocyte_Ly6Chi_vs_Lo_VlnPlot.tiff",
  plot        = p_mono_vln,
  path        = file.path(atlas_fig_dir, "03_Myeloid"),
  device      = "tiff",
  width       = 220,
  height      = 80,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```






```{r Endothelial + Fibroblast annotation at res0.8}

# ══════════════════════════════════════════════════════════════
# Endothelial SUBTYPES
# ══════════════════════════════════════════════════════════════

res     <- 0.8
new_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas)      <- new_col
DefaultAssay(seurat_atlas) <- "SCT"


#General ECs
p_dot_EC <- DotPlotSymbols(seurat_atlas,
                             symbols  = c("Pecam1","Cdh5","Cd36","Fabp4",
                                          "Eng","Esam","Kdr"),
                             mapping  = gene_map,
                             group.by = new_col) +
  ggtitle("General EC markers")

#Arterial ECs
p_dot_Art <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("Id1","Mgll"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("Arterial EC markers")

#Venous ECs
p_dot_Ven <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("Plvap","Vwf"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("Venous EC markers")

#LECs
p_dot_LEC <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("Lyve1","Prox1","Flt4",
                                           "Nrp2","Cd31","Pdpn","Reln"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("LEC markers")

#Immunomodulatory ECs
p_dot_Imm <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("Ptprc","Cdh5"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("Immunomodulatory EC markers")

#Save Endothelial plots
endo_plots <- list(
  "Res08_EC_general"        = p_dot_EC,
  "Res08_EC_arterial"       = p_dot_Art,
  "Res08_EC_venous"         = p_dot_Ven,
  "Res08_LEC"               = p_dot_LEC,
  "Res08_EC_immunomod"      = p_dot_Imm
)

for (nm in names(endo_plots)) {
  ggsave(
    filename    = paste0(nm, ".tiff"),
    plot        = endo_plots[[nm]],
    path        = file.path(atlas_fig_dir, "05_Endothelial"),
    device      = "tiff",
    width       = 220,
    height      = 130,
    units       = "mm",
    dpi         = 600,
    compression = "lzw",
    bg          = "white"
  )
}


# ══════════════════════════════════════════════════════════════
# FIBROBLAST SUBTYPES
# ══════════════════════════════════════════════════════════════

#RCFs 
p_dot_RCF <- DotPlotSymbols(seurat_atlas,
                              symbols  = c("Cthrc1","Acta2","Fn1",
                                           "Col1a1","Postn","Tagln"),
                              mapping  = gene_map,
                              group.by = new_col) +
  ggtitle("RCF markers")

#Matrifibrocytes
p_dot_Matr <- DotPlotSymbols(seurat_atlas,
                               symbols  = c("Comp","Sfrp2","Cilp",
                                            "Eln","Wisp2"),
                               mapping  = gene_map,
                               group.by = new_col) +
  ggtitle("Matrifibrocyte markers")

#Ccl2 fibroblasts
p_dot_Ccl2 <- DotPlotSymbols(seurat_atlas,
                               symbols  = c("Mt2","Ccl2","Timp1",
                                            "Mt1","Prg4","Angptl4"),
                               mapping  = gene_map,
                               group.by = new_col) +
  ggtitle("Ccl2 fibroblast markers")

#Prolif fibroblasts
p_dot_ProlFib <- DotPlotSymbols(seurat_atlas,
                                  symbols  = c("Stmn1","H2afz","Cks2",
                                               "Cenpa","Acta2","Hmgb2"),
                                  mapping  = gene_map,
                                  group.by = new_col) +
  ggtitle("Prolif fibroblast markers")


p_dot_EC
p_dot_Art
p_dot_Ven
p_dot_LEC
p_dot_Imm
p_dot_RCF
p_dot_Matr
p_dot_Ccl2
p_dot_ProlFib


#Save Fibroblast plots
fibro_plots <- list(
  "Res08_RCF"               = p_dot_RCF,
  "Res08_Matrifibrocytes"   = p_dot_Matr,
  "Res08_Ccl2_fibroblasts"  = p_dot_Ccl2,
  "Res08_Prolif_fibroblasts"= p_dot_ProlFib
)

for (nm in names(fibro_plots)) {
  ggsave(
    filename    = paste0(nm, ".tiff"),
    plot        = fibro_plots[[nm]],
    path        = file.path(atlas_fig_dir, "06_Fibroblast_Stromal"),
    device      = "tiff",
    width       = 220,
    height      = 130,
    units       = "mm",
    dpi         = 600,
    compression = "lzw",
    bg          = "white"
  )
}


```






```{r  Final annotation UMAP at res 0.8}
res     <- 0.8
res_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res_col
col_lab <- "lineage_r.08"

map_08 <- c(
  `2`  = "MHCII Mac",
  `3`  = "CCR2+ Mac",
  `7`  = "TLF Mac",
  `14` = "TLF Mac",
  `9`  = "Ly6C-hi Mono",
  `22` = "Ly6C-lo Mono",
  `12` = "B Cells",
  `17` = "ARG1+ Mac",
  `24` = "Prolif TLF Mac",
  `19` = "NKT",
  `20` = "DCs",
  `1`  = "Neutrophils",
  `26` = "Arterial ECs",
  `21` = "Venous ECs",
  `5`  = "ECs",
  `6`  = "RCFs",
  `4`  = "RCFs"
)

cl_ids <- Idents(seurat_atlas)
annot  <- map_08[as.character(cl_ids)]
annot[is.na(annot)] <- "Other"
seurat_atlas@meta.data[[col_lab]] <- annot

lvl_order <- c(
  "Ly6C-hi Mono","Ly6C-lo Mono","Neutrophils","DCs",
  "MHCII Mac","CCR2+ Mac","TLF Mac","ARG1+ Mac","Prolif TLF Mac",
  "B Cells","NKT",
  "Arterial ECs","Venous ECs","ECs","RCFs","Other"
)
vals <- seurat_atlas@meta.data[[col_lab]]
seurat_atlas@meta.data[[col_lab]] <- factor(vals,
                                             levels = intersect(lvl_order,
                                                                unique(vals)))
pal_hc <- c(
  "Ly6C-hi Mono"   = "#D55E00",
  "Ly6C-lo Mono"   = "#E69F00",
  "Neutrophils"    = "#0072B2",
  "DCs"            = "#56B4E9",
  "MHCII Mac"      = "#009E73",
  "CCR2+ Mac"      = "#851033",
  "TLF Mac"        = "#2C7FB8",
  "ARG1+ Mac"      = "#33A02C",
  "Prolif TLF Mac" = "#A6D854",
  "B Cells"        = "#8C564B",
  "NKT"            = "#CC79A7",
  "Arterial ECs"   = "#FB9A99",
  "Venous ECs"     = "#B15928",
  "ECs"            = "#6A3D9A",
  "RCFs"           = "#F0E442",
  "Other"          = "#BDBDBD"
)

seurat_atlas[[col_lab]] <- NULL
seurat_atlas <- Seurat::AddMetaData(
  seurat_atlas,
  metadata = setNames(
    data.frame(lineage = factor(annot, levels = lvl_order),
               row.names = colnames(seurat_atlas)),
    col_lab
  )
)
Seurat::Idents(seurat_atlas) <- col_lab

if (is.null(seurat_atlas@misc)) seurat_atlas@misc <- list()
seurat_atlas@misc$annotation_maps <- modifyList(
  seurat_atlas@misc$annotation_maps %||% list(),
  list("r0.8" = map_08)
)
seurat_atlas@misc$palettes <- modifyList(
  seurat_atlas@misc$palettes %||% list(),
  list("lineage_r.08" = pal_hc)
)


Idents(seurat_atlas) <- "lineage_r.08"
DefaultAssay(seurat_atlas) <- "SCT"

umap_df$celltype_08 <- as.character(seurat_atlas$lineage_r.08)

centroids_08 <- umap_df %>%
  filter(!celltype_08 %in% c("Other","Unassigned stromal")) %>%
  group_by(celltype_08) %>%
  summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")

p_final_umap <- ggplot(
  umap_df[order(umap_df$celltype_08 %in% c("Other","Unassigned stromal"),
                decreasing = TRUE), ],
  aes(x = UMAP_1, y = UMAP_2, color = celltype_08)) +
  geom_point(size = 0.25, alpha = 0.8, stroke = 0) +
  scale_color_manual(values = pal_hc,
                     guide = guide_legend(
                       override.aes = list(size = 4, alpha = 1),
                       ncol = 1)) +
  geom_text_repel(data = centroids_08,
                  aes(x = UMAP_1, y = UMAP_2, label = celltype_08),
                  color = "black", size = 2.5, fontface = "bold",
                  box.padding = 0.3, point.padding = 0.1,
                  segment.color = "grey50", segment.size = 0.25,
                  max.overlaps = 30, seed = 42) +
  theme_classic(base_size = 11) +
  theme(axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.border = element_blank(), panel.background = element_blank(),
        plot.background = element_blank(), legend.position = "right",
        legend.text = element_text(size = 8, face = "bold"),
        legend.title = element_blank(),
        legend.key.size = unit(0.35, "cm"),
        plot.margin = margin(5, 5, 5, 5)) +
  annotate("segment",
           x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+3.2,
           y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+0.2,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("segment",
           x = min(umap_df$UMAP_1)+0.2, xend = min(umap_df$UMAP_1)+0.2,
           y = min(umap_df$UMAP_2)+0.2, yend = min(umap_df$UMAP_2)+3.75,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("text", x = min(umap_df$UMAP_1)+1.7, y = min(umap_df$UMAP_2)-0.5,
           label = "UMAP-1", size = 3, color = "black",
           hjust = 0.5, fontface = "bold") +
  annotate("text", x = min(umap_df$UMAP_1)-0.5, y = min(umap_df$UMAP_2)+2.0,
           label = "UMAP-2", size = 3, color = "black",
           hjust = 0.5, angle = 90, fontface = "bold")

p_final_umap


ggsave(
  filename = "Atlas_final_annotation_UMAP.tiff",
  plot     = p_final_umap,
  path     = file.path(atlas_fig_dir, "07_Final_Annotation"),
  device   = "tiff",
  width    = 200,
  height   = 150,
  units    = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```



```{r Macrophage/Monocyte Only UMAP}

mac_ids <- c("Ly6C-hi Mono","Ly6C-lo Mono",
             "MHCII Mac","CCR2+ Mac","TLF Mac",
             "ARG1+ Mac","Prolif TLF Mac")

mac_obj <- subset(seurat_atlas,
                   subset = lineage_r.08 %in% mac_ids)
mac_obj$lineage_r.08 <- droplevels(factor(mac_obj$lineage_r.08,
                                           levels = mac_ids))

umap_mac <- as.data.frame(Embeddings(mac_obj, "umap"))
colnames(umap_mac) <- c("UMAP_1", "UMAP_2")
umap_mac$celltype <- mac_obj$lineage_r.08

centroids_mac <- umap_mac %>%
  group_by(celltype) %>%
  summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")

pal_mac <- pal_hc[mac_ids]

p_mac_umap <- ggplot(umap_mac,
                      aes(x = UMAP_1, y = UMAP_2, color = celltype)) +
  geom_point(size = 0.4, alpha = 0.8, stroke = 0) +
  scale_color_manual(values = pal_mac,
                     guide = guide_legend(
                       override.aes = list(size = 4, alpha = 1))) +
  geom_text_repel(data = centroids_mac,
                  aes(x = UMAP_1, y = UMAP_2, label = celltype),
                  color = "black", size = 3, fontface = "bold",
                  box.padding = 0.4, segment.color = "grey50",
                  segment.size = 0.3, max.overlaps = 20, seed = 42) +
  theme_classic(base_size = 11) +
  theme(axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.border = element_blank(), panel.background = element_blank(),
        plot.background = element_blank(), legend.position = "right",
        legend.text = element_text(size = 9, face = "bold"),
        legend.title = element_blank(), plot.margin = margin(5, 5, 5, 5)) +
  annotate("segment",
           x = min(umap_mac$UMAP_1)+0.2, xend = min(umap_mac$UMAP_1)+3.2,
           y = min(umap_mac$UMAP_2)+0.2, yend = min(umap_mac$UMAP_2)+0.2,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("segment",
           x = min(umap_mac$UMAP_1)+0.2, xend = min(umap_mac$UMAP_1)+0.2,
           y = min(umap_mac$UMAP_2)+0.2, yend = min(umap_mac$UMAP_2)+3.75,
           arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
           color = "black", linewidth = 0.8) +
  annotate("text", x = min(umap_mac$UMAP_1)+1.7, y = min(umap_mac$UMAP_2)-0.5,
           label = "UMAP-1", size = 3, color = "black",
           hjust = 0.5, fontface = "bold") +
  annotate("text", x = min(umap_mac$UMAP_1)-0.5, y = min(umap_mac$UMAP_2)+2.0,
           label = "UMAP-2", size = 3, color = "black",
           hjust = 0.5, angle = 90, fontface = "bold")

p_mac_umap


ggsave(
  filename    = "Macrophage_monocyte_UMAP.tiff",
  plot        = p_mac_umap,
  path        = file.path(atlas_fig_dir, "03_Myeloid"),
  device      = "tiff",
  width       = 180,
  height      = 130,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
```




```{r Filtering Apoptotic Cells }

Idents(seurat_atlas) <- new_col
DefaultAssay(seurat_atlas) <- "SCT"

apoptotic_markers <- c("Bax","Bak1","Casp3","Casp9","Ddit3","Atf4","Mki67")

p_apop_dot <- DotPlotSymbols(seurat_atlas,
                               symbols  = apoptotic_markers,
                               mapping  = gene_map,
                               group.by = new_col) +
  ggtitle("Apoptosis markers")
p_apop_dot

seurat_atlas <- AddModuleScoreSymbols(
  seurat_atlas,
  symbols = c("Bax","Bak1","Bbc3","Bcl2l11",
              "Casp3","Casp8","Casp9",
              "Ddit3","Atf4","Atf3",
              "Hspa1a","Hspa1b"),
  mapping = gene_map,
  name    = "ApoptosisScores",
  assay   = "SCT"
)

p_apop_vln <- VlnPlot(seurat_atlas,
                       features = "ApoptosisScores1",
                       group.by = new_col,
                       pt.size  = 0) +
  theme_classic(base_size = 10) +
  theme(axis.text.x     = element_text(angle = 45, hjust = 1, size = 8,
                                        face = "bold"),
        axis.title.x    = element_blank(),
        axis.title.y    = element_text(size = 9),
        axis.line       = element_line(linewidth = 0.5),
        axis.ticks      = element_line(linewidth = 0.4),
        legend.position = "none",
        plot.title      = element_text(size = 10, face = "bold")) +
  ggtitle("Apoptosis module score per cluster") +
  ylab("Apoptosis score")

p_apop_vln

ggsave(
  filename    = "QC_apoptosis_dotplot.tiff",
  plot        = p_apop_dot,
  path        = file.path(atlas_fig_dir, "08_QC"),
  device      = "tiff",
  width       = 260,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
ggsave(
  filename    = "QC_apoptosis_vlnplot.tiff",
  plot        = p_apop_vln,
  path        = file.path(atlas_fig_dir, "08_QC"),
  device      = "tiff",
  width       = 260,
  height      = 100,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
```



```{r Fibroblast module score FeaturePlots}

fibro_panels <- list(
  RCFs             = c("Cthrc1","Acta2","Fn1","Col1a1","Postn","Tagln"),
  Matrifibrocytes  = c("Comp","Sfrp2","Cilp","Eln","Wisp2"),
  Ccl2_fibroblasts = c("Mt2","Ccl2","Timp1","Mt1","Prg4","Angptl4"),
  Prolif_RCFs      = c("Stmn1","H2afz","Cks2","Cenpa","Mki67","Hmgb2")
)

for (nm in names(fibro_panels)) {
  seurat_atlas <- AddModuleScoreSymbols(
    seurat_atlas,
    symbols = fibro_panels[[nm]],
    mapping = gene_map,
    name    = paste0("Fib_", nm),
    assay   = "SCT"
  )
}

fibro_score_features <- c("Fib_RCFs1","Fib_Matrifibrocytes1",
                            "Fib_Ccl2_fibroblasts1","Fib_Prolif_RCFs1")

p_fibro_scores <- FeaturePlot(seurat_atlas,
                               features  = fibro_score_features,
                               reduction = "umap",
                               order     = TRUE,
                               ncol      = 2) &
  scale_color_gradient(low = "grey90", high = "#8E44AD") &
  theme_classic(base_size = 10) &
  theme(axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.border = element_blank(), plot.title = element_text(size = 9, face = "bold"))

p_fibro_scores[[1]] <- p_fibro_scores[[1]] + ggtitle("RCFs score")
p_fibro_scores[[2]] <- p_fibro_scores[[2]] + ggtitle("Matrifibrocytes score")
p_fibro_scores[[3]] <- p_fibro_scores[[3]] + ggtitle("Ccl2 fibroblasts score")
p_fibro_scores[[4]] <- p_fibro_scores[[4]] + ggtitle("Prolif RCFs score")

p_fibro_scores

ggsave(
  filename    = "Fibroblast_module_scores_FeaturePlots.tiff",
  plot        = p_fibro_scores,
  path        = file.path(atlas_fig_dir, "06_Fibroblast_Stromal"),
  device      = "tiff",
  width       = 180,
  height      = 150,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)


```



```{r Differential expression cluster 11,8,23}

DefaultAssay(seurat_atlas) <- "SCT"
Idents(seurat_atlas)       <- new_col
seurat_atlas <- PrepSCTFindMarkers(seurat_atlas)

#FindMarkers
markers_11 <- FindMarkers(seurat_atlas, ident.1 = 11,
                           min.pct = 0.25, logfc.threshold = 0.25)
markers_23 <- FindMarkers(seurat_atlas, ident.1 = 23,
                           min.pct = 0.25, logfc.threshold = 0.25)
markers_8  <- FindMarkers(seurat_atlas, ident.1 = 8,
                           min.pct = 0.25, logfc.threshold = 0.25)

#Annotate with gene symbols
annotate_markers <- function(markers_df, gene_map) {
  markers_df$ensembl_gene_id <- rownames(markers_df)
  out <- merge(markers_df, gene_map, by = "ensembl_gene_id", all.x = TRUE)
  colnames(out)[colnames(out) == "mgi_symbol"] <- "gene_symbol"
  out <- out[order(out$avg_log2FC, decreasing = TRUE), ]
  return(out)
}

markers_11_annot <- annotate_markers(markers_11, gene_map)
markers_23_annot <- annotate_markers(markers_23, gene_map)
markers_8_annot  <- annotate_markers(markers_8,  gene_map)


print(head(markers_11_annot[, c("gene_symbol","avg_log2FC","p_val_adj")], 20))
print(head(markers_23_annot[, c("gene_symbol","avg_log2FC","p_val_adj")], 20))
print(head(markers_8_annot[,  c("gene_symbol","avg_log2FC","p_val_adj")], 20))

#Volcano plot function
make_volcano <- function(markers_annot, cluster_label,
                          fc_thresh = 1, pval_thresh = 0.05) {
  df <- markers_annot
  df$sig <- "NS"
  df$sig[df$avg_log2FC >  fc_thresh & df$p_val_adj < pval_thresh] <- "Up"
  df$sig[df$avg_log2FC < -fc_thresh & df$p_val_adj < pval_thresh] <- "Down"
  df$sig <- factor(df$sig, levels = c("Up","Down","NS"))

  top_up   <- head(df[df$sig == "Up",   ], 10)
  top_down <- head(df[df$sig == "Down", ], 10)
  to_label <- rbind(top_up, top_down)

  ggplot(df, aes(x = avg_log2FC,
                  y = -log10(p_val_adj + 1e-300),
                  color = sig)) +
    geom_point(size = 0.8, alpha = 0.7, stroke = 0) +
    geom_text_repel(data         = to_label,
                    aes(label    = gene_symbol),
                    size         = 2.5,
                    fontface     = "italic",
                    max.overlaps = 20,
                    seed         = 42,
                    segment.size = 0.3,
                    segment.color = "grey50") +
    scale_color_manual(values = c("Up"   = "#C0392B",
                                   "Down" = "#2980B9",
                                   "NS"   = "grey70"),
                       name = NULL) +
    geom_vline(xintercept = c(-fc_thresh, fc_thresh),
               linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_hline(yintercept = -log10(pval_thresh),
               linetype = "dashed", color = "grey50", linewidth = 0.4) +
    theme_classic(base_size = 10) +
    theme(
      plot.title      = element_text(face = "bold", size = 10, hjust = 0.5),
      axis.text       = element_text(size = 8, color = "black", face = "bold"),
      axis.title      = element_text(size = 9, face = "bold"),
      axis.line       = element_line(linewidth = 0.5),
      axis.ticks      = element_line(linewidth = 0.4),
      legend.position = "top",
      legend.text     = element_text(size = 8)
    ) +
    xlab("avg log2FC") +
    ylab("-log10(adj p-value)") +
    ggtitle(paste0("Cluster ", cluster_label, " - differential markers"))
}

#Build volcano plots
p_vol_11 <- make_volcano(markers_11_annot, "11")
p_vol_23 <- make_volcano(markers_23_annot, "23")
p_vol_8  <- make_volcano(markers_8_annot,  "8")

p_vol_11
p_vol_23
p_vol_8

#Save volcano plots
for (nm in list(list(p_vol_11, "Cluster11"),
                list(p_vol_23, "Cluster23"),
                list(p_vol_8,  "Cluster8"))) {
  ggsave(
    filename    = paste0("DEG_volcano_", nm[[2]], ".tiff"),
    plot        = nm[[1]],
    path        = file.path(atlas_fig_dir, "07_Final_Annotation"),
    device      = "tiff",
    width       = 120,
    height      = 100,
    units       = "mm",
    dpi         = 600,
    compression = "lzw",
    bg          = "white"
  )
}


#Stromal validation FeaturePlots 
stromal_labels <- c("Fibroblasts","VSMCs",
                     "Cardiomyocytes","Fibroblast progenitors",
                     "RCFs","Transitional fibroblasts","Prolif RCFs",
                     "Matrifibrocytes","Ccl2 fibroblasts",
                     "Unassigned stromal")

stromal_obj <- subset(seurat_atlas,
                       subset = lineage_r.08 %in% stromal_labels)
DefaultAssay(stromal_obj) <- "SCT"

p_stromal_check <- FeaturePlotSymbols(
  stromal_obj,
  symbols    = c("S100a4","Fap","Postn",
                 "Tagln","Myh11","Pdgfrb"),
  mapping    = gene_map,
  ncol       = 3,
  order      = TRUE,
  raster     = FALSE,
  pt.size    = 0.5,
  min.cutoff = 0,
  max.cutoff = "q95"
)

p_stromal_check

ggsave("Cluster1123_stromal_FeaturePlots.tiff",
       plot   = p_stromal_check,
       path   = file.path(atlas_fig_dir, "07_Final_Annotation"),
       device = "pdf",
       width  = 220, height = 150, units = "mm")

#Cardiomyocyte validation FeaturePlots
cardio_obj <- subset(seurat_atlas,
                      subset = lineage_r.08 == "Cardiomyocytes")
DefaultAssay(cardio_obj) <- "SCT"

p_cardio_check <- FeaturePlotSymbols(
  cardio_obj,
  symbols    = c("Tnnt2","Myh6","Myl2","Ttn"),
  mapping    = gene_map,
  ncol       = 4,
  order      = TRUE,
  raster     = FALSE,
  pt.size    = 0.8,
  min.cutoff = 0,
  max.cutoff = "q95"
)

p_cardio_check

ggsave(
  filename    = "Cluster8_cardiomyocyte_FeaturePlots.tiff",
  plot        = p_cardio_check,
  path        = file.path(atlas_fig_dir, "07_Final_Annotation"),
  device      = "tiff",
  width       = 220,
  height      = 80,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

# Save marker tables
write.csv(markers_11_annot,
          file.path(atlas_fig_dir, "07_Final_Annotation",
                    "DEG_cluster11_annotated.csv"),
          row.names = FALSE)
write.csv(markers_23_annot,
          file.path(atlas_fig_dir, "07_Final_Annotation",
                    "DEG_cluster23_annotated.csv"),
          row.names = FALSE)
write.csv(markers_8_annot,
          file.path(atlas_fig_dir, "07_Final_Annotation",
                    "DEG_cluster8_annotated.csv"),
          row.names = FALSE)

```






```{r Reannotation of UMAP}
res     <- 0.8
res_col <- paste0("integrated_snn_res.", res)
Idents(seurat_atlas) <- res_col

map_08 <- c(
  # Macs & monos
  `2`  = "MHCII Mac",
  `3`  = "CCR2+ Mac",
  `7`  = "TLF Mac",
  `14` = "TLF Mac",
  `9`  = "Ly6C-hi Mono",
  `22` = "Ly6C-lo Mono",
  `12` = "B Cells",
  `17` = "ARG1+ Mac",
  `24` = "Prolif TLF Mac",
  `19` = "NKT",
  `20` = "DCs",
  `1`  = "Neutrophils",

  # Endothelial
  `26` = "Arterial ECs",
  `21` = "Venous ECs",
  `5`  = "ECs",

  # Fibroblast lineage
  `16` = "Fibroblast progenitors",
  `0`  = "Matrifibrocytes",
  `15` = "Ccl2 fibroblasts",
  `13` = "Ccl2 fibroblasts",
  `4`  = "Transitional fibroblasts",
  `6`  = "VSMCs",
  `10` = "RCFs",
  `25` = "Prolif RCFs",

  # Stromal branches
  `11` = "Cardiomyocytes",
  `8`  = "Cardiomyocytes",
  `23` = "VSMCs"
)

col_lab <- "lineage_r.08"
cl_ids  <- Idents(seurat_atlas)
annot   <- map_08[as.character(cl_ids)]
annot[is.na(annot)] <- "Unassigned stromal"
seurat_atlas@meta.data[[col_lab]] <- annot

lvl_order <- c(
  "Ly6C-hi Mono","Ly6C-lo Mono","Neutrophils","DCs",
  "MHCII Mac","CCR2+ Mac","TLF Mac","ARG1+ Mac","Prolif TLF Mac",
  "B Cells","NKT",
  "Arterial ECs","Venous ECs","ECs",
  "Fibroblast progenitors",
  "RCFs","Transitional fibroblasts","Prolif RCFs",
  "Matrifibrocytes",
  "Ccl2 fibroblasts","Fibroblasts","VSMCs",
  "Cardiomyocytes",
  "Unassigned stromal"
)

vals <- seurat_atlas@meta.data[[col_lab]]
seurat_atlas@meta.data[[col_lab]] <- factor(vals,
                                             levels = intersect(lvl_order,
                                                                unique(vals)))

pal_hc <- c(
  "Ly6C-hi Mono"             = "#D55E00",
  "Ly6C-lo Mono"             = "#E69F00",
  "Neutrophils"              = "#0072B2",
  "DCs"                      = "#56B4E9",
  "MHCII Mac"                = "#009E73",
  "CCR2+ Mac"                = "#851033",
  "TLF Mac"                  = "#2C7FB8",
  "ARG1+ Mac"                = "#33A02C",
  "Prolif TLF Mac"           = "#A6D854",
  "B Cells"                  = "#8C564B",
  "NKT"                      = "#CC79A7",
  "Arterial ECs"             = "#FB9A99",
  "Venous ECs"               = "#B15928",
  "ECs"                      = "#6A3D9A",
  "Fibroblast progenitors"   = "#D13F21",
  "RCFs"                     = "#F0E442",
  "Transitional fibroblasts" = "#FFD700",
  "Prolif RCFs"              = "#FFB347",
  "Matrifibrocytes"          = "#FF7F00",
  "Ccl2 fibroblasts"         = "#800080",
  "Fibroblasts"              = "#A0522D",
  "VSMCs"                    = "#6421D1",
  "Cardiomyocytes"           = "#7A9134",
  "Unassigned stromal"       = "grey80"
)

pal_plot <- pal_hc[levels(seurat_atlas@meta.data[[col_lab]])]

print(sort(table(seurat_atlas$lineage_r.08), decreasing = TRUE))

# UMAP Creation
Idents(seurat_atlas)       <- col_lab
DefaultAssay(seurat_atlas) <- "SCT"

umap_df$celltype_08 <- as.character(seurat_atlas$lineage_r.08)

centroids_08 <- umap_df %>%
  filter(!celltype_08 %in% c("Unassigned stromal")) %>%
  group_by(celltype_08) %>%
  summarise(UMAP_1 = median(UMAP_1),
            UMAP_2 = median(UMAP_2),
            .groups = "drop")

p_final_umap <- ggplot(
  umap_df[order(umap_df$celltype_08 == "Unassigned stromal",
                decreasing = TRUE), ],
  aes(x = UMAP_1, y = UMAP_2, color = celltype_08)) +
  geom_point(size = 0.25, alpha = 0.8, stroke = 0) +
  scale_color_manual(values = pal_hc,
                     guide  = guide_legend(
                       override.aes = list(size = 4, alpha = 1),
                       ncol = 1)) +
  geom_text_repel(data          = centroids_08,
                  aes(x = UMAP_1, y = UMAP_2, label = celltype_08),
                  color         = "black",
                  size          = 2.5,
                  fontface      = "bold",
                  box.padding   = 0.3,
                  point.padding = 0.1,
                  segment.color = "grey50",
                  segment.size  = 0.25,
                  max.overlaps  = 30,
                  seed          = 42) +
  theme_classic(base_size = 11) +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    panel.border     = element_blank(),
    panel.background = element_blank(),
    plot.background  = element_blank(),
    legend.position  = "right",
    legend.text      = element_text(size = 8, face = "bold"),
    legend.title     = element_blank(),
    legend.key.size  = unit(0.35, "cm"),
    plot.margin      = margin(5, 5, 5, 5)
  ) +
  annotate("segment",
           x    = min(umap_df$UMAP_1) + 0.2,
           xend = min(umap_df$UMAP_1) + 3.2,
           y    = min(umap_df$UMAP_2) + 0.2,
           yend = min(umap_df$UMAP_2) + 0.2,
           arrow     = arrow(length = unit(0.2, "cm"), type = "closed"),
           color     = "black", linewidth = 0.8) +
  annotate("segment",
           x    = min(umap_df$UMAP_1) + 0.2,
           xend = min(umap_df$UMAP_1) + 0.2,
           y    = min(umap_df$UMAP_2) + 0.2,
           yend = min(umap_df$UMAP_2) + 3.75,
           arrow     = arrow(length = unit(0.2, "cm"), type = "closed"),
           color     = "black", linewidth = 0.8) +
  annotate("text",
           x      = min(umap_df$UMAP_1) + 1.7,
           y      = min(umap_df$UMAP_2) - 0.5,
           label  = "UMAP-1", size = 3, color = "black",
           hjust  = 0.5, fontface = "bold") +
  annotate("text",
           x      = min(umap_df$UMAP_1) - 0.5,
           y      = min(umap_df$UMAP_2) + 2.0,
           label  = "UMAP-2", size = 3, color = "black",
           hjust  = 0.5, angle = 90, fontface = "bold")

p_final_umap


ggsave(
  filename    = "Atlas_final_annotation_UMAP.tiff",
  plot        = p_final_umap,
  path        = file.path(atlas_fig_dir, "07_Final_Annotation"),
  device      = "tiff",
  width       = 200,
  height      = 150,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```

```{r}
# ── Master DotPlot at resolution 0.8 (Fig 1C) ────────────────────────────────
Idents(seurat_atlas)       <- "lineage_r.08"
DefaultAssay(seurat_atlas) <- "SCT"

master_markers_08 <- c(
  # Myeloid — monocytes
  "Ly6c2", "Ccr2", "Plac8",           # Ly6C-hi Mono
  "Cx3cr1", "Treml4",                  # Ly6C-lo Mono
  # Myeloid — macrophages
  "Timd4", "Lyve1", "Folr2", "Mrc1",  # TLF Mac
  "Spp1", "Trem2",                     # ARG1+ Mac
  "H2-Ab1", "Cd74", "Cd14",           # MHCII Mac
  "Bcl2a1a", "Cyp4f18",               # CCR2+ Mac (Ccr2 removed — already above)
  "Mki67", "Birc5",                    # Prolif TLF Mac
  # Other immune
  "Ly6g", "S100a8",                    # Neutrophils
  "Itgax", "Siglech",                  # DCs
  "Cd79a", "Ms4a1",                    # B Cells
  "Ncr1", "Klrb1c", "Gzmb",           # NKT
  # Endothelial
  "Pecam1", "Cdh5", "Kdr",            # ECs
  "Id1", "Mgll",                       # Arterial ECs
  "Plvap", "Vwf",                      # Venous ECs
  # Fibroblast / Stromal
  "Col1a1", "Pdgfra",                  # General fibroblast
  "Cthrc1", "Acta2", "Tagln",         # RCFs
  "Comp", "Sfrp2", "Cilp",            # Matrifibrocytes
  "Ccl2", "Mt1", "Timp1",             # Ccl2 fibroblasts
  "Stmn1", "Hmgb2",                   # Prolif RCFs (H2afz → Hmgb2)
  "Rgs5", "Kcnj8",                    # VSMCs
  "Scn5a", "Tnnt2", "Myl2"           # Cardiomyocytes
)

# Deduplicate defensively before passing
master_markers_08 <- unique(master_markers_08)

p_master_dot_08 <- DotPlotSymbols(
  seurat_atlas,
  symbols  = master_markers_08,
  mapping  = gene_map,
  group.by = "lineage_r.08"
)

p_master_dot_08

ggsave(
  filename    = "Atlas_master_dotplot_res08.tiff",
  plot        = p_master_dot_08,
  path        = file.path(atlas_fig_dir, "07_Final_Annotation"),
  device      = "tiff",
  width       = 380,
  height      = 180,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
  

```




```{r}
saveRDS(seurat_atlas, "/projects/b1246/Isaac/seurat_atlas.rds")

```



###########################################

# CREATING THE SINGLE R COMPATIBLE OBJECT

###########################################

```{r SINGLE R ATLAS CREATION}

## --- Setup
Atlas <- seurat_atlas
DefaultAssay(Atlas) <- "RNA"

## --- Extract raw counts from Seurat v5 Assay5
expr_counts <- Atlas[["RNA"]]$counts

## --- Ensure unique gene names
expr_counts <- expr_counts[!duplicated(rownames(expr_counts)), , drop = FALSE]

## --- Align labels with counts
cells_counts <- colnames(expr_counts)              # cells in counts
cells_meta   <- rownames(Atlas@meta.data)          # cells in metadata

# Subset metadata to same cells and order
labels <- as.character(Atlas@meta.data[cells_counts, "lineage_r.08"])

## --- Keep only valid cells
keep <- !is.na(labels) & labels != ""
expr_counts <- expr_counts[, keep, drop = FALSE]
labels <- labels[keep]

## --- Build SingleCellExperiment
library(SingleCellExperiment)
sce_reference <- SingleCellExperiment(assays = list(counts = expr_counts))
colData(sce_reference)$label <- labels

## --- Filter: keep labels with >=20 cells
cell_counts <- table(labels)
good_labels <- names(cell_counts[cell_counts >= 20])
sce_reference <- sce_reference[, labels %in% good_labels]
labels <- colData(sce_reference)$label

## --- Aggregate to pseudobulk by label
library(scuttle)
sce_reference_pb <- aggregateAcrossCells(
  sce_reference,
  ids = labels
)

## --- Normalize (adds logcounts)
library(scater)
sce_reference_pb <- logNormCounts(sce_reference_pb)

## --- Sanity checks
dim(sce_reference_pb)              # genes × pseudobulk samples
colnames(sce_reference_pb)         # pseudobulk labels kept
assayNames(sce_reference_pb)       # "counts", "logcounts"
table(labels)                      # cell counts per label (post-filter)

## --- Barplot (after filtering)
barplot(table(labels), las = 2, cex.names = 0.7,
        main = "Cells per pseudobulk label (≥20 filter)", col = "steelblue")



```

#Saving reference

```{r}
saveRDS(sce_reference_pb, 
        file = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/atlas_reference.rds")
```




###########################################
PSEUDOTIME 
##########################################


```{r Pseudotime Set up}

library(monocle3)
library(SeuratWrappers)
library(dplyr)
library(ggplot2)
library(patchwork)

pseudotime_dir <- file.path(atlas_fig_dir, "09_Pseudotime")
dir.create(pseudotime_dir, recursive = TRUE, showWarnings = FALSE)

```


```{r Function definition}
style_monocle_plot <- function(p, title_label = NULL) {
  
  p <- p +
    theme_classic(base_size = 11) +
    theme(
      axis.line        = element_blank(),
      axis.ticks       = element_blank(),
      axis.text        = element_blank(),
      axis.title       = element_blank(),
      panel.border     = element_blank(),
      panel.background = element_blank(),
      panel.grid       = element_blank(),
      plot.background  = element_blank(),
      plot.title       = element_blank(),
      legend.position  = "right",
      legend.text      = element_text(size = 8, face = "bold"),
      legend.title     = element_blank(),
      legend.key.size  = unit(0.35, "cm"),
      plot.margin      = margin(5, 5, 5, 5)
    )
  
  x_min <- min(p$data$data_dim_1, na.rm = TRUE)
  y_min <- min(p$data$data_dim_2, na.rm = TRUE)
  
  p <- p +
    annotate("segment",
             x = x_min + 0.2, xend = x_min + 3.2,
             y = y_min + 0.2, yend = y_min + 0.2,
             arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
             color = "black", linewidth = 0.8) +
    annotate("segment",
             x = x_min + 0.2, xend = x_min + 0.2,
             y = y_min + 0.2, yend = y_min + 3.75,
             arrow = arrow(length = unit(0.2,"cm"), type = "closed"),
             color = "black", linewidth = 0.8) +
    annotate("text",
             x = x_min + 1.7, y = y_min - 0.5,
             label = "UMAP-1", size = 3, color = "black",
             hjust = 0.5, fontface = "bold") +
    annotate("text",
             x = x_min - 0.5, y = y_min + 2.0,
             label = "UMAP-2", size = 3, color = "black",
             hjust = 0.5, angle = 90, fontface = "bold")
  
  return(p)
}

make_pt_vln <- function(cds_obj, title_label) {
  pt_df <- data.frame(
    pseudotime = pseudotime(cds_obj),
    celltype   = cds_obj@colData$lineage_r.08,
    stringsAsFactors = FALSE
  )
  pt_df <- pt_df[is.finite(pt_df$pseudotime), ]

  ct_order <- pt_df %>%
    group_by(celltype) %>%
    summarise(med_pt = median(pseudotime), .groups = "drop") %>%
    arrange(med_pt) %>%
    pull(celltype)

  pt_df$celltype <- factor(pt_df$celltype, levels = ct_order)

  ggplot(pt_df, aes(x = celltype, y = pseudotime, fill = celltype)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.85) +
    geom_boxplot(width = 0.1, outlier.size = 0, fill = "white",
                 color = "grey30", linewidth = 0.4) +
    scale_fill_manual(values = pal_hc, guide = "none") +
    scale_y_continuous(expand = c(0, 0)) +
    coord_flip() +
    theme_classic(base_size = 10) +
    theme(
      axis.text.y        = element_text(size = 8, face = "bold",
                                         color = "black"),
      axis.text.x        = element_text(size = 8, color = "black",
                                         face = "bold"),
      axis.title.x       = element_text(size = 9, face = "bold"),
      axis.title.y       = element_blank(),
      axis.line          = element_line(linewidth = 0.5),
      axis.ticks         = element_line(linewidth = 0.4),
      panel.grid.major.x = element_line(color = "grey90",
                                         linewidth = 0.3),
      plot.margin        = margin(5, 10, 5, 5)
    ) +
    xlab("") +
    ylab("Pseudotime")
}

pal_hc <- c(
  "Ly6C-hi Mono"             = "#D55E00",
  "Ly6C-lo Mono"             = "#E69F00",
  "Neutrophils"              = "#0072B2",
  "DCs"                      = "#56B4E9",
  "MHCII Mac"                = "#009E73",
  "CCR2+ Mac"                = "#851033",
  "TLF Mac"                  = "#2C7FB8",
  "ARG1+ Mac"                = "#33A02C",
  "Prolif TLF Mac"           = "#A6D854",
  "B Cells"                  = "#8C564B",
  "NKT"                      = "#CC79A7",
  "Arterial ECs"             = "#FB9A99",
  "Venous ECs"               = "#B15928",
  "ECs"                      = "#6A3D9A",
  "Fibroblast progenitors"   = "#D13F21",
  "RCFs"                     = "#F0E442",
  "Transitional fibroblasts" = "#FFD700",
  "Prolif RCFs"              = "#FFB347",
  "Matrifibrocytes"          = "#FF7F00",
  "Ccl2 fibroblasts"         = "#800080",
  "Fibroblasts"              = "#A0522D",
  "VSMCs"                    = "#6421D1",
  "Cardiomyocytes"           = "#7A9134",
  "Unassigned stromal"       = "grey80",
  "Other"                    = "grey80"
)
```

```{r  Convert Seurat to CellDataSet}
library(monocle3)

# ══════════════════════════════════════════════════════════════
# STEP 1 — Convert Seurat to CellDataSet 
# ══════════════════════════════════════════════════════════════

# Use lineage_r.08 as the cell type label
Idents(seurat_atlas)       <- "lineage_r.08"
DefaultAssay(seurat_atlas) <- "RNA"

# Join layers — required for Seurat v5
seurat_atlas <- JoinLayers(seurat_atlas)

# Extract components manually (bypasses deprecated PackageCheck)
counts    <- GetAssayData(seurat_atlas, assay = "RNA", layer = "counts")
cell_meta <- seurat_atlas@meta.data
gene_meta <- data.frame(
  gene_short_name = rownames(counts),
  row.names       = rownames(counts)
)

# Build CDS manually
cds <- new_cell_data_set(
  expression_data = counts,
  cell_metadata   = cell_meta,
  gene_metadata   = gene_meta
)

# Transfer lineage_r.08 labels to CDS
cds@colData$lineage_r.08 <- seurat_atlas$lineage_r.08

# Transfer UMAP coordinates from Seurat
reducedDim(cds, "UMAP") <- Embeddings(seurat_atlas, "umap")

print(table(cds@colData$lineage_r.08))

```

```{r Preprocess and cluster within Monocle3}

cds <- preprocess_cds(cds, num_dim = 30)

# Seurat's UMAP - preserving embedding
cds <- reduce_dimension(cds,
                         reduction_method    = "UMAP",
                         preprocess_method   = "PCA",
                         umap.fast_sgd       = FALSE,
                         cores               = 1,
                         verbose             = TRUE)

#Seurat UMAP coordinates directly
reducedDim(cds, "UMAP") <- Embeddings(seurat_atlas, "umap")

# Clustering cells using Seurat's cluster assignments
cds@clusters$UMAP$clusters <- factor(seurat_atlas$lineage_r.08)
cds@clusters$UMAP$partitions <- factor(rep(1, ncol(cds)))


```




```{r Learn trajectory graph}

# ── Run pseudotime separately per biological compartment ───────
# Rationale: immune and stromal cells have independent trajectories
# Forcing them into one graph creates biologically impossible paths

# Myeloid compartment only
myeloid_types <- c("Ly6C-hi Mono","Ly6C-lo Mono","Neutrophils","DCs",
                    "MHCII Mac","CCR2+ Mac","TLF Mac","ARG1+ Mac",
                    "Prolif TLF Mac","B Cells","NKT")

cds_myeloid <- cds[, cds@colData$lineage_r.08 %in% myeloid_types]

cds_myeloid <- cluster_cells(cds_myeloid,
                              reduction_method = "UMAP",
                              verbose          = TRUE)

# Override partitions
cds_myeloid@clusters$UMAP$partitions <- setNames(
  factor(rep(1, ncol(cds_myeloid))),
  colnames(cds_myeloid)
)

cds_myeloid <- learn_graph(cds_myeloid,
                             use_partition       = FALSE,
                             close_loop          = FALSE,
                             learn_graph_control = list(
                               minimal_branch_len = 10
                             ),
                             verbose = TRUE)

root_myeloid <- colnames(cds_myeloid)[
  cds_myeloid@colData$lineage_r.08 == "Ly6C-hi Mono"
]
cds_myeloid <- order_cells(cds_myeloid, root_cells = root_myeloid)


# Stromal compartment only
stromal_types <- c("Fibroblast progenitors","RCFs",
                    "Transitional fibroblasts","Prolif RCFs",
                    "Matrifibrocytes","Ccl2 fibroblasts",
                    "Fibroblasts","VSMCs",
                    "Cardiomyocytes","Arterial ECs","Venous ECs","ECs")

cds_stromal <- cds[, cds@colData$lineage_r.08 %in% stromal_types]

cds_stromal <- cluster_cells(cds_stromal,
                              reduction_method = "UMAP",
                              verbose          = TRUE)

cds_stromal@clusters$UMAP$partitions <- setNames(
  factor(rep(1, ncol(cds_stromal))),
  colnames(cds_stromal)
)

cds_stromal <- learn_graph(cds_stromal,
                             use_partition       = FALSE,
                             close_loop          = FALSE,
                             learn_graph_control = list(
                               minimal_branch_len = 10
                             ),
                             verbose = TRUE)

# Root = Fibroblast progenitors — earliest stromal state
root_stromal <- colnames(cds_stromal)[
  cds_stromal@colData$lineage_r.08 == "Fibroblast progenitors"
]
cds_stromal <- order_cells(cds_stromal, root_cells = root_stromal)


# ── Extract trajectory segments from CDS ─────────────────────
get_trajectory_segments <- function(cds_obj) {
  graph  <- cds_obj@principal_graph$UMAP
  coords <- t(cds_obj@principal_graph_aux$UMAP$dp_mst)
  edges  <- igraph::as_edgelist(graph)
  
  data.frame(
    x    = coords[edges[,1], 1],
    y    = coords[edges[,1], 2],
    xend = coords[edges[,2], 1],
    yend = coords[edges[,2], 2]
  )
}

# ── Build trajectory plot from scratch ───────────────────────
make_trajectory_plot <- function(cds_obj, pal) {
  
  umap_coords <- reducedDim(cds_obj, "UMAP")
  plot_data <- data.frame(
    UMAP_1   = umap_coords[, 1],
    UMAP_2   = umap_coords[, 2],
    celltype = cds_obj@colData$lineage_r.08
  )
  
  segs <- get_trajectory_segments(cds_obj)
  
  p <- ggplot(plot_data,
               aes(x = UMAP_1, y = UMAP_2, color = celltype)) +
    geom_point(size = 0.5, alpha = 0.8, stroke = 0) +
    geom_segment(data = segs,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 color     = "black",
                 linewidth = 0.8,
                 inherit.aes = FALSE) +
    scale_color_manual(values = pal,
                       guide  = guide_legend(
                         override.aes = list(size = 4, alpha = 1),
                         ncol = 1)) +
    theme_classic(base_size = 11) +
    theme(
      axis.line        = element_blank(),
      axis.ticks       = element_blank(),
      axis.text        = element_blank(),
      axis.title       = element_blank(),
      panel.border     = element_blank(),
      panel.background = element_blank(),
      plot.background  = element_blank(),
      legend.position  = "right",
      legend.text      = element_text(size = 8, face = "bold"),
      legend.title     = element_blank(),
      legend.key.size  = unit(0.4, "cm"),
      plot.margin      = margin(5, 5, 5, 5)
    )
  
  # Add axis arrows
  x0 <- min(plot_data$UMAP_1) + 0.2
  y0 <- min(plot_data$UMAP_2) + 0.2
  
  p <- p +
    annotate("segment",
             x = x0, xend = x0 + 3.0,
             y = y0, yend = y0,
             arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
             color = "black", linewidth = 0.7) +
    annotate("segment",
             x = x0, xend = x0,
             y = y0, yend = y0 + 3.0,
             arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
             color = "black", linewidth = 0.7) +
    annotate("text",
             x = x0 + 1.5, y = y0 - 0.6,
             label = "UMAP-1", size = 3,
             color = "black", hjust = 0.5, fontface = "bold") +
    annotate("text",
             x = x0 - 0.6, y = y0 + 1.5,
             label = "UMAP-2", size = 3,
             color = "black", hjust = 0.5,
             angle = 90, fontface = "bold")
  return(p)
}

p_traj_myeloid <- make_trajectory_plot(cds_myeloid, pal_hc)
p_traj_stromal <- make_trajectory_plot(cds_stromal, pal_hc)

p_traj_myeloid
p_traj_stromal

#save
ggsave(
  filename    = "Pseudotime_myeloid_trajectory.tiff",
  plot        = p_traj_myeloid,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 180,
  height      = 130,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
ggsave(
  filename    = "Pseudotime_stromal_trajectory.tiff",
  plot        = p_traj_stromal,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 180,
  height      = 130,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```




```{r Order cells / set root}

root_myeloid <- colnames(cds_myeloid)[
  cds_myeloid@colData$lineage_r.08 == "Ly6C-hi Mono"
]
cds_myeloid <- order_cells(cds_myeloid, root_cells = root_myeloid)

root_stromal <- colnames(cds_stromal)[
  cds_stromal@colData$lineage_r.08 == "Fibroblast progenitors"
]
cds_stromal <- order_cells(cds_stromal, root_cells = root_stromal)

make_pseudotime_plot <- function(cds_obj) {
  
  umap_coords <- reducedDim(cds_obj, "UMAP")
  pt          <- pseudotime(cds_obj)
  
  plot_data <- data.frame(
    UMAP_1     = umap_coords[, 1],
    UMAP_2     = umap_coords[, 2],
    pseudotime = ifelse(is.infinite(pt), NA, pt)
  )
  
  # Plot finite pseudotime cells on top
  plot_data <- plot_data[order(is.na(plot_data$pseudotime),
                                plot_data$pseudotime,
                                na.last = FALSE), ]
  
  segs <- get_trajectory_segments(cds_obj)
  
  p <- ggplot(plot_data,
               aes(x = UMAP_1, y = UMAP_2, color = pseudotime)) +
    geom_point(size = 0.5, alpha = 0.8, stroke = 0) +
    geom_segment(data = segs,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 color       = "black",
                 linewidth   = 0.8,
                 inherit.aes = FALSE) +
    scale_color_viridis_c(
      option  = "plasma",
      name    = "Pseudotime",
      na.value = "grey85",
      guide   = guide_colorbar(
        barwidth    = 0.5,
        barheight   = 4,
        ticks       = TRUE,
        title.theme = element_text(size = 9, face = "bold", angle = 0),
        label.theme = element_text(size = 8, face = "bold")
      )
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.line        = element_blank(),
      axis.ticks       = element_blank(),
      axis.text        = element_blank(),
      axis.title       = element_blank(),
      panel.border     = element_blank(),
      panel.background = element_blank(),
      plot.background  = element_blank(),
      legend.position  = "right",
      legend.text      = element_text(size = 8, face = "bold"),
      legend.title     = element_text(size = 9, face = "bold"),
      legend.key.size  = unit(0.4, "cm"),
      plot.margin      = margin(5, 5, 5, 5)
    )
  
  # Add axis arrows
  x0 <- min(plot_data$UMAP_1) + 0.2
  y0 <- min(plot_data$UMAP_2) + 0.2
  
  p <- p +
    annotate("segment",
             x = x0, xend = x0 + 3.0,
             y = y0, yend = y0,
             arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
             color = "black", linewidth = 0.7) +
    annotate("segment",
             x = x0, xend = x0,
             y = y0, yend = y0 + 3.0,
             arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
             color = "black", linewidth = 0.7) +
    annotate("text",
             x = x0 + 1.5, y = y0 - 0.6,
             label = "UMAP-1", size = 3,
             color = "black", hjust = 0.5, fontface = "bold") +
    annotate("text",
             x = x0 - 0.6, y = y0 + 1.5,
             label = "UMAP-2", size = 3,
             color = "black", hjust = 0.5,
             angle = 90, fontface = "bold")
  return(p)
}

p_pt_myeloid_umap <- make_pseudotime_plot(cds_myeloid)
p_pt_stromal_umap <- make_pseudotime_plot(cds_stromal)

p_pt_myeloid_umap
p_pt_stromal_umap


ggsave(
  filename    = "Pseudotime_myeloid_umap.tiff",
  plot        = p_pt_myeloid_umap,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 160,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
ggsave(
  filename    = "Pseudotime_stromal_umap.tiff",
  plot        = p_pt_stromal_umap,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 160,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```



```{r Pseudotime VlnPlot per cell type}


p_pt_vln_myeloid <- make_pt_vln(cds_myeloid, "Myeloid")
p_pt_vln_stromal <- make_pt_vln(cds_stromal, "Stromal")

p_pt_vln_myeloid
p_pt_vln_stromal

ggsave(
  filename    = "Pseudotime_myeloid_vlnplot.tiff",
  plot        = p_pt_vln_myeloid,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 120,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

ggsave(
  filename    = "Pseudotime_stromal_vlnplot.tiff",
  plot        = p_pt_vln_stromal,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 120,
  height      = 140,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)
```
#Plots including days_post_MI on the x-axis (Reviewer Clarification)

```{r Pseudotime Vln plot (Myeloid)}

cds_myeloid_pt$day_post_MI <- NA_real_

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "EMTAB7895_1d"
] <- 1

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "EMTAB7895_3d"
] <- 3

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "EMTAB7895_7d"
] <- 7

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "EMTAB9816_MI1D"
] <- 1

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "EMTAB9816_MI3D"
] <- 3

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "EMTAB9816_MI7D"
] <- 7

cds_myeloid_pt$day_post_MI[
  cds_myeloid_pt$orig.ident == "GSE119355"
] <- 11



post_MI_pseudotime_plot <- function(cds_obj) {

  pt <- pseudotime(cds_obj)

  plot_data <- data.frame(
    pseudotime  = ifelse(is.infinite(pt), NA, pt),
    day_post_MI = cds_obj$day_post_MI,
    celltype    = cds_obj$lineage_r.08,
    dataset     = cds_obj$orig.ident
  )

  plot_data <- subset(
    plot_data,
    !is.na(pseudotime) & !is.na(day_post_MI)
  )

  plot_data$day_post_MI <- factor(
    plot_data$day_post_MI,
    levels = c(1, 3, 7, 11)
  )

  ggplot(
    plot_data,
    aes(
      x = day_post_MI,
      y = pseudotime
    )
  ) +
    geom_violin(
      trim = FALSE,
      width = 0.85
    ) +
    geom_boxplot(
      width = 0.12,
      outlier.shape = NA
    ) +
    labs(
      x = "Days post-MI",
      y = "Monocle3 pseudotime"
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.title = element_text(face = "bold"),
      axis.text = element_text(face = "bold"),
      plot.margin = margin(5, 5, 5, 5)
    )
}

p_pt_myeloid_post_MI <- post_MI_pseudotime_plot(cds_myeloid_pt)

p_pt_myeloid_post_MI


ggsave(
  filename = "Pseudotime_myeloid_post_MI.pdf",
  plot     = p_pt_myeloid_post_MI,
  path     = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/Figures/09_Pseudotime",
  device   = "pdf",
  width    = 120,
  height   = 120,
  units    = "mm",
  bg       = "white"
)

ggsave(
  filename    = "Pseudotime_myeloid_post_MI.tiff",
  plot        = p_pt_myeloid_post_MI,
  path        = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/Figures/09_Pseudotime",
  device      = "tiff",
  width       = 120,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)



```


```{r Pseudotime Vln plot (Stromal)}

cds_stromal_pt$day_post_MI <- NA_real_

cds_stromal_pt$day_post_MI[
  cds_stromal_pt$orig.ident == "EMTAB7895_1d"
] <- 1

cds_stromal_pt$day_post_MI[
  cds_stromal_pt$orig.ident == "EMTAB7895_3d"
] <- 3

cds_stromal_pt$day_post_MI[
  cds_stromal_pt$orig.ident == "EMTAB7895_7d"
] <- 7

cds_stromal_pt$day_post_MI[
  cds_stromal_pt$orig.ident == "EMTAB9816_MI1D"
] <- 1

cds_stromal_pt$day_post_MI[
  cds_stromal_pt$orig.ident == "EMTAB9816_MI3D"
] <- 3

cds_stromal_pt$day_post_MI[
  cds_stromal_pt$orig.ident == "EMTAB9816_MI7D"
] <- 7




p_pt_stromal_post_MI <- post_MI_pseudotime_plot(cds_stromal_pt)

p_pt_stromal_post_MI


ggsave(
  filename = "Pseudotime_stromal_post_MI.pdf",
  plot     = p_pt_stromal_post_MI,
  path     = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/Figures/09_Pseudotime",
  device   = "pdf",
  width    = 120,
  height   = 120,
  units    = "mm",
  bg       = "white"
)

ggsave(
  filename    = "Pseudotime_stromal_post_MI.tiff",
  plot        = p_pt_stromal_post_MI,
  path        = "/projects/b1246/Isaac/MERTK/Atlas_Datasets/Figures/09_Pseudotime",
  device      = "tiff",
  width       = 120,
  height      = 120,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```





```{r Calculating Median Inferred pseudotime}

pt <- pseudotime(cds_myeloid_pt)

check_df <- data.frame(
  pseudotime = pt,
  day_post_MI = cds_myeloid_pt$day_post_MI
)

check_df <- subset(
  check_df,
  is.finite(pseudotime) & !is.na(day_post_MI)
)

aggregate(
  pseudotime ~ day_post_MI,
  data = check_df,
  FUN = median
)

pt_stromal <- pseudotime(cds_stromal_pt)

check_stromal <- data.frame(
  pseudotime  = pt_stromal,
  day_post_MI = cds_stromal_pt$day_post_MI
)

check_stromal <- subset(
  check_stromal,
  is.finite(pseudotime) & !is.na(day_post_MI)
)

aggregate(
  pseudotime ~ day_post_MI,
  data = check_stromal,
  FUN = median
)



```
























```{r Key gene expression along pseudotime}

rowData(cds_myeloid)$gene_short_name <- rownames(cds_myeloid)

# Key myeloid genes
key_genes_symbols <- c(
  "Ly6c2", "Ccr2",      # Early monocyte
  "Adgre1", "Mrc1",     # Macrophage differentiation
  "Arg1", "Retnla",     # Inflammatory resolution
  "Timd4", "Lyve1",     # Tissue remodeling
  "Mki67"               # Proliferation
)

print(head(rownames(cds_myeloid), 5))

key_genes_found <- key_genes_symbols[
  key_genes_symbols %in% rownames(cds_myeloid)
]
print(key_genes_found)

# If not found by symbol, trying Ensembl
if (length(key_genes_found) == 0) {
  sym_to_ens     <- setNames(gene_map$ensembl_gene_id,
                              toupper(gene_map$mgi_symbol))
  key_genes_ens  <- sym_to_ens[toupper(key_genes_symbols)]
  key_genes_ens  <- key_genes_ens[!is.na(key_genes_ens)]
  key_genes_found <- key_genes_ens[
    key_genes_ens %in% rownames(cds_myeloid)
  ]
  print(key_genes_found)
}

if (length(key_genes_found) > 0) {
  p_gene_pseudotime <- plot_genes_in_pseudotime(
    cds_myeloid[key_genes_found, ],
    color_cells_by      = "lineage_r.08",
    min_expr            = 0.1,
    ncol                = 3,
    label_by_short_name = TRUE
  ) +
    scale_color_manual(values = pal_hc,
                       guide  = guide_legend(
                         override.aes = list(size = 2),
                         ncol = 2)) +
    theme(strip.text  = element_text(face = "bold.italic", size = 8),
          legend.text = element_text(size = 7))

  p_gene_pseudotime
  
} 

ggsave(
  filename    = "Pseudotime_gene_expression_myeloid.tiff",
  plot        = p_gene_pseudotime,
  path        = pseudotime_dir,
  device      = "tiff",
  width       = 220,
  height      = 180,
  units       = "mm",
  dpi         = 600,
  compression = "lzw",
  bg          = "white"
)

```


```{r Saving cds objects}

saveRDS(cds_myeloid,
        "/projects/b1246/Isaac/MERTK/Atlas_Datasets/cds_myeloid_pt.rds")

saveRDS(cds_stromal,
        "/projects/b1246/Isaac/MERTK/Atlas_Datasets/cds_stromal_pt.rds")

sink(file.path(atlas_fig_dir, "session_info_atlas_creation.txt"))
print(sessionInfo())
sink()
```
